# Course information {-}

[Download](https://github.com/geanders/RProgrammingForResearch/raw/master/slides/CourseOverview.pdf) a pdf of the lecture slides covering this topic.

## Course overview

This document provides the course notes for Colorado State University's ERHS 535 course for Fall 2016. The course offers in-depth instruction on data collection, data management, programming, and visualization, using data examples relevant to academic research.

## Time and place

This course meets in Room 120 of the Environmental Health Building on Mondays and Wednesdays, 10:00 am--12:00 pm. Exceptions to these meeting times are: 

- There will be no meeting on Wednesday, Aug. 31.
- There will be no meeting on Labor Day (Monday, Sept. 5).
- To make up for missing class on Aug. 31, we will have a supplemental class on Friday, Sept. 9, 10:00 am--12:00 pm. You **will not** lose attendance points if you cannot attend this class, but **will** be responsible for the material covered.
- There are no course meetings the week of Thanksgiving.

## Detailed schedule

Here is a more detailed view of the schedule for this course for Fall 2016:

```{r echo = FALSE}
a <- data.frame(date = c("Aug. 22, 24",
                         "Aug. 29",
                         "Sept. 7, Sept. 9*", 
                         "Sept. 12, 14",
                         "Sept. 19, 21",
                         "Sept. 26, 28",
                         "Oct. 3, 5",
                         "Oct. 10, 12",
                         "Oct. 17, 19",
                         "Oct. 24, 26",
                         "Oct. 31, Nov. 2",
                         "Nov. 7, 9", 
                         "Nov. 14, 16", 
                         "Nov. 28, 30",
                         "Dec. 5, 7",
                         "Week of Dec. 12"),
                level = c("Preliminary",
                          rep("Basic", 4),
                          rep("Intermediate", 4),
                          rep("Advanced", 6), 
                          " "),
                content = c("R Preliminaries",
                            rep(c("Entering and cleaning data",
                            "Exploring data", 
                            "Reporting data results",
                            "Reproducible Research"), 2),
                            "Entering and cleaning data",
                            "Exploring data", 
                            "Reporting data results",
                            "Mapping in R",
                            "Package development 1",
                            "Package development 2",
                            "Group presentations"), 
                due = c("", "Quiz", "Quiz", "Quiz, HW #1",
                        "Quiz", "Quiz, HW #2", "Quiz", "Quiz, HW #3",
                        "Quiz", "HW #4", "Project proposal", "HW #5",
                        "", "HW #6", "Project draft", "Final project"))
knitr::kable(a, col.names = c("Dates", "Level", "Lecture content", "Graded items"))
```

## Grading

Course grades will be determined by the following five components:

```{r echo = FALSE}
a <- data.frame(a = c("Final group project",
                      "Weekly in-class quizzes, weeks 2-9",
                      "Homework",
                      "Attendance and class participation",
                      "Weekly in-course group exercises"),
                b = c(30, 25, 25, 10, 10))
knitr::kable(a, col.names = c("Assessment component", 
                              "Percent of grade"))
```

### Attendance and class participation

Because so much of the learning for this class is through interactive work in class, it is critical that you come to class. Out of a possible 10 points for class attendance, you will get:

- **10 points** if you attend all classes
- **8 points** if you miss one class
- **6 points** if you miss two classes
- **4 points** if you miss three classes
- **2 points** if you miss four classes
- **0 points** if you miss five or more classes

## Weekly in-course group exercises

Part of each class will be spent doing in-course group exercises. Ten points of your final grade will be based on your participation in these exercises. As long as you are in class and participate in these exercises, you will get full credit for this component. If you miss a class, to get credit towards this component of your grade, you will need to turn in a one-page document describing what you learned from doing the in-course exercise on your own time. All in-class exercises are included in the online course book at the end of the chapter on the associated material.

### In-class quizzes

You will have eight total in-class quizzes. You will have one for each of the Week 2--9 class meetings. There will be *at least* 10 questions per quiz. You will get 1/3 point for each correct answer. If you do the math, you can get full credit for this if you get at least 75% of your answers right. You can not get more than the maximum of 25 points for this component-- once you reach 25 points on quizzes, you will have achieved full credit for the quiz component of the course grade.

All quiz questions will be multiple choice, matching, or some other form of "close-answered" question (i.e., no open-response-style questions). You can not make up a quiz for a class period you missed. You can still get full credit on your total possible quiz points if you miss a class, but it means you will have to work harder and get more questions right for days you are in class. 

Because grading format for these quizzes allows for you to miss some questions and still get the full quiz credit for the course, I will not ever re-consider the score you got on a previous quiz, give points back for a wrong answer on a poorly-worded question, etc. However, if a lot of people got a particular question wrong, I will be sure to cover it in the next class period. Also, especially if a question was poorly worded and caused confusion, I will work a similar question into a future quiz-- in addition to the 10 guaranteed questions for that quiz-- so every student will have the chance to get an extra 1/3 point of credit for the question.

The "Vocabulary" appendix of our online book has the list of material for which you will be responsible for this quiz. Most of the functions and concepts will have been covered in class, but some may not. You are responsible for going through the list and, if there are things you don't know or remember from class, learning them. To do this, you can use help functions in R, Google, StackOverflow, books on R, ask a friend, and any other resource you can find.

In general, using R frequently in your research or other coursework will help you to prepare and do well on these quizzes. 

### Homework

There will be six homework assignments, starting a few weeks into the course and then due approximately every two weeks (see the detailed schedule in the online course book for exact due dates).

Homeworks should be done individually. You will get many chances to work with others during in-course exercises and your final group project, but these homeworks should be a chance to assess how well you understand and can use the course material on your own. 

Homeworks will be graded for correctness, but some partial credit will be given for questions you try but fail to answer correctly. If you can't completely do a required task, be sure to show and explain what you tried to do to complete it. 

Homework is due by the start of class on the due date. Your grade will be reduced by 10 points for each day it is late, and will receive no credit if it is late by over a week. 

### Final group project

You will do the final group project in groups of 2--3. The final product will be a statistical blog post-style article of 1,500 words or less and an accompanying Shiny web application. Come up with an interesting question you'd love to get the answer to that you think you can find data to help you answer. You will need to use the data you find, and R, to write your article. The final product will be a Word document created from an RMarkdown file and an accompanying Shiny web application.

Here are some articles to give you an idea of the style and content for this project:

- [Does Christmas come earlier each year?](http://www.statslife.org.uk/culture/1892-does-christmas-really-come-earlier-every-year)
- [Hilary: the most poisoned baby name in US history](http://hilaryparker.com/2013/01/30/hilary-the-most-poisoned-baby-name-in-us-history/)
- [Every Guest Jon Stewart Ever Had On "The Daily Show"](http://fivethirtyeight.com/datalab/every-guest-jon-stewart-ever-had-on-the-daily-show/)
- [Should Travelers Avoid Flying Airlines That Have Had Crashes in the Past?](http://fivethirtyeight.com/features/should-travelers-avoid-flying-airlines-that-have-had-crashes-in-the-past/)
- [Billion-Dollar Billy Beane](http://fivethirtyeight.com/features/billion-dollar-billy-beane/)

You will have in-class group work time during weeks 10--15 to work on this. This project will also require some work with your group outside of class. You will be able to get feedback from me through weekly informal written reports in these weeks. I will also provide feedback and help during the in-class group work time. 

The final group project will be graded with A through F, with the following point values (out of 25 possible):

- **25 points** for an A
- **20 points** for a B
- **15 points** for a C
- **10 points** for a D
- **5 points** for an F

If you turn nothing in, you will get **0 points**.

We will discuss expectations for this project, create groups, etc. around the middle of the semester. The focus for this will be on finding, cleaning, and using good data to answer an interesting question, and on presenting, summarizing, and explaining the data well.

## Helpful books for learning R

There are three publishers that are leaders in good books for learning R:

- [O'Reilly](http://shop.oreilly.com/category/browse-subjects/programming/r.do?sortby=publicationDate&page=all)
- [No Starch Press](https://www.nostarch.com)
- [Springer](http://www.springer.com/generic/search/results?SGWID=4-40109-24-653415-0&sortOrder=relevance&searchType=ADVANCED_CDA&language=en&searchScope=editions&resultStart=1&queryText=%22+R+%22)

Some particular books you might want to check out:

- [R for Data Science](http://r4ds.had.co.nz)
- [R for Dummies](http://discovery.library.colostate.edu/Record/.b45296674)
- [R in a Nutshell](http://discovery.library.colostate.edu/Record/.b40438880)
- [R Cookbook](http://discovery.library.colostate.edu/Record/.b36840282)
- [R Graphics Cookbook](http://www.amazon.com/R-Graphics-Cookbook-Winston-Chang/dp/1449316956/ref=sr_1_1?ie=UTF8&qid=1440997472&sr=8-1&keywords=r+graphics+cookbook)
- [A Beginner's Guide to R](http://discovery.library.colostate.edu/Record/.b44709365)
- [Roger Peng's Leanpub books](https://leanpub.com/u/rdpeng)
