\mainmatter

# Course information {-}

[Download](https://github.com/geanders/RProgrammingForResearch/raw/master/slides/CourseOverview.pdf) a pdf of the lecture slides covering this topic.

## Course overview

This document provides the course notes for Colorado State University's ERHS 535 course for Fall 2018. The course offers in-depth instruction on data collection, data management, programming, and visualization, using data examples relevant to academic research.

## Time and place

This course meets in Room 120 of the Environmental Health Building on Mondays and Wednesdays, 10:00 am--12:00 pm. Exceptions to these meeting times are: 

- There will be no meeting on Labor Day (Monday, Sept. 3).
- There are no course meetings the week of Thanksgiving (week of Nov. 19).
- I will be away from Fort Collins for two course dates (Aug. 27 and 29). I will videotape the lectures for these two class dates and post them online. I will offer a (voluntary) session on Aug. 31, 10:00-11:00 am, for anyone who would like to join to work on the in-course group exercises with classmates and with my feedback. 
<!-- - To make up for missing class on Aug. 31, we will have a supplemental class on Friday, Sept. 9, 10:00 am--12:00 pm. You **will not** lose attendance points if you cannot attend this class, but **will** be responsible for the material covered. -->

## Detailed schedule

Here is a more detailed view of the schedule for this course for Fall 2016:

```{r echo = FALSE}
a <- data.frame(date = c("Aug. 20, 22",
                         "Aug. 27, 29",
                         "Sept. 5", 
                         "Sept. 10, 12",
                         "Sept. 17, 19",
                         "Sept. 24, 26",
                         "Oct. 1, 3",
                         "Oct. 8, 10",
                         "Oct. 15, 17",
                         "Oct. 22, 24",
                         "Oct. 29, 31",
                         "Nov. 5, 7", 
                         "Nov. 12, 14", 
                         "Nov. 26, 28",
                         "Dec. 3, 5",
                         "Week of Dec. 10"),
                level = c("Preliminary",
                          rep("Basic", 4),
                          rep("Intermediate", 4),
                          rep("Advanced", 6), 
                          " "),
                content = c("R Preliminaries",
                            rep(c("Entering and cleaning data",
                            "Exploring data", 
                            "Reporting data results",
                            "Reproducible Research"), 2),
                            "Entering and cleaning data",
                            "Exploring data", 
                            "Reporting data results",
                            "Mapping in R",
                            "Package development 1",
                            "Package development 2",
                            "Group presentations"), 
                due = c("", "", "Quiz (W)", "Quiz (M), HW #1 (W)",
                        "Quiz (M)", "Quiz (M), HW #2 (W)", 
                        "Quiz (M)", "Quiz (M), HW #3 (W)",
                        "Quiz (M), Group choices (M)", "Quiz (M), Project proposal (M), HW #4 (W)",
                        "", "HW #5 (W)",
                        "", "HW #6 (W)", "Project draft (M)",
                        "Final project (M)"))
knitr::kable(a, col.names = c("Dates", "Level", "Lecture content", "Graded items"))
```

## Grading

Course grades will be determined by the following five components:

```{r echo = FALSE}
a <- data.frame(a = c("Final group project",
                      "Weekly in-class quizzes, weeks 3-10",
                      "Homework",
                      "Attendance and class participation",
                      "Weekly in-course group exercises"),
                b = c(30, 25, 25, 10, 10))
knitr::kable(a, col.names = c("Assessment component", 
                              "Percent of grade"))
```

### Attendance and class participation

Because so much of the learning for this class is through interactive work in class, it is critical that you come to class. Out of a possible 10 points for class attendance, you will get:

- **10 points** if you attend all classes
- **8 points** if you miss one class
- **6 points** if you miss two classes
- **4 points** if you miss three classes
- **2 points** if you miss four classes
- **0 points** if you miss five or more classes

Exceptions: 

- Attendance on the first day of class (Aug. 20) will not be counted.
- If you miss classes for "University-sanctioned"" activities. These can include attending a conference, travel to collect data for your dissertation), For these absences, you must inform prior to the date that you will be absence. No points will be lost for attendance if you provide a signed letter from your research advisor by Dec. 11, 2017 (start of finals week), and you can make arrangements with me to make up any missed work. For more details, see [CSU's Academic Policies on Course Attendance](http://catalog.colostate.edu/general-catalog/academic-standards/academic-policies/).
- If you have to miss class for a serious medical issue (e.g., operation, sickness severe enough to require a doctor's visit), the absence will be excused if you bring in a note from a doctor of other medical professional giving the date you missed and that it was for a serious medical issue.


### Weekly in-course group exercises

Part of each class will be spent doing in-course group exercises. Ten points of your final grade will be based on your participation in these exercises. As long as you are in class and participate in these exercises, you will get full credit for this component. If you miss a class, to get credit towards this component of your grade, you will need to turn in a one-page document describing what you learned from doing the in-course exercise on your own time. All in-class exercises are included in the online course book at the end of the chapter on the associated material.

### In-class quizzes

You will have eight total in-class quizzes. You will have one for each of the Week 3--10 class meetings. There will be *at least* 10 questions per quiz. You will get 1/3 point for each correct answer. If you do the math, you can get full credit for this if you get at least 75% of your answers right. You can not get more than the maximum of 25 points for this component-- once you reach 25 points on quizzes, you will have achieved full credit for the quiz component of the course grade.

All quiz questions will be multiple choice, matching, or some other form of "close-answered" question (i.e., no open-response-style questions). You can not make up a quiz for a class period you missed. You can still get full credit on your total possible quiz points if you miss a class, but it means you will have to work harder and get more questions right for days you are in class. 

Because grading format for these quizzes allows for you to miss some questions and still get the full quiz credit for the course, I will not ever re-consider the score you got on a previous quiz, give points back for a wrong answer on a poorly-worded question, etc. However, if a lot of people got a particular question wrong, I will be sure to cover it in the next class period. Also, especially if a question was poorly worded and caused confusion, I will work a similar question into a future quiz-- in addition to the 10 guaranteed questions for that quiz-- so every student will have the chance to get an extra 1/3 point of credit for the question.

The "Vocabulary" appendix of our online book has the list of material for which you will be responsible for this quiz. Most of the functions and concepts will have been covered in class, but some may not. You are responsible for going through the list and, if there are things you don't know or remember from class, learning them. To do this, you can use help functions in R, Google, StackOverflow, books on R, ask a friend, and any other resource you can find. The final version of the Vocabulary list you will be responsible will be posted by the Wednesday evening before each quiz.

In general, using R frequently in your research or other coursework will help you to prepare and do well on these quizzes. 

### Homework

There will be six homework assignments, starting a few weeks into the course and then due approximately every two weeks (see the detailed schedule in the online course book for exact due dates).

The first homework (HW #1) should be done individually. For some other homeworks, you may be given the option to work in small groups of approximately three students. 

For later homeworks, a subset of the full set of questions will be selected for which I will do a detailed grading of the code itself, with substantial feedback on coding. All other questions in the homework will be graded for completeness and based on the final answer produced. For homeworks with a heavy coding component, I will provide solution code for all questions. 

Homework is due by the start of class on the due date. Your grade will be reduced by 10 points for each day it is late, and will receive no credit if it is late by over a week. 

### Final group project

For the final project, you will work in small groups (3--4 people) on an R programming challenge. The final grade will be based on the resulting R software, as well as on a short group presentation and written report describing your work. You will be given a lot of in-class time during the last third of the semester to work with your group on this project, and you will also need to spend some time working outside of class to complete the project. More details on this project will be provided later in the semester.

<!-- ### Final group project -->

<!-- You will do the final group project in groups of 2--3. The final product will be a statistical blog post-style article of 1,500 words or less and an accompanying Shiny web application. Come up with an interesting question you'd love to get the answer to that you think you can find data to help you answer. You will need to use the data you find, and R, to write your article. The final product will be a Word document created from an RMarkdown file and an accompanying Shiny web application. -->

<!-- Here are some articles to give you an idea of the style and content for this project: -->

<!-- - [Does Christmas come earlier each year?](http://www.statslife.org.uk/culture/1892-does-christmas-really-come-earlier-every-year) -->
<!-- - [Hilary: the most poisoned baby name in US history](http://hilaryparker.com/2013/01/30/hilary-the-most-poisoned-baby-name-in-us-history/) -->
<!-- - [Every Guest Jon Stewart Ever Had On "The Daily Show"](http://fivethirtyeight.com/datalab/every-guest-jon-stewart-ever-had-on-the-daily-show/) -->
<!-- - [Should Travelers Avoid Flying Airlines That Have Had Crashes in the Past?](http://fivethirtyeight.com/features/should-travelers-avoid-flying-airlines-that-have-had-crashes-in-the-past/) -->
<!-- - [Billion-Dollar Billy Beane](http://fivethirtyeight.com/features/billion-dollar-billy-beane/) -->

<!-- You will have in-class group work time during weeks 10--15 to work on this. This project will also require some work with your group outside of class. You will be able to get feedback from me through weekly informal written reports in these weeks. I will also provide feedback and help during the in-class group work time.  -->

<!-- The final group project will be graded with A through F, with the following point values (out of 30 possible): -->

<!-- - **30 points** for an A -->
<!-- - **25 points** for a B -->
<!-- - **20 points** for a C -->
<!-- - **15 points** for a D -->
<!-- - **10 points** for an F -->

<!-- If you turn nothing in, you will get **0 points**. -->

<!-- #### Final presentation -->

<!-- - In total, the group's presentation should last 25-30 minutes. There will then be 5 minutes for questions.  -->
<!-- - Split the presentation up into three parts: (1) the main presentation (10-15 minutes), (2) a tutorial-style discussion of how you used R to do the project (10 minutes) and (3) an overview of your Shiny app (5 minutes). -->
<!-- - The main presentation part should include the following sections:  -->
<!--     + **Research question**: In one sentence, what is the main thing you were trying to figure out? -->
<!--     + **Introduction**: Why did you decide to do this project? You must convince us that your project answers an interesting and important question.  -->
<!--     + **Methods**: Where did you find data to answer the question? Why are they appropriate? Are there any limitations we should keep in mind? How did you investigate the data to try to answer your question? This should not include R code (save that for the tutorial part), but rather should use language like "To determine if ... was associated with ..., we measured the correlation ...". It's fine for this project if the Methods are fairly simple ("We investigated the distribution of ... using boxplots ...", "We took the mean and interquartile range of ...", "We mapped state-level averages of ...", etc.). Why do you choose to use the Methods you used? Why do you think they're appropriate and useful for your project? -->
<!--     + **Results**: What did you find out? Most of these slides should be figures or tables. Discuss your interpretation of your results as you present them. Ideally, you should be able to show your main results in about 4 slides, with one figure or table per slide. -->
<!--     + **Conclusions**: So what? How do your results compare with what other people have found out about your research question? Based on what you found, are there now other things you want to check out? -->
<!-- - The tutorial part should include the following sections:  -->
<!--     + **Overview of your approach in R**: Step us through a condensed version of how you did your project -->
<!--     + **Interesting packages / techniques**: Spend a bit more time on any parts that you found particularly interesting or exciting. Were there packages you used that were helpful that we haven't talked about in class? Did you find out how to do anything that you think other students could use in the future? Did you end up writing a lot of functions to use? Did you have an interesting way of sharing code and data among your group members? -->
<!--     + **Lessons learned**: If you were to do this project again from scratch, what would you do differently? Were there any big wrong turns along the way? Did you find out how to do something late in the project that would have saved you time if you'd started using it earlier? -->

<!-- #### Final report -->

<!-- The final report should not exceed 1,500 words. You should aim for no more than five figures and tables.  -->

<!-- In addition to the good examples linked above, you can find another example of the type of document we're looking for [here](http://fivethirtyeight.com/features/the-20-most-extreme-cases-of-the-book-was-better-than-the-movie/) from FiveThirtyEight. This example would have received an excellent score if it had been turned in for this class because it is clearly and engagingly written, it presents figures and tables that directly help to answer its main question and that are clearly explained and attractively presented, and its author has convinced me that this is an interesting question worth reading an article about. Notice that it is not very long, only has three figures and tables, and uses fairly simple analysis. -->

<!-- I will assess the final report on the following criteria: -->

<!-- - Is it written with correct spelling and grammar? -->
<!-- - Is it very clear what your over-arching research question is? -->
<!-- - Have you made a convincing case that this is an important or interesting problem? You could meet this criterion even by convincing me that this is a problem that just one of you is passionate about (as an example, see [here](http://hilaryparker.com/2013/01/30/hilary-the-most-poisoned-baby-name-in-us-history/)). -->
<!-- - Are the data that you chose to use reasonable for answering the question? Have you explained any caveats or limitations to the data that I should keep in mind when interpreting your results? As an example of how to do this for an analysis with secondary (imperfect) data, see how [this post](http://fivethirtyeight.com/features/the-20-most-extreme-cases-of-the-book-was-better-than-the-movie/) handles describing the data it uses, particularly in footnotes 1 and 3 and the sentences in the main text that correspond to them. -->
<!-- - Have you explained the way you analyzed the data clearly enough that I think that I could reproduce your analysis if I had your data? Have you explained a bit why your method of analyzing the data is appropriate for your question? Have you let me know about major caveats or limitations related to the methods of analysis you're using? -->
<!-- - Have you presented figures and / or tables with results that help answer your main research question? Is it clear what each is showing and how I should interpret it? (For a nice example of explaining how to interpret results, see footnote 4 [here](http://fivethirtyeight.com/features/the-20-most-extreme-cases-of-the-book-was-better-than-the-movie/).) Have you explained and interpreted your main results in the text? Have you pointed out any particularly interesting observations (interesting outliers, for example)?  -->
<!-- - When I'm finished with your article, do I have more insight into your research question than when I started?  -->
<!-- - If you include a quote or a figure from an outside source, you **must** include a full reference for it. Otherwise, I am okay with you doing referencing more in a blog-post style. That is, if you are repeating another person's ideas or findings, you must reference it, but you may use a web link rather than writing out full references. You do not need to include references of any type for standard analysis techniques (for example, you would not need to include a reference from a Stats book if you are fitting a regression model).  -->

<!-- #### Shiny app -->

<!-- Finally, you should create a Shiny app to provide an interactive visualization related to your research question. Expectations for that are:  -->

<!-- - It should work.  -->
<!-- - It should include text that is clearly written, without grammatical errors or typos. Any graphics are easily to interpret and follow some of the principles of good graphics covered in class.  -->
<!-- - It includes at least two rendered outputs (e.g., one plot and one table; two plots).  -->
<!-- - It is self-contained-- in other words, a user shouldn't need to read your report or hear your explain the app to understand it. It should include enough information on the app for a user to figure out how to use the app and interpret the output.  -->
<!-- - It is publicly available online. The easiest way to do this is to create a free account at shinyapps.io and publish to that. Only one person in the group needs to publish the app.  -->


## Course set-up

Please be sure you have the latest version of R and RStudio (Desktop version, Open Source edition) installed. Both are free for anyone to download. Also, be sure to sign up for a free GitHub account.

Here are useful links for this set-up: 

- R: https://cran.r-project.org 
- RStudio: https://www.rstudio.com/products/rstudio/#Desktop 
- Sign-up for a GitHub account: https://github.com

## Coursebook

This coursebook will serve as the only required textbook for this course. I am still in the process of editing and adding to this book, so content may change somewhat over the semester (particularly for the second half of the book, which is currently in a rawer draft than the beginning of the book). We typically cover about a chapter of the book each week of the course. 

This coursebook includes: 

- Links to the slides presented in class for each topic
- In-course exercises, typically including links to the data used in the exercise
- An appendix with homework assignments
- A list of vocabulary and concepts that should be mastered for each quiz

If you find any typos or bugs, or if you have any suggestions for how the book can be improved, feel free to post it on the book's [GitHub Issues page](https://github.com/geanders/RProgrammingForResearch/issues).

This book was developed using Yihui Xie's phenomenal [bookdown](https://bookdown.org) framework. The book is built using code that combines R code, data, and text to create a book for which R code and examples can be re-executed every time the book is re-built, which helps identify bugs and broken code examples quickly. The online book is hosted using GitHub's free [GitHub Pages](https://pages.github.com). All material for this book is available and can be explored at [the book's GitHub repository](https://github.com/geanders/RProgrammingForResearch).

### Other helpful books (not required)

The best book to supplement the coursebook and lectures for this course is [R for Data Science](http://r4ds.had.co.nz), by Garrett Grolemund and Hadley Wickham. The entire book is freely available online through the same format at the coursebook. You can also purchase a paper version of the book (published by O'Reilly) through Amazon, Barnes & Noble, etc., for around $40. This book is an excellent and up-to-date reference by some of the best R programmers in the world. 

There are a number of other useful books available on general R programming, including:

- [R for Dummies](https://colostate-primo.hosted.exlibrisgroup.com/primo-explore/fulldisplay?docid=01COLSU_ALMA51267598310003361&context=L&vid=01COLSU&lang=en_US&search_scope=Everything&adaptor=Local%20Search%20Engine&tab=default_tab&query=any,contains,r%20for%20dummies&sortby=rank&offset=0)
- [R Cookbook](https://colostate-primo.hosted.exlibrisgroup.com/primo-explore/fulldisplay?docid=01COLSU_ALMA21203304500003361&context=L&vid=01COLSU&lang=en_US&search_scope=Everything&adaptor=Local%20Search%20Engine&tab=default_tab&query=any,contains,r%20cookbook&sortby=rank&offset=0)
- [R Graphics Cookbook](http://www.amazon.com/R-Graphics-Cookbook-Winston-Chang/dp/1449316956/ref=sr_1_1?ie=UTF8&qid=1440997472&sr=8-1&keywords=r+graphics+cookbook)
- [Roger Peng's Leanpub books](https://leanpub.com/u/rdpeng)
- Various books on [bookdown.org](www.bookdown.org)

The R programming language is used extensively within certain fields, including statistics and bioinformatics. If you are using R for a specific type of analysis, you will be able to find many books with advice on using R for both general and specific statistical analysis, including many available in print or online through the CSU library. 
