# Course information

## Course overview

This document provides the course notes for Colorado State University's ERHS 535 course for Fall 2016. The course offers in-depth instruction on data collection, data management, programming, and visualization, using data examples relevant to academic research.

## Detailed schedule

```{r echo = FALSE}
a <- data.frame(date = c("Aug. 22, 24",
                         "Aug. 29",
                         "Sept. 7, Sept. 9*", 
                         "Sept. 12, 14",
                         "Sept. 19, 21",
                         "Sept. 26, 28",
                         "Oct. 3, 5",
                         "Oct. 10, 12",
                         "Oct. 17, 19",
                         "Oct. 24, 26",
                         "Oct. 31, Nov. 2",
                         "Nov. 7, 9", 
                         "Nov. 14, 16", 
                         "Nov. 28, 30",
                         "Dec. 5, 7",
                         "Week of Dec. 12"),
                level = c("Preliminary",
                          rep("Basic", 4),
                          rep("Intermediate", 4),
                          rep("Advanced", 6), 
                          " "),
                content = c("Preliminaries in R",
                            rep(c("Entering and cleaning data",
                            "Exploring data", 
                            "Reporting data results",
                            "Reproducible Research"), 2),
                            "Entering and cleaning data",
                            "Exploring data", 
                            "Reporting data results",
                            "Mapping in R",
                            "Package development 1",
                            "Package development 2",
                            "Group presentations"), 
                due = c("", "Quiz", "Quiz", "Quiz, HW #1",
                        "Quiz", "Quiz, HW #2", "Quiz", "Quiz, HW #3",
                        "Quiz", "HW #4", "Project proposal", "HW #5",
                        "", "HW #6", "Project draft", "Final project"))
knitr::kable(a, col.names = c("Dates", "Level", "Lecture content", "Graded items"))
```

## Grading

Course grades will be determined by the following five components:

```{r echo = FALSE}
a <- data.frame(a = c("Final group project",
                      "Weekly in-class quizzes, weeks 2-9",
                      "Homework",
                      "Attendance and class participation",
                      "Weekly in-course group exercises"),
                b = c(30, 25, 25, 10, 10))
knitr::kable(a, col.names = c("Assessment component", 
                              "Percent of grade"))
```

