# Exploring data #1

[Download](https://github.com/geanders/RProgrammingForResearch/raw/master/slides/CourseNotes_Week3.pdf) a pdf of the lecture slides covering this topic.

## Data from a package

So far we've covered three ways to get data into R:

1. From flat files (either on your computer or online)
2. From files like SAS and Excel
3. From R objects (i.e., using `load()`)

Many R packages come with their own data, which is very easy to load and use.

For example, the `faraway` package has a dataset called `worldcup` that you'll use today. To load it, use the `data()` function once you've loaded the package:

```{r}
library(faraway)
data("worldcup")
```

Unlike most data objects you'll work with, the data that comes with an R package will often have its own help file. You can access this using the `?` operator:

```{r, eval = FALSE}
?worldcup
```

To find out all the datasets that are available in the packages you currently have loaded, run `data()` without an option inside the parentheses:

```{r, eval = FALSE}
data()
```

As a note, you can similarly use `library()`, without the name of a package, to list all of the packages you have installed that you could call with `library()`:

```{r, eval = FALSE}
library()
```


## Plots to explore data

Plots can be invaluable in exploring your data. In this section, I will focus on teaching you the basics of `ggplot2` plotting. This section will focus on making **useful**, rather than **attractive** graphs, since we are focusing on exploring rather than presenting data at this stage. In a later section, I will discuss customization of `ggplot2` plots, to help you make more attractive plots that would go into final reports. 

This chapter covers how to make the following types of basic plots using functions from the `ggplot2` package (another member of the tidyverse!):

```{r}
library(ggplot2)
```

The basic steps behind creating a plot with `ggplot2` are:

1. You'll create an object of the `ggplot` class, typically specifying the **data** and some or all of the **aesthetics**; 
2. You'll add on **geoms** and other elements to create and customize the plot, using `+`.

*Note*: To avoid errors, end lines with `+`, don't start lines with it. 

### Initializing a ggplot object

Use the following conventions to initialize a ggplot object:

```{r, eval = FALSE}
## Generic code
object <- ggplot(dataframe, aes(x = column_1, y = column_2))
```

Notice that you first specify the dataframe with the data you want to plot and then you might specify either **mappings** or constant values for some or all of the aesthetics (`aes`). 

### Plot aesthetics

**Aesthetics** are elements that can show certain elements of the data. For example, color might show gender, x-position might show height, and y-position might show weight. 

```{r echo = FALSE, warning = FALSE, fig.align = "center", out.width = "0.6\\textwidth", message = FALSE, fig.width = 5, fig.height = 3}
library(dplyr)
data("nepali") 
nepali %>%
  tbl_df() %>% 
  distinct(id, .keep_all = TRUE) %>%
  mutate(sex = factor(sex, levels = c(1, 2), labels = c("Male", "Female"))) %>%
  ggplot(aes(x = ht, y = wt, color = sex)) + 
  geom_point() + 
  xlab("Height (cm)") + ylab("Weight (kg)")
```

In this graph, the mapped aesthetics are color, x, and y. 

*Note*: Any of these aesthetics could also be given a constant value, instead of being mapped to an element of the data. For example, all the points could be red, instead of showing gender. 

Here are some common plot aesthetics you might want to specify: 

```{r echo = FALSE}
aes_vals <- data.frame(aes = c("`x`", "`y`", "`shape`",
                               "`color`", "`fill`", "`size`",
                               "`alpha`", "`linetype`"),
                       desc = c("Position on x-axis", 
                                "Position on y-axis", 
                                "Shape",
                                "Color of border of elements", 
                                "Color of inside of elements",
                                "Size", 
                                "Transparency (1: opaque; 0: transparent)",
                                "Type of line (e.g., solid, dashed)"))
knitr::kable(aes_vals, col.names = c("Code", "Description"))
```

Which aesthetics you must specify depend on which geoms (more on those in a second) you're adding to the plot. You can find out the aesthetics you can use for a geom in the "Aesthetics" section of the geom's help file (e.g., `?geom_point`). Required aesthetics are in bold in this section of the help file and optional ones are not. 

### Adding geoms

Next, you'll want to add one or more `geoms` to create the plot. You can add these with `+` after the `ggplot` statement to initialize the ggplot object. Some of the most common geoms are:

```{r echo = FALSE}
plot_funcs <- data.frame(type = c("Histogram (1 numeric variable)",
                                  "Scatterplot (2 numeric variables)",
                                  "Boxplot (1 numeric variable, possibly 1 factor variable)",
                                  "Line graph (2 numeric variables)"), 
                         ggplot2_func = c("`geom_histogram`",
                                          "`geom_point`",
                                          "`geom_boxplot`",
                                          "`geom_line`"))
knitr::kable(plot_funcs, col.names = c("Plot type",
                                       "ggplot2 function"))
```

### Constant aesthetics

Instead of mapping an aesthetic to an element of your data, you can use a constant value for it. For example, you may want to make all the points green, rather than having color map to gender: 

```{r echo = FALSE, warning = FALSE, fig.align = "center", out.width = "0.6\\textwidth", message = FALSE, fig.width = 5, fig.height = 3}
nepali %>%
  tbl_df() %>% 
  distinct(id, .keep_all = TRUE) %>%
  mutate(sex = factor(sex, levels = c(1, 2), labels = c("Male", "Female"))) %>%
  ggplot(aes(x = ht, y = wt)) + 
  geom_point(color = "darkgreen") + 
  xlab("Height (cm)") + ylab("Weight (kg)")
```

In this case, you'll define that aesthetic when you add the geom, outside of an `aes` statement. 

If you're using a constant shape, you specify the shape with a number. Here are the shapes that correspond to the numbers 1 to 25:

```{r echo = FALSE, fig.width = 5, fig.height = 3, fig.align = "center"}
x <- rep(1:5, 5)
y <- rep(1:5, each = 5)
shape <- 1:25
to_plot <- data_frame(x = x, y = y, shape = shape)
ggplot(to_plot, aes(x = x, y = y)) + 
  geom_point(shape = shape, size = 4, color = "black", fill = "red") + 
  geom_text(label = shape, nudge_x = -0.25) +
  xlim(c(0.5, 5.5)) + 
  theme_void() + 
  scale_y_reverse()
```

If you are using a constant color, R has character names for different colors. For example: 

```{r echo = FALSE, fig.width = 5, fig.height = 3, fig.align = "center"}
x <- rep(0, 6)
y <- 1:6
color <- c("blue", "blue4", "darkorchid", "deepskyblue2", 
           "steelblue1", "dodgerblue3")
to_plot <- data_frame(x = x, y = y, color = color)
ggplot(to_plot, aes(x = x, y = y)) + 
  geom_point(color = color, size = 2) + 
  geom_text(label = color, hjust = 0, nudge_x = 0.05) + 
  theme_void() + 
  xlim(c(-1, 1.5)) +
  scale_y_reverse()
```

Google "R colors" and search the images to find links to listings of different R colors.

### Useful plot additions

There are also a number of elements that you can add onto a `ggplot` object using `+`. A few very frequently used ones are: 

```{r echo = FALSE}
plot_adds <- data.frame(add = c("`ggtitle`",
                                "`xlab`, `ylab`",
                                "`xlim`, `ylim`"),
                        descrip = c("Plot title",
                                    "x- and y-axis labels",
                                    "Limits of x- and y-axis"))
knitr::kable(plot_adds, col.names = c("Element", "Description"))
```

### Example dataset

For the example plots, I'll use a dataset in the `faraway` package called `nepali`. This gives data from a study of the health of a group of Nepalese children. 

```{r}
library(faraway)
data(nepali)
```

I'll be using functions from `dplyr` and `ggplot2`, so those need to be loaded:

```{r message = FALSE, warning = FALSE}
library(dplyr)
library(ggplot2)
```

Each observation is a single measurement for a child; there can be multiple observations per child. I'll subset out child id, sex, weight, height, and age, and I'll limit to each child's first measurement. 

```{r message = FALSE}
nepali <- nepali %>%
  # Subset to certain columns
  select(id, sex, wt, ht, age) %>%
  # Convert id and sex to factors
  mutate(id = factor(id),
         sex = factor(sex, levels = c(1, 2),
                      labels = c("Male", "Female"))) %>%
  # Limit to first obs. per child
  distinct(id, .keep_all = TRUE)
```

The data now looks like:

```{r}
head(nepali)
```

### Histograms

For `geom_histogram()`, the main aesthetic is `x`, the (numeric) vector for which you want to create a histogram:

```{r, fig.width = 2.5, fig.height = 2, message = FALSE, warning = FALSE, fig.align = "center"}
ggplot(nepali, aes(x = ht)) + 
  geom_histogram()
```

You can add some elements to the histogram, like `ggtitle`, `xlab`, and `xlim`:

```{r, fig.width = 2.5, fig.height = 2, fig.align = "center", message = FALSE, warning = FALSE}
ggplot(nepali, aes(x = ht)) + 
  geom_histogram(fill = "lightblue", color = "black") + 
  ggtitle("Height of children") + 
  xlab("Height (cm)") + xlim(c(0, 120))
```

`geom_histogram` also has its own special argument, `bins`. You can use this to change the number of bins that are used to make the histogram:

```{r, fig.width = 2.5, fig.height = 2, fig.align = "center", warning = FALSE, message = FALSE}
ggplot(nepali, aes(x = ht)) + 
  geom_histogram(fill = "lightblue", color = "black",
                 bins = 40) 
```

### Scatterplots

You can use the `geom_point` geom to create a scatterplot. For example, to create a scatterplot of height versus age for the Nepali data: 

```{r, fig.width = 2.5, fig.height = 2, warning = FALSE, fig.align = "center"}
ggplot(nepali, aes(x = ht, y = wt)) + 
  geom_point()
```

Again, you can use some of the options and additions to change the plot appearance:

```{r, fig.width = 2.5, fig.height = 2, fig.align = "center", message = FALSE, warning = FALSE}
ggplot(nepali, aes(x = ht, y = wt)) + 
  geom_point(color = "blue", size = 0.5) + 
  ggtitle("Weight versus Height") + 
  xlab("Height (cm)") + ylab("Weight (kg)")
```

You can also try mapping another variable, `sex`, to the `color` aesthetic:

```{r, fig.width = 3.5, fig.height = 2, fig.align = "center", message = FALSE, warning = FALSE}
ggplot(nepali, aes(x = ht, y = wt, color = sex)) + 
  geom_point(size = 0.5) + 
  ggtitle("Weight versus Height") + 
  xlab("Height (cm)") + ylab("Weight (kg)")
```

### Boxplots 

To create a boxplot, use `geom_boxplot`: 

```{r, fig.height = 2, fig.width = 2, warning = FALSE, fig.align="center"}
ggplot(nepali, aes(x = 1, y = ht)) + 
  geom_boxplot() + 
  xlab("")+ ylab("Height (cm)")
```

You can also create separate boxplots, one for each level of a factor. In this case, you'll need to include two aesthetics (`x` and `y`) when you initialize the ggplot object. 

```{r, fig.height = 2, fig.width = 2.5, fig.align = "center", warning = FALSE}
ggplot(nepali, aes(x = sex, y = ht, group = sex)) + 
  geom_boxplot() + 
  xlab("Sex")+ ylab("Height (cm)") 
```

### Extensions of `ggplot2`

There are lots of R extensions for creating other interesting plots. For example, you can use the `ggpairs` function from the `GGally` package to plot all pairs of scatterplots for several variables. Notice how this output shows continuous and binary variables differently.

```{r out.width = "\\textwidth", fig.align = "center", message = FALSE, warning = FALSE}
library(GGally)
ggpairs(nepali[, c("sex", "wt", "ht", "age")])
```

## Simple statistics functions

Here are some simple statistics functions you will likely use often:

Function  | Description
--------- | -----------------
`range()` | Range (minimum and maximum) of vector 
`min()`, `max()` | Minimum or maximum of vector
`mean()`, `median()` | Mean or median of vector
`table()` | Number of observations per level for a factor vector
`cor()` | Determine correlation(s) between two or more vectors
`summary()` | Summary statistics, depends on class

All of these take, as the main argument, the vector(s) for which you want the statistic. If there are missing values in the vector, you'll need to add an option to say what to do when them (e.g., `na.rm` or `use="complete.obs"`).

```{r}
mean(nepali$wt, na.rm = TRUE)
range(nepali$ht, na.rm = TRUE)
table(nepali$sex)
```

The `cor` function can take two or more vectors. If you give it multiple values, it will give the correlation matrix for all the vectors.

```{r}
cor(nepali$wt, nepali$ht, use = "complete.obs")
cor(nepali[ , c("wt", "ht", "age")], use = "complete.obs")
```

R supports object-oriented programming. This shows up with `summary()`. R looks to see what type of object it's dealing with, and then uses a method specific to that object type. 

```{r}
summary(nepali$wt)
summary(nepali$sex)
```

We'll see more of this when we do regression models.

You can also perform many of these tasks using `dplyr`. For example, to get the mean weight, you can use the `summarize` function: 

```{r}
nepali %>%
  summarize(mean_wt = mean(wt, na.rm = TRUE))
```

If you want to get summaries by group using `dplyr` (e.g., mean weight by sex), use `group_by` before running `summarize`:

```{r}
nepali %>%
  group_by(sex) %>%
  summarize(mean_wt = mean(wt, na.rm = TRUE))
```

## Logical vectors

Last week, you learned a lot about logical statements and how to use them with the `subset()` function. You can also use logical vectors, created with these statements, for a lot of other things. For example, you can use them directly in the square bracket indexing (`[..., ...]`).

A logical statement run on a vector will create a logical vector the same length as the original vector:

```{r}
is_male <- nepali$sex == "Male"
length(nepali$sex)
length(is_male)
```

The logical vector will have the value `TRUE` at any position where the original vector met the logical condition you tested, and `FALSE` anywhere else:

```{r}
head(nepali$sex)
head(is_male)
```

You can "flip" this logical vector (i.e., change every `TRUE` to `FALSE` and vice-versa) using `!`:

```{r}
head(is_male)
head(!is_male)
```

You can do a few cool things now with this vector. For example, you can use it with indexing to pull out just the rows where `is_male` is `TRUE`:

```{r}
head(nepali[is_male, ])
```

Or, with `!`, just the rows where `is_male` is `FALSE`:

```{r}
head(nepali[!is_male, ])
```

You can also use `sum()` and `table()` to find out how many males and females are in the dataset:

```{r}
sum(is_male); sum(!is_male)
table(is_male)
```


## In-course exercise

### Loading data from an R package

The data we'll be using today is from a dataset called `worldcup` in the package `faraway`. Load that data so you can use it on your computer (note: you will need to load and install the `faraway` package to do this). Use the help file for the data to find out more about the dataset. Use some basic functions, like `head`, `tail`, `colnames`, `str`, and `summary` to check out the data a bit. See if you can figure out:

- What variables are included in this dataset? (Check the column names.)
- What class is each column currently? In particular, which are numbers and which are factors?

#### Example R code:

Load the `faraway` package using `load()` and then load the data using `data()`:

```{r}
## Uncomment the next line if you need to install the package
# install.packages("faraway")
library(faraway)
data("worldcup")
```

Check out the help file for the `worldcup` dataset to find out more about the data. (Note: Only datasets that are parts of packages will have help files.)

```{r, eval = FALSE}
?worldcup
```

Check out the data a bit:

```{r}
str(worldcup)
head(worldcup)
tail(worldcup)
colnames(worldcup)
summary(worldcup)
```

### Basic plots of the data

Use some basic plots to check out this data. Try the following:

- Plot histograms of all the numeric variables (`Time`, `Shot`, `Passes`, `Tackles`, `Saves`)
- Plot scatterplots of different combinations of numeric variables (e.g., `Time` vs. `Shots`). Try doing this using the `geom_point()` geom from `ggplot2`. Also try doing it using the `ggpairs()` function from the `GGally` package, to plot several of these at the same time. Try using different constant or mapped values with the `color` aesthetic. 
- Create boxplots of `Time`, `Shots`, `Passes` and `Saves` by position.
- Go online and find out which teams were the top four teams in this World Cup (i.e., first through fourth places). Create a `top_teams` subset with just these teams. Plot boxplots of `Shots` and `Saves` by team for just these teams. 
- Did you notice any interesting features of the data when you did any of the graphs in this section?

#### Example R code:

Use histograms to explore the distribution of different variables. If you want to change the number of bins in the histogram, try playing around with the `bins` and `binwidth` arguments. You can use the `bins` argument to say how many bins you want (e.g., `bins = 50`). You can use the `binwidth` argument to say how wide you want the bins to be (e.g., `binwidth = 10` if you wanted bins to be 10 units wide, in the units of the variable mapped to the `x` aesthetic. Try using `fill` and `color` to change the appearance of the plot. Google "R colors" and search the images to find links to listings of different R colors.

```{r, message = FALSE, fig.align = "center", fig.width = 5, fig.height = 3}
library(ggplot2)
ggplot(worldcup, aes(x = Time)) + 
  geom_histogram()

ggplot(worldcup, aes(x = Time)) + 
  geom_histogram(bins = 50)

ggplot(worldcup, aes(x = Time)) + 
  geom_histogram(binwidth = 100)

ggplot(worldcup, aes(x = Time)) + 
  geom_histogram(binwidth = 50, color = "white", fill = "cyan4")
```

Create a scatterplot of `Time` versus `Passes`. To change the size of the points, use the `size` argument (use a number lower than 1 for smaller points, higher than 1 for larger points). Try changing the color and transparency of the points using the aesthetics `color` and `alpha`. Try using color to show each player's position by mapping `Position` to the `color` aesthetic. 

```{r, fig.align = "center", fig.width = 5, fig.height = 3}
ggplot(worldcup, aes(x = Time, y = Passes)) + 
  geom_point()

ggplot(worldcup, aes(x = Time, y = Passes)) + 
  geom_point(size = 0.5)

ggplot(worldcup, aes(x = Time, y = Passes)) + 
  geom_point(size = 2, color = "blue", alpha = 0.25)

ggplot(worldcup, aes(x = Time, y = Passes, color = Position)) + 
  geom_point()
```

Use the `ggpairs` function from the `GGally` package to plot scatterplots of all combinations of several numeric variables.

```{r, message=FALSE, warning=FALSE, fig.align = "center"}
library(GGally)
library(dplyr)
ggpairs(select(worldcup, Time, Shots, Passes, Tackles, Saves))
```

To create a boxplot of `Shots` by `Position`, you can use `geom_boxplot`:

```{r, fig.width=5, fig.height=3, fig.align = "center"}
ggplot(worldcup, aes(x = Position, y = Shots)) + 
  geom_boxplot()
```

The top four teams in this World Cup were Spain, the Netherlands, Germany, and Uruguay. Create a subset with just the data for these four teams:

```{r}
top_teams <- worldcup %>%
  filter(Team %in% c("Spain", "Netherlands", "Germany", "Uruguay")) %>% 
  mutate(Team = factor(Team))
```

This dataset will still have all the levels saved for the `Team` factor, even though it isn't using them all. You can re-set this by resetting `Team` as a factor, which is what I've done with the `mutate` line. When R creates a factor from a vector, its default is to only use as levels the values that show up in the vector. 

Now, you can plot the boxplots, mapping `Team` to the `x` aesthetic and `Shots` or `Saves` to the `y` aesthetic:

```{r, fig.width = 6, fig.height = 3, fig.align = "center"}
ggplot(top_teams, aes(x = Team, y = Shots)) + 
  geom_boxplot() + 
  ggtitle("Shots")

ggplot(top_teams, aes(x = Team, y = Saves)) + 
  geom_boxplot() + 
  ggtitle("Saves")
```

#### If you have extra time:

If you wanted to do the same plot for several different variables, you could loop through your code (we'll be covering more about loops in a few weeks). For example, you could create histograms for all of the numeric variables (if you do this in RStudio, you'll need to use the arrows on the plot window to move through and see all the different plots once you've created them):

```{r, fig.width = 3, fig.height = 2.5}
## Create an object with the column names for all of the numeric variables
my_vars <- colnames(worldcup)[3:7]

## Loop through all of those variables. Print out a histogram with the 
## variable, and have it print on the plot, as the main title, the 
## column name for that variable
for(var in my_vars){
  worldcup$to_plot <- worldcup[ , var]
  a <- ggplot(worldcup, aes(x = to_plot)) + 
    geom_histogram(bins = 20, color = "white", fill = "navy") + 
    xlab(var) + 
    ggtitle(paste("Histogram of", var))
  plot(a)
}
```

A few things to note in this example: 

- To map an element of the data to an aesthetic, it's easiest if that element is saved in a column in the dataframe. Within this loop, I'm making an extra column called `to_plot`, where I'm copying the column of the variable I want to plot each time the loop runs. That way, I can always use `x = to_plot` in the aesthetic mapping for the ggplot object. 
- If you run code to create a ggplot object within a loop, it won't automatically print. Instead, you need to use `print` to get the object to print out. One way to do that is to save the final ggplot object as an R object (here I'm saving it to `a`) and then use the `print` function to print that object. 
- Next week, we'll talk some about faceting, which can create multiple plots by variable like this in a lot less code. However, it's useful at this point to start thinking about how to extend code to use in loops, to save yourself time when you need to repeat something similar many times.

### Exploring the data using simple statistics and logical statements

Next, try checking out the data using some basic commands for simple statistics, like `mean()`, `range()`, `max()`, and `min()`. Use these, along with some logical statements, to help you answer the following questions:

- What is the range of time that players spent in the game? Who played the most World Cup time in this World Cup? For the minimum of the range of `Time`, how many players played this amount of time?
- What is the mean number of saves that players made? What is the mean number of saves just among the goalkeepers? How many of the players are goalkeepers? Did any non-goalkeeper make a save?

#### Example R code:

Use `range()` to find out the range of time these players played in the World Cup. 
```{r}
range(worldcup$Time)
```

To figure out who played the most time, you need to subset out the rows of the dataset where the `Time` variable equals the maximum of the `Time` variable for the whole dataset. There are a few ways to do that. Here I'm showing two: (1) using logic within the "square-bracket indexing", to pull out just rows where it is TRUE that the `Time` for that row equals `max(worldcup$Time)` and (2) using `filter` from the `dplyr` package to filter down to rows where where it is TRUE that the `Time` for that row equals `max(Time)` for the whole dataset. 

```{r}
max(worldcup$Time)
head(worldcup$Time == max(worldcup$Time))

worldcup[worldcup$Time == max(worldcup$Time), ]

worldcup %>%
  filter(Time == max(Time))
```

*Note*: You may have noticed that you lost the players names when you did this using the `dplyr` pipechain. That's because `dplyr` functions convert the data to a dataframe format that does not include rownames. If you want to keep players' names, use `mutate` to move those names from the rownames of the data into a column in the dataframe:

```{r}
worldcup %>%
  mutate(Name = rownames(worldcup)) %>%
  filter(Time == max(Time))
```

To calculate the mean number of saves among all the players, use the `mean` function, either by itself or within a `summarize` call:

```{r}
mean(worldcup$Saves)

worldcup %>%
  summarize(mean_saves = mean(Saves))
```

For the next parts of the question, it will be convenient to have a logical vector for whether each player is a goalkeeper, so here's how you would create that:

```{r}
goalie <- worldcup$Position == "Goalkeeper"
```

This new object, `goalie`, is a vector the same length as `worldcup$Position`. Each element of `goalie` says whether it is TRUE or FALSE that `worldcup$Position` is equal to "Goalkeeper" at that spot on the `worldcup$Position` vector. 

```{r}
head(goalie)
```

The `summary()` function will count up the total number of times that `goalie` is TRUE and FALSE.

```{r}
summary(goalie)
```

There are a few ways to use this vector to figure out how many players were goalkeepers. First, you could use `summary` (which I just showed) or `table`, and just read how many times this vector has the value `TRUE`. Second, since R saves logical vectors with `TRUE` as 1 and `FALSE` as 0, you could just the `sum` function to add up the vector to find out how often it's `TRUE` (`sum` adds up every value in the vector).

```{r}
table(goalie)
sum(goalie)
```

You could also answer this question by using `summarize` from `dplyr`. You need to `group_by` player position and then you can use the `n` function in `summarize` to count up the total number of observations in each group:

```{r}
worldcup %>%
  group_by(Position) %>%
  summarize(n_players = n())
```

Now, you can answer the questions about mean saves for goalies and max saves for non-goalies. First, try doing that using the `goalie` logical vector you created. If you put `goalie` in the square bracket indexing for the dataframe as the rows value (i.e., the index before the comma), R will subset out just the rows where `goalie` is equal to TRUE. If you put `!goalie` in the square bracket indexing as the rows value, R will just subset out the rows where `goalie` is equal to FALSE. You can use this index subsetting to figure out the mean number of saves per goalie and also whether any non-goalie made a save (by checking the maximum value or range of saves for non-goalies).

```{r}
head(worldcup[goalie, ])
mean(worldcup[goalie, "Saves"])
range(worldcup[!goalie, "Saves"])
```

You could also answer this quesiton using a `dplyr` pipe chain to summarize the data after grouping it by position:

```{r}
worldcup %>%
  group_by(Position) %>%
  summarize(number_players = n(), 
            mean_saves = mean(Saves),
            max_saves = max(Saves))
```

### Using regression models to explore data

For this part of the exercise, you'll use a dataset on weather, air pollution, and mortality counts in Chicago, IL. This dataset is called `chicagoNMMAPS` and is part of the `dlnm` package. Change the name of the dataframe to something shorter, like `chic`. Check out the data a bit to see what variables you have, and then perform the following tasks:

- Write out (on paper, not in R) the regression equation for regressing dewpoint temperature on temperature. 
- Try fitting a linear regression of dew point temperature (`dptp`) on temperature (`temp`). (Bonus points: Notice anything that seems unusual about these two variables in this dataset? You can find out with `summary`, but it helps if you know a bit about what dewpoint temperature measures.)
Save this model as the object `mod_1`. 
- Based on this regression, does there seem to be a relationship between temperature and dewpoint temperature in Chicago? (Hint: Try using `summary()` on the model object to get more information about the model you fit.) What is the p-value for the coefficient for temperature?
- Plot temperature (x-axis) versus dewpoint temperature (y-axis) for Chicago. Add in the regression line from the model you fit.
- Use `plot()` on the model object to check if some of the assumptions for the regression model seem appropriate.
- Try fitting the regression as a GLM, using `glm()`. Are your coefficients different?
- Does PM10 vary by day of the week? (Hint: The `dow` variable is a factor that gives day of the week. You can do an ANOVA analysis by fitting a linear model using this variable as the independent variable, and then run `anova()` on that model, and R will compare it to an intercept-only model.) What day of the week is PM10 generally highest? (Check the model coefficients to figure this out.) Try to write out (on paper) the regression equation for the model you're fitting.
- Try using `glm()` to run a Poisson regression of respiratory deaths (`resp`) on temperature during summer days. Start by creating a subset with just summer days called `summer`. (Hint: Use the `month` variable to do this-- just pull out the subset where the month is 6, 7, or 8, for June, July, and August.) Try to write out the regression equation for the model you're fitting.
- The coefficient for the temperature variable in this model is our best estimate (based on this model) of the **log relative risk** for a one degree Celcius increase in temperature. What is the **relative risk** associated with a one degree Celsius increase?

#### Example R code:

Install and load the `dlnm` package and then load the `chicagoNMMAPS` data. Change the name of the dataframe to `chic`, so it will be shorter to call for the rest of your work. 

```{r, message = FALSE, warning = FALSE}
# install.packages("dlnm")
library(dlnm)
data("chicagoNMMAPS")
chic <- chicagoNMMAPS
```

Fit a linear regression of `dptp` on `temp` and save as the object `mod_1`:

```{r}
mod_1 <- lm(dptp ~ temp, data = chic)
mod_1
```

Use `summary()` to check out a bit more about the model you fit. 

```{r}
summary(mod_1)
```

There does seem to be an association between temperature and dewpoint temperature: a unit increase in temperature is associated with a `r round(coef(mod_1)[2], 1)` unit increase in dewpoint temperature. The p-value for the temperature coefficient is <2e-16. This is far below 0.05, which suggests we would be very unlikely to see such a strong association by chance if the null hypothesis, that the two variables are not associated, were true.

Plot these two variables and add in the regression line from the model (note: I've used the `col = ` option to make the color of the points gray). Use `abline()` to add the regression line for the model you fit. 

```{r}
ggplot(chic, aes(x = temp, y = dptp)) + 
  geom_point(size = 0.5, col = "gray") + 
  geom_smooth(method = "lm", color = "black")

mod_coefs <- coef(mod_1)
ggplot(chic, aes(x = temp, y = dptp)) + 
  geom_point(size = 0.5, col = "gray") + 
  geom_abline(aes(intercept = mod_coefs[1], slope = mod_coefs[2]))
```

Plot some various plots to check model assumptions for the model you fit using the `plot()` function on your model object:

```{r, fig.width = 6, fig.height = 6}
par(mfrow = c(2, 2)) # Set to four plots per panel -- 2 rows, 2 columns
plot(mod_1)
par(mfrow = c(1, 1)) # Reset to one plot per panel
```

Try fitting the model using `glm()`. Call it `mod_1a`. Compare the coefficients for the two models. You can use the `coef()` function on an `lm` or `glm` object to pull out just the model coefficients.

```{r}
mod_1a <- glm(dptp ~ temp, data = chic)

coef(mod_1)
coef(mod_1a)
```

The results from the two models are identical.

Fit a model of PM10 regressed on day of week, where day of week is a factor. 

```{r}
mod_2 <- lm(pm10 ~ dow, data = chic)
summary(mod_2)
```

Use the `anova()` command to compare this model to a model with only an intercept (i.e., one that only fits a global mean and uses that as the expected value for all of the observations).

```{r}
anova(mod_2)
```

The p-value for an ANOVA of the model with day-of-week coefficients versus the model that just has an intercept is < 2.2e-16. This is well below 0.05, which suggests that day-of-week is associated with PM10 concentration, as a model that includes day-of-week does a much better job of explaining variation in PM10 than a model without it does. (Note, too, that the F value and Pr(>F) for the `anova()` call are identical to the F-statistic information given in the `summary()` of the model object. This will always be true when you're using `anova()` to compare a model to a model with just an intercept.)

Use a boxplot to visually compare PM10 by day of week. 

```{r, fig.height = 3, fig.width = 6}
ggplot(chic, aes(x = dow, y = pm10)) + 
  geom_boxplot()
```

Now try the same plot, but try using the `ylim = ` option to change the limits on the y-axis for the graph, so you can get a better idea of the pattern by day of week (some of the extreme values are very high, which makes it hard to compare by eye when the y-axis extends to include them all).

```{r, fig.height = 3, fig.width = 6}
ggplot(chic, aes(x = dow, y = pm10)) + 
  geom_boxplot() + 
  ylim(c(0, 100))
```

Create a subset called `summer` with just the summer days:

```{r}
summer <- chic %>%
  filter(month %in% 6:8)
```

Use `glm()` to fit a Poisson model of respiratory deaths regressed on temperature. Since you want to fit a Poisson model, use the option `family = poisson(link = "log")`. 

```{r}
mod_3 <- glm(resp ~ temp, data = summer,
             family = poisson(link = "log"))
summary(mod_3)
```

Use the fitted model coefficient to determine the relative risk for a one degree Celcius increase in temperature. First, remember that you can use the `coef()` function to read out the model coefficients. The second of these is the value for the temperature coefficient. That means that you can use indexing (`[2]`) to get just that value. That's the log relative risk; take the exponent to get the relative risk.

```{r}
coef(mod_3)
coef(mod_3)[2]
exp(coef(mod_3)[2])
```
