# Reporting data results #1

[Download](https://github.com/geanders/RProgrammingForResearch/raw/master/slides/CourseNotes_Week4.pdf) a pdf of the lecture slides covering this topic.

## Guidelines for good plots

There are a number of very thoughtful books and articles about creating graphics that effectively communicate information. \medskip

Some of the authors I highly recommend (and from whose work I've pulled the guidelines for good graphics we'll talk about this week) are: 

- Edward Tufte
- Howard Wainer
- Stephen Few
- Nathan Yau

You should plan, in particular, to read *The Visual Display of Quantitative Information* by Edward Tufte before you graduate.

This week, we'll focus on six guidelines for good graphics, based on the writings of these and other specialists in data display. \medskip

The guidelines are: 

1. Aim for high data density. 
2. Use clear, meaningful labels.
3. Provide useful references.
4. Highlight interesting aspects of the data. 
5. Make order meaningful. 
6. When possible, use small multiples.

For the examples, I'll use `dplyr` for data cleaning and, for plotting, the packages `ggplot2`, `gridExtra`, and `ggthemes`.

```{r message = FALSE}
library(dplyr)

library(ggplot2)
library(gridExtra)
library(ggthemes)
```

You can load the data for today's examples with the following code: 

```{r message = FALSE}
library(faraway)
data(nepali)
data(worldcup)

library(dlnm)
data(chicagoNMMAPS)
chic <- chicagoNMMAPS
chic_july <- chic %>%
  filter(month == 7 & year == 1995)
```

## High data density

Guideline 1: **Aim for high data density.** \bigskip

You should try to increase, as much as possible, the **data to ink ratio** in your graphs. This is the ratio of "ink" providing information to all ink used in the figure. \medskip

One way to think about this is that the only graphs you make that use up a lot of your printer's ink should be packed with information.


The two graphs below show the same information. Compare the amount of ink used in the left plot to the amount used in the right plot to see how graphs with the same information can have very different data densities. \bigskip

```{r echo = FALSE, fig.height = 3, fig.width = 8}
a <- ggplot(worldcup, aes(Position)) + 
        geom_bar() + coord_flip() + 
        ylab("Number of players") + 
        ggtitle("1. Lower data density")

ex <- group_by(worldcup, Position) %>%
        summarise(n = n())
b <- ggplot(ex, aes(x = n, y = Position)) + 
        geom_point() +  
        xlab("Number of players") +  ylab("") + 
        theme_few() + 
        xlim(0, 250) + 
        ggtitle("2. Higher data density")
grid.arrange(a, b, ncol = 2)
```

The two graphs below show another example of very different data densities in two plots showing the same information: \bigskip

```{r echo = FALSE, fig.height = 3, fig.width = 8}
a <- ggplot(chic_july, aes(x = date, y = death)) + 
        geom_area(fill = "black") + 
        xlab("Day in July 1995") + 
        ylab("All-cause deaths") + 
        ylim(0, 450) + 
        theme_excel() + 
        ggtitle("1. Lower data density")

b <- ggplot(chic_july, aes(x = as.POSIXlt(date)$mday,
                           y = death)) + 
        geom_line() + 
        xlab("Day in July 1995") + 
        ylab("All-cause deaths") + 
        ylim(0, 450) + 
        theme_tufte() + 
        ggtitle("2. Higher data density")

grid.arrange(a, b, ncol = 2)
```

One quick way to increase data density in `ggplot2` is to change the *theme* for the plot. This essentially changes the "background" elements to a plot, including elements like the plot grid, background color, and the font used for labeling. \bigskip

Some themes come with `ggplto2`, including:

- `theme_bw`
- `theme_minimal`
- `theme_void` 

The `ggthemes` packages has some excellent additional themes.

The following slides show some examples of the effects of using different themes. The following code creates a plot of daily deaths in Chicago in July 1995:

```{r}
chic_plot <- ggplot(chic_july, aes(x = date, y = death))  +
        geom_point(color = "red")
```

Next, we can see how the graph looks with the default theme and with other themes.

The left graph shows the graph with the default theme, while the right shows the effect of adding the black-and-white theme that comes with `ggplot2` as `theme_bw`:

```{r, fig.height = 3, fig.width = 8}
a <- chic_plot
b <- chic_plot + theme_bw()
grid.arrange(a, b, ncol = 2)
```

Stephen Few theme: 

```{r, fig.height = 3, fig.width = 8}
a <- chic_plot
b <- chic_plot + theme_few()
grid.arrange(a, b, ncol = 2)
```

Edward Tufte theme: 

```{r, fig.height = 3, fig.width = 8}
a <- chic_plot
b <- chic_plot + theme_tufte()
grid.arrange(a, b, ncol = 2)
```

You can even use themes to add some questionable choices for different elements. For example, `ggthemes` includes an Excel theme: 

```{r, fig.height = 3, fig.width = 8}
a <- chic_plot
b <- chic_plot + theme_excel()
grid.arrange(a, b, ncol = 2)
```


## Meaningful labels

Guideline 2: **Use clear, meaningful labels.** \bigskip

Graph defaults often use abbreviations for axis labels and other labeling. Further, text labels can sometimes be aligned in a way that makes them hard to read. The plots below give an example of the same information shown without (left) and with (right) clear, meaningful labels. \bigskip

```{r echo = FALSE, fig.height = 3, fig.width = 8}
ex <- worldcup
ex$Position <- factor(ex$Position, 
                      levels = c("Defender",
                                 "Forward",
                                 "Goalkeeper",
                                 "Midfielder"),
                      labels = c("DF", "FW",
                                 "GK", "MF"))
a <- ggplot(ex, aes(Position)) + 
        geom_bar() + 
        ylab("Pls") + 
        xlab("Pos") +
        ggtitle("1. Murkier labels") + 
        theme(axis.text.x = 
                      element_text(angle = 90,
                                   vjust = 0.5,
                                   hjust=1))

b <- ggplot(worldcup, aes(Position)) + 
        geom_bar(fill = "lightgray") + coord_flip() + 
        ylab("Number of players") + xlab("") + 
        theme_tufte() + 
        ggtitle("2. Clearer labels")

grid.arrange(a, b, ncol = 2)
```

There are a few strategies you can use to make labels clearer: 

- Add `xlab` and `ylab` elements to the plot, rather than relying on the column names in the original data. This can also be done with `scale` elements (e.g., `scale_x_continuous`), which give you more power to also make other changes to the x- and y-axes.
- Include units of measurement in axis titles when relevant. If units are dollars or percent, check out the `scales` package, which allows you to add labels directly to axis elements by including arguments like `labels = percent` in `scale` elements.
- If the x-variable requires longer labels (player positions in the example above), consider flipping the coordinates, rather than abbreviating or rotating the labels. You can use `coord_flip` to do this.

## References

Guideline 3: **Provide useful references.** \bigskip

Data is easier to interpret when you add references. For example, if you show what it typical, it helps viewers interpret how unusual outliers are. The graph below on the right has added shading showing the range of daily deaths in July in Chicage for 1990--1994 and 1996--2000, to clarify how unusual July 1995 was. \bigskip

```{r echo = FALSE, fig.height = 3, fig.width = 8}
chic_july <- subset(chic, month == 7 & year == 1995)
chic_july_ref <- filter(chic, month == 7 & 
                        year %in% c(1990:1994,
                                    1996:2000)) %>%
        summarise(mean = mean(death),
                  min = min(death),
                  max = max(death))
ref_points <- data.frame(date = c(-2, 33, 33, -2),
        death = c(rep(chic_july_ref$max, 2),
                  rep(chic_july_ref$min, 2)))
        
a <- ggplot(chic_july, aes(x = as.POSIXlt(date)$mday,
                           y = death)) + 
        geom_line() + 
        xlab("Day in July 1995") + 
        ylab("All-cause deaths") + 
        ylim(0, 450) + 
        theme_tufte() + 
        ggtitle("1. No reference")

b <- ggplot(chic_july, aes(x = as.POSIXlt(date)$mday,
                           y = death)) + 
        xlab("Day in July 1995") + 
        ylab("All-cause deaths") + 
        ylim(0, 450) + 
        theme_tufte() + 
        geom_polygon(aes(x = date, y = death), 
                     data = ref_points,
                     color = "lightgray", 
                     alpha = 0.1) + 
        geom_line() + 
        ggtitle("2. Reference")

grid.arrange(a, b, ncol = 2)
```

Another useful way to add references is to add a linear or smooth fit to the data, to help clarify trends in the data. \bigskip

```{r echo = FALSE, message = FALSE, fig.width = 8, fig.height = 4}
ex <- subset(worldcup, Position == "Forward")
a <- ggplot(ex, aes(x = Passes, y = Shots)) + 
        geom_point(size = 1.5) + 
        theme_few()  + 
        ggtitle("1. No reference")

b <- ggplot(ex, aes(x = Passes, y = Shots)) + 
        geom_point(size = 1.5) + 
        theme_few()  + 
        geom_smooth() + 
        ggtitle("2. Reference")

grid.arrange(a, b, ncol = 2)
```

You can use the function `geom_smooth` to add a smooth or linear reference line:

```{r fig.width = 3, fig.height = 3, out.width = "0.4\\textwidth", fig.align = "center"}
ggplot(worldcup, aes(x = Passes, y = Shots)) + 
  geom_point() + theme_few() + 
  geom_smooth(method = "lm")
```

The most useful `geom_smooth` parameters to know are: 

- `method`: The default is to add a loess curve if the data includes less than 1000 points and a generalized additive model for 1000 points or more. However, you can change to show the fitted line from a linear model using `method = "lm"` or from a generalized linear model using `method = "glm"`.
- `span`: How wiggly or smooth the smooth line should be (smaller value: more wiggly; larger value: more smooth)
- `se`: TRUE or FALSE, indicating whether to include shading for 95% confidence intervals.
- `level`: Confidence level for confidence interval (e.g., `0.90` for 90% confidence intervals)

Lines and polygons can also be useful for adding references. Useful geoms include: 

- `geom_hline`, `geom_vline`: Add a horizontal or vertical line
- `geom_abline`: Add a line with an intercept and slope
- `geom_polygon`: Add a filled polygon
- `geom_path`: Add an unfilled polygon

When adding these references: 

- Add reference elements first, so they will be plotted under the data, instead of on top of it. 
- Use `alpha` to add transparency to these elements. 
- Use colors that are unobtrusive (e.g., grays)
- For lines, consider using non-solid line types (e.g., `linetype = 3`)

## Highlighting

Guideline 3: **Highlight interesting aspects.** \bigskip

Consider adding elements to highlight noteworthy elements of the data. For example, in the graph on the right, the days of a major heat wave have been highlighted with a red line. \bigskip

```{r echo = FALSE, fig.height = 3, fig.width = 8}
chic_july <- subset(chic, month == 7 & year == 1995)
chic_july_ref <- filter(chic, month == 7 & 
                        year %in% c(1990:1994,
                                    1996:2000)) %>%
        summarise(mean = mean(death),
                  min = min(death),
                  max = max(death))
ref_points <- data.frame(date = c(-2, 33, 33, -2),
        death = c(rep(chic_july_ref$max, 2),
                  rep(chic_july_ref$min, 2)))
hw <- data.frame(date = c(12, 16, 16, 12),
                 death = c(425, 425, 0, 0))
        
a <- ggplot(chic_july, aes(x = as.POSIXlt(date)$mday,
                           y = death)) + 
        xlab("Day in July 1995") + 
        ylab("All-cause deaths") + 
        ylim(0, 450) + 
        theme_tufte() + 
        geom_polygon(aes(x = date, y = death), 
                     data = ref_points,
                     color = "lightgray", 
                     alpha = 0.1) + 
        geom_line() + 
        ggtitle("1. No highlighting")

b <- ggplot(chic_july, aes(x = as.POSIXlt(date)$mday,
                           y = death)) + 
        xlab("Day in July 1995") + 
        ylab("All-cause deaths") + 
        ylim(0, 450) + 
        theme_tufte() + 
        geom_polygon(aes(x = date, y = death), 
                     data = ref_points,
                     color = "lightgray", 
                     alpha = 0.1) + 
        geom_line(aes(x = date, y = death),
                  data = hw[1:2, ],
                     color = "red",
                  size = 2) +
        geom_line() + 
        ggtitle("2. With highlighting")

grid.arrange(a, b, ncol = 2)
```

In the below graphs, the names of the players with the most shots and passes have been added to highlight these unusual points. \bigskip

```{r echo = FALSE, message = FALSE, fig.width = 8, fig.height = 4}
ex <- subset(worldcup, Position == "Forward")
a <- ggplot(ex, aes(x = Passes, y = Shots)) + 
        geom_point(size = 1.5, alpha = 0.5) + 
        theme_few()  + 
        ggtitle("1. No highlighting")

most_shots <- ex[which.max(ex$Shots), ]
most_passes <- ex[which.max(ex$Passes), ]
b <- ggplot(ex, aes(x = Passes, y = Shots)) + 
        geom_point(size = 1.5, alpha = 0.5) + 
        theme_few()  + 
        ggtitle("2. Highlighting") + 
        geom_text(data = most_shots,
           label = paste(rownames(most_shots), ",",
                                most_shots$Team, " "),
                  colour = "blue", size = 3,
                  hjust = 1, vjust = 0.4) + 
        geom_text(data = most_passes,
           label = paste(rownames(most_passes), ",",
                                most_passes$Team, " "),
                  colour = "blue", size = 3,
                  hjust = 1, vjust = 0.4) 

grid.arrange(a, b, ncol = 2)
```

One helpful way to annotate is with text, using `geom_text()`. For this, you'll first need to create a dataframe with the hottest day in the data:

```{r}
hottest_day <- chic_july %>%
  filter(temp == max(temp))
hottest_day[ , 1:6]
```

```{r fig.height = 3, fig.width = 4, out.width = "0.7\\textwidth", fig.align = "center"}
chic_plot + geom_text(data = hottest_day, 
                      label = "Max",
                      size = 3)
```

With `geom_text`, you'll often want to use position adjustment (the `position` parameter) to move the text so it won't be right on top of the data points: 

```{r fig.height = 3, fig.width = 4, out.width = "0.5\\textwidth", fig.align = "center", message = FALSE}
chic_plot + geom_text(data = hottest_day, 
                      label = "Max",
                      size = 3, hjust = 0, vjust = -1)
```

You can also use lines to highlight. For this, it is often useful to create a new dataframe with data for the reference. To add a line for the Chicago heat wave, I've added a dataframe called `hw` with the relevant date range. I'm setting the y-value to be high enough (425) to ensure the line will be placed above the mortality data. 

```{r}
hw <- data.frame(date = c(as.Date("1995-07-12"), 
                          as.Date("1995-07-16")),
                 death = c(425, 425))
        
b <- chic_plot + 
        geom_line(data = hw,
                  aes(x = date, y = death),
                  size = 2)
```

```{r fig.height = 3, fig.width = 4, out.width = "0.7\\textwidth", fig.align = "center"}
b
```

## Order

Guideline 4: **Make order meaningful.** \bigskip

You can make the ranking of data clearer from a graph by using order to show rank. Often, factor or categorical variables are ordered by something that is not interesting, like alphabetical order. 

```{r echo = FALSE, fig.width = 8, fig.height = 5}
ex <- group_by(worldcup, Team) %>%
        summarise(mean_time = mean(Time))

a <- ggplot(ex, aes(x = mean_time, y = Team)) + 
        geom_point() + 
        theme_few() + 
        xlab("Mean time per player (minutes)") + ylab("") + 
        ggtitle("1. Alphabetical order")

ex2 <- arrange(ex, mean_time) %>%
        mutate(Team = factor(Team, levels = Team))
b <- ggplot(ex2, aes(x = mean_time, y = Team)) + 
        geom_point() + 
        theme_few() + 
        xlab("Mean time per player (minutes)") +  ylab("") + 
        ggtitle("2. Meaningful order")

grid.arrange(a, b, ncol = 2)
```

You can re-order factor variables in a graph by resetting the factor using the `factor` function and changing the order that levels are included in the `levels` parameter. 

## Small multiples

Guideline 5: **When possible, use small multiples.** \bigskip

*Small multiples* are graphs that use many small plots showing the same thing for different facets of the data. For example, instead of using color in a single plot to show data for males and females, you could use two small plots, one each for males and females. \bigskip

Typically, in small multiples, all plots with use the same x- and y-axes. This makes it easier to compare across plots, and it also allows you to save room by limiting axis annotation.

```{r echo = FALSE, message = FALSE, fig.height = 6, fig.width = 8}
ex <- subset(worldcup, Position %in% c("Forward",
                                       "Midfielder"))
ex2 <- group_by(ex, Team) %>%
        summarise(mean = mean(Shots)) %>%
        arrange(desc(mean))
ex$Team <- factor(ex$Team, 
                      levels = ex2$Team)
a <- ggplot(ex, aes(x = Time, y = Shots)) + 
        geom_point() + 
        theme_few() + 
        facet_wrap(~ Team, ncol = 8) + 
        geom_smooth(method = "lm", se = FALSE)
a
```

You can use the `facet` functions to create small multiples. This separates the graph into several small graphs, one for each level of a factor. \bigskip

The `facet` functions are: 

- `facet_grid()`
- `facet_wrap()`

For example, to create small multiples by sex for the Nepali dataset, when plotting height versus weight, you can call:

```{r warning = FALSE, fig.width = 8, fig.height = 3}
ggplot(nepali, aes(ht, wt)) + 
        geom_point() + 
        facet_grid(. ~ sex)
```

The `facet_grid` function can facet by one or two variables. One will be shown by rows, and one by columns:

```{r eval = FALSE}
## Generic code
facet_grid([factor for rows] ~ [factor for columns])
```

The `facet_wrap()` function can only facet by one variable, but it can "wrap" the small graphs for that variable, so the don't all have to be in one row or column:

```{r eval = FALSE}
## Generic code
facet_wrap(~ [factor for faceting], ncol = [# of columns])
```

Often, when you do faceting, you'll want to re-name your factors levels or re-order them. For this, you'll need to use the `factor()` function on the original vector. For example, to rename the `sex` factor levels from "1" and "2" to "Male" and "Female", you can run:

```{r}
nepali <- nepali %>%
  mutate(sex = factor(sex, levels = c(1, 2), 
                      labels = c("Male", "Female")))
```

Notice that the labels for the two graphs have now changed:

```{r warning = FALSE, fig.width = 8, fig.height = 3}
ggplot(nepali, aes(ht, wt)) + 
        geom_point() + 
        facet_grid(. ~ sex)
```

To re-order the factor, and show the plot for "Female" first, you can use `factor` to change the order of the levels:

```{r}
nepali <- nepali %>%
  mutate(sex = factor(sex, levels = c("Female", "Male")))
```

Now notice that the order of the plots has changed:

```{r warning = FALSE, fig.width = 8, fig.height = 3}
ggplot(nepali, aes(ht, wt)) + 
        geom_point() + 
        facet_grid(. ~ sex)
```

## To find out more

Some excellent further references for plotting are:

- R Graphics Cookbook (book and website)
- Google images

## In-course exercise

### Designing a plot

For today's exercise, you'll be building a plot using the `worldcup` data from the `faraway` package. First, load in that data. 

Next, say you want to look at the relationship between the number of minutes that a player played in the 2010 World Cup (`Time`) and the number of shots the player took on goal (`Shots`). On a sheet of paper, and talking with your partner, decide how the two of you would design a plot to explore and present this relationship. How would you incorporate some of the principles of creating good graphs?

#### Example R code

```{r}
library(faraway)
data(worldcup)
head(worldcup, 2)
```

This dataset has the players' names as rownames, rather than in a column. Once we start using `dplyr` functions, we'll lose these rownames. Therefore, start by converting the rownames to a column called `Player`:

```{r message = FALSE}
library(dplyr)
worldcup <- worldcup %>%
  mutate(Player = rownames(worldcup))
head(worldcup, 2)
```


### Basic scatterplots with R base graphics and ggplot

Perform the following tasks:

- Install and load the `ggplot2` and `ggthemes` packages. 
- For the `worldcup` data, plot a scatterplot of Time (on the x-axis) versus Shots. 
- Add some ggplot elements to make this plot a bit more attractive. (For example, change the x- and y-axis labels and add a title.) 

#### Example R code

Install and load the `ggplot2` package:

```{r}
# install.packages("ggplot2")
library(ggplot2)
# install.packages("ggthemes")
library(ggthemes)
```

Use the R base graphics to create a scatterplot of Time versus Shots:

```{r, fig.width = 5, fig.height = 4}
ggplot(worldcup, aes(x = Time, y = Shots)) + 
        geom_point()
```

Use some of the plotting options to improve the graph's appearance:

```{r, fig.width = 5, fig.height = 4}
ggplot(worldcup, aes(x = Time, y = Shots)) + 
  geom_point() + 
  xlab("Minutes played") + 
  ylab("Shots on goal") + 
  ggtitle("Time played vs. shots taken\n World Cup 2010")
```

### Fancier graphs

In this section, we'll work on creating a plot like this:

```{r, fig.width = 8, fig.height = 2.5, echo = FALSE}
most_shots <- worldcup[which.max(worldcup$Shots), ]

top_four <- c("Spain", "Germany", "Uruguay", "Netherlands")
worldcup$top_four <- factor(worldcup$Team %in% top_four,
                            levels = c(TRUE, FALSE),
                            labels = c("Top 4", "Other"))

worldcup$Position <- factor(worldcup$Position,
                            levels = c("Goalkeeper",
                                       "Defender", 
                                       "Midfielder",
                                       "Forward"))

ggplot(worldcup, aes(x = Time, y = Shots, color = top_four)) + 
        geom_point(size = 1.8, alpha = 0.7) + 
        facet_grid(. ~ Position) + 
        scale_x_continuous("Time played in World Cup (minutes)",
                           breaks = 180 * 0:7) +
        theme_few() + 
        geom_text(data = most_shots,
                  aes(label = paste0(Player, ",", Team, " ")),
                  colour = "black", size = 3,
                  hjust = 1, vjust = 0.4) +
        scale_color_discrete(name = "Team's final\n ranking")

data(worldcup)
worldcup <- worldcup %>%
  mutate(Player = rownames(worldcup))
```

Try the following tasks:

- First, before you start coding, talk with your group members about how this graph is different from the simple one you created with `ggplot` in the last section. Also discuss what you can figure out from this new graph that was less clear from the simpler one you created before.
- Use the `xlab()` function to make a clearer title for the x-axis. (You may have already written this code in the last section of this exercise.)
- Often, in graphs with a lot of points, it's hard to see some of the points, because they overlap other points. Three strategies to address this are: (a) make the points smaller; (b) make the points somewhat transparent; and (c) jitter the points. Try doing the first two with the simple ggplot scatterplot you created in the previous section of Shots by Time.
- Create a new column in the `worldcup` data called `top_four` that specifies whether or not the `Team` for that observation was one of the top four teams in the tournament (Netherlands, Uruguay, Spain, and Germany). Make the colors of the points correspond to whether the team was a top-four team.
- Create small multiples. The relationship between time played and shots taken is probably different by the players' positions. Use faceting to create different graphs for each position.
- Make order count: What order are the faceted graphs currently in? Offensive players have more chances to take shots than defensive players, so that might be a useful ordering for the facets. Re-order the `Position` factor column to go from nearest your own goal to nearest the opponents goal, and then re-plot the graph from the previous step.
- Highlighting interesting data: Who had the most shots in the 2010 World Cup? Was he on a top-four team? Use `geom_text()` to label his point on the graph with his name.
- Increase data density: Try changing the theme, to come up with a graph with a bit less non-data ink. From the `ggthemes` package (you'll need to install it if you don't already have it), try some of the following themes: `theme_few()`, `theme_tufte()`, `theme_stata()`, `theme_fivethirtyeight()`, `theme_economist_white()`,
`theme_wsj()`.

#### Example R code

As a reminder, here's the code to do a simple scatterplot ot Shots by Time for the `worldcup` data:

```{r fig.width = 3, fig.height = 3}
ggplot(worldcup, aes(x = Time, y = Shots)) +
        geom_point()
```

To add a clearer x-axis label that the current one, use `xlab()`:

```{r fig.width = 3, fig.height = 3}
ggplot(worldcup, aes(x = Time, y = Shots)) +
        geom_point() + 
        xlab("Time played in World Cup (minutes)")
```

To make the points smaller, use the `size` option in `geom_point()` (smaller than about 2 = smaller than default, larger than about 2 = larger than default):

```{r fig.width = 3, fig.height = 3}
ggplot(worldcup, aes(x = Time, y = Shots)) +
        geom_point(size = 1)  + 
        xlab("Time played in World Cup (minutes)")
```

To make the points semi-transparent, use the `alpha` option in `geom_point()` (closer to 0 = more tranparent, closer to 1 = more opaque):

```{r fig.width = 3, fig.height = 3}
ggplot(worldcup, aes(x = Time, y = Shots)) +
        geom_point(alpha = 0.5)  + 
        xlab("Time played in World Cup (minutes)")
```

To jitter the points some, use the `position = "jitter"` option in `geom_point()`:

```{r fig.width = 3, fig.height = 3}
ggplot(worldcup, aes(x = Time, y = Shots)) +
        geom_point(size = 1.5, position = "jitter",
                   alpha = 0.5)  + 
        xlab("Time played in World Cup (minutes)")
```

As an alternative, you could also jitter the points by using `geom_jitter` rather than `geom_point`:

```{r fig.width = 3, fig.height = 3}
ggplot(worldcup, aes(x = Time, y = Shots)) +
        geom_jitter(size = 1.5, alpha = 0.5, width = 0.25)  + 
        xlab("Time played in World Cup (minutes)")
```

```{block type = "rmdnote"}
"Jittering" the points means adding some extra random noise in either the x- or y-direction, or in both directions. This technique can be particularly useful when you are trying to plot points for which one dimension is categorical, rather than continuous. You can specify which direction (x and / or y) is jittered, as well as the amount of noise to add. To find out more, check out the helpfile for the `geom_jitter` function. 
```

To create a new column called `top_four`, first create vector that lists those top four teams, then create a logical vector in the dataframe for whether the team for that observation is in one of the top four teams:

```{r}
worldcup <- worldcup %>%
  mutate(top_four = Team %in% c("Spain", "Germany",  
                                "Uruguay", "Netherlands"))
summary(worldcup$top_four)
```

To color points by this variable, use `color = ` in the `aes()` part of the `ggplot()` call:

```{r fig.width = 5, fig.height = 3}
ggplot(worldcup, aes(x = Time, y = Shots,
                     color = top_four)) +
        geom_point(size = 1.5, position = "jitter",
                   alpha = 0.5)  + 
        xlab("Time played in World Cup (minutes)")
```

To create nicer labels for the legend for color, convert the `top_four` column into the factor class, with the labels you want to use in the figure legend:

```{r}
worldcup <- worldcup %>%
  mutate(top_four = factor(top_four, levels = c(TRUE, FALSE),
                            labels = c("Top 4", "Other")))
summary(worldcup$top_four)
```

```{r fig.width = 5, fig.height = 3}
ggplot(worldcup, aes(x = Time, y = Shots,
                     color = top_four)) +
        geom_point(size = 1.5, position = "jitter",
                   alpha = 0.5)  + 
        xlab("Time played in World Cup (minutes)")
```

As a note, you can use the `scale_color_discrete()` function to put in a nicer legend title:

```{r fig.width = 5, fig.height = 3}
ggplot(worldcup, aes(x = Time, y = Shots,
                     color = top_four)) +
        geom_point(size = 1.5, position = "jitter",
                   alpha = 0.5)  + 
        xlab("Time played in World Cup (minutes)") + 
        scale_color_discrete(name = "Team's final\n ranking")
```

To create small multiples, use the `facet_grid()` command:

```{r fig.width = 10, fig.height = 3}
ggplot(worldcup, aes(x = Time, y = Shots,
                     color = top_four)) +
        geom_point(size = 1.5, position = "jitter",
                   alpha = 0.5)  + 
        xlab("Time played in World Cup (minutes)") + 
        scale_color_discrete(name = "Team's final\n ranking") + 
        facet_grid(. ~ Position)
```

To re-order the `Position` column of the dataframe, use the `levels` option of the `factor()` function. This re-sets how R saves the order of the levels of this factor.

```{r}
worldcup <- worldcup %>%
  mutate(Position = factor(Position, 
                           levels = c("Goalkeeper", "Defender",
                                      "Midfielder", "Forward")))
levels(worldcup$Position)
```

```{block type = "rmdnote"}
Note from this code example that you can use the `levels` function to find out the levels and their order for a factor-class vector.
```

Then use the same code from before for your plot:

```{r fig.width = 10, fig.height = 3}
ggplot(worldcup, aes(x = Time, y = Shots,
                     color = top_four)) +
        geom_point(size = 1.5, position = "jitter",
                   alpha = 0.5)  + 
        xlab("Time played in World Cup (minutes)") + 
        scale_color_discrete(name = "Team's final\n ranking") + 
        facet_grid(. ~ Position)
```

You can use the `filter` function with a logical statement comparing `Shots` to the maximum value of `Shots` (`max(Shots)`) to filter down to the row or rows of the player or players with the most shots:

```{r}
most_shots <- worldcup %>%
  filter(Shots == max(Shots))
most_shots
```

Use `geom_text()` to label his point on the graph with his name. You may need to mess around with some of the options in `geom_text()`, like `size`, `hjust`, and `vjust` (`hjust` and `vjust` say where, in relation to the point location, to put the label), to get something you're happy with. Also, I pasted on an extra space at the end of the player's name, to add some padding so the label wouldn't be right on top of the point.

```{r fig.width = 10, fig.height = 3}
ggplot(worldcup, aes(x = Time, y = Shots,
                     color = top_four)) +
        geom_point(size = 1.5, position = "jitter",
                   alpha = 0.5)  + 
        xlab("Time played in World Cup (minutes)") + 
        scale_color_discrete(name = "Team's final\n ranking") + 
        facet_grid(. ~ Position) + 
        geom_text(data = most_shots,
                  aes(label = paste(Player, " ")),
                  colour = "black", size = 3,
                  hjust = 1, vjust = 0.4)
```

Try out different themes for the plot. First, I'll save everything we've done so far as the object `shot_plot`, then I'll try adding different themes:

```{r fig.width = 10, fig.height = 3}
shot_plot <- ggplot(worldcup, aes(x = Time, y = Shots,
                     color = top_four)) +
        geom_point(size = 1.5, position = "jitter",
                   alpha = 0.5)  + 
        xlab("Time played in World Cup (minutes)") + 
        scale_color_discrete(name = "Team's final\n ranking") + 
        facet_grid(. ~ Position) + 
        geom_text(data = most_shots,
                  aes(label = paste(Player, " ")),
                  colour = "black", size = 3,
                  hjust = 1, vjust = 0.4)

shot_plot + theme_few()
shot_plot + theme_tufte()
shot_plot + theme_wsj()
shot_plot + theme_fivethirtyeight()
shot_plot + theme_stata()
shot_plot + theme_economist_white()
```

### Data visualization cheatsheet

RStudio comes with some excellent cheatsheets, which provide quick references to functions and code you might find useful for different tasks. For this part of the group exercise, you'll explore their cheatsheet for data visualization, both to learn some new `ggplot2` code and to become familiar with how to use this cheatsheet as you do your own analysis. 

- Open the data visualization cheatsheet. You can do this from RStudio by going to "Help" -> "Cheatsheets" -> "Data Visualization with ggplot2".
- Notice that different sections give examples with some datasets that come with either base R or ggplot2. For example, under the "Graphical Primitives" section, there is code defining the object `a` as a ggplot object using the "seals" dataset: `a <- ggplot(seals, aes(x = long, y = lat))`. 
- Go through the cheatsheet and list all of the example datasets that are used in this cheatsheet. Open their helpfiles to learn more about the data. 
- Create the example datasets `a` through `l` and `s` through `t` using the code given on the cheatsheet. 
- Pick at least one example to try out from  each of the following sections: "Graphical Primitives", "One Variable", at least three subsections of "Two Variables", "Three Variables", "Scales", "Faceting", and "Position Adjustments". As you try these, try to figure out any aesthetics that you aren't familiar with (e.g., `ymin`, `ymax`). Also, use helpfiles for the geoms to look up parameters you aren't familiar with (e.g., `stat` for `geom_area`). If you can't figure out how to interpret a plot, check the helpfile for the associated geom. **Note**: For the `n` geom used in "scales", it should be defined as `n <- d + geom_bar(aes(fill = fl))`. 

#### Example R code

The code for opening the helpfiles for the example datasets is:

```{r eval = FALSE}
?seals
?economics
?mpg
?diamonds
?USArrests
```

Note that, for `USArrests`, only some of the columns are pulled out (e.g., `murder = USArrests$murder`) to use in the `data` example dataframe. Further, the "Visualizing error" examples use a dataframe created specifically for these examples, called `df`. 

```{block type = "rmdnote"}
Some of the base R and ggplot2 example datasets have gotten to be fairly well-known. Some that you'll see very often in examples are the `iris`, `mpg`, and `diamonds` datasets.
```

All of the code to create the datasets `a` through `l` and `s` through `t` is given somewhere on the cheatsheet. Here it is in full: 

```{r message = FALSE}
a <- ggplot(seals, aes(x = long, y = lat))
b <- ggplot(economics, aes(date, unemploy))
c <- ggplot(mpg, aes(hwy))
d <- ggplot(mpg, aes(fl))
e <- ggplot(mpg, aes(cty, hwy))
f <- ggplot(mpg, aes(class, hwy))
g <- ggplot(diamonds, aes(cut, color))
h <- ggplot(diamonds, aes(carat, price))
i <- ggplot(economics, aes(date, unemploy))
df <- data.frame(grp = c("A", "B"), fit = 4.5, se = 1:2)
j <- ggplot(df, aes(grp, fit, ymin = fit - se, ymax = fit + se))
data <- data.frame(murder = USArrests$Murder,
                   state = tolower(rownames(USArrests)))
map <- map_data("state")
k <- ggplot(data, aes(fill = murder))
seals$z <- with(seals, sqrt(delta_long^2 + delta_lat^2))
l <- ggplot(seals, aes(long, lat))
s <- ggplot(mpg, aes(fl, fill = drv))
t <- ggplot(mpg, aes(cty, hwy)) + geom_point()
```

Notice that, in some places, the aesthetics are defined using the full aesthetic name-value pair (e.g., `aes(x = long, y = lat)`), while in other places the code relies on position for defining which column of a dataframe maps to which aesthetic (e.g., `aes(cty, hwy)` or `aes(fl)`). Either is fine, although relying on position can result in errors if you are not very familiar with the order in which parameters are defined for a function.

This code will vary based on the examples you try, but here is some code for one set of examples:

```{r eval = FALSE}
b + geom_ribbon(aes(ymin = unemploy - 900, ymax = unemploy + 900))
c + geom_dotplot()
f + geom_violin(scale = "area")
h + geom_hex()
j + geom_pointrange()
k + geom_map(aes(map_id = state), map = map) + 
  expand_limits(x = map$long, y = map$lat)
l + geom_contour(aes(z = z))
n <- d + geom_bar(aes(fill = fl))
n + scale_fill_brewer(palette = "Blues")
o <- c + geom_dotplot(aes(fill = ..x..))
o + scale_fill_gradient(low = "red", high = "yellow")
t + facet_grid(year ~ fl)
s + geom_bar(position = "fill")
```

