# Reporting results #2a

[Download](https://github.com/geanders/RProgrammingForResearch/raw/master/slides/CourseNotes_Week9a.pdf) a pdf of the lecture slides covering this topic.

```{r echo = FALSE, message = FALSE, warning = FALSE}
knitr::opts_knit$set(error = TRUE)
```

## Example data

This week, we'll be using some example data from NOAA's Storm Events Database. 
This data lists major weather-related storm events during 2017.
For each event, it includes information like the start and end dates, where it
happened, associated deaths, injuries, and property damage, and some other
characteristics.

See the in-course
exercises for this week for more on getting and cleaning this data. As part of the in-course
exercise, you'll be making the following plot and saving it as the object `storm_plot`: 

```{r echo = FALSE, message = FALSE, warning = FALSE, fig.width = 6, fig.height = 3}
library(readr)
library(dplyr)
library(lubridate)
library(stringr)
library(tidyr)
library(ggplot2)

storms_2017 <- read_csv("data/StormEvents_details-ftp_v1.0_d2017_c20180918.csv",
                        progress = FALSE)
storms_2017 <- storms_2017 %>% 
  select(BEGIN_DATE_TIME, END_DATE_TIME, 
         EPISODE_ID:STATE_FIPS, EVENT_TYPE:CZ_NAME, SOURCE,
         BEGIN_LAT:END_LON) %>% 
  mutate(BEGIN_DATE_TIME = dmy_hms(BEGIN_DATE_TIME),
         END_DATE_TIME = dmy_hms(END_DATE_TIME),
         STATE = str_to_title(STATE),
         CZ_NAME = str_to_title(CZ_NAME)) %>% 
  filter(CZ_TYPE == "C") %>% 
  select(-CZ_TYPE) %>% 
  mutate(STATE_FIPS = str_pad(STATE_FIPS, 2, side = "left", pad = "0"),
         CZ_FIPS = str_pad(CZ_FIPS, 3, side = "left", pad = "0")) %>% 
  unite(fips, STATE_FIPS, CZ_FIPS, sep = "") %>% 
  rename_all(funs(str_to_lower(.)))

data("state")
us_state_info <- data_frame(state = state.name, 
                            area = state.area,
                            region = state.region)
state_storms <- storms_2017 %>% 
  group_by(state) %>% 
  count() %>% 
  ungroup() %>% 
  right_join(us_state_info, by = "state")

storm_plot <- ggplot(state_storms, aes(x = area, y = n)) + 
  geom_point(aes(color = region)) + 
  labs(x = "Land area (square miles)", 
       y = "# of storm events in 2017")
storm_plot
```

We'll be using the data and this plot in the next sections.

## `ggplot2` extras and extensions

### `scales` package

The `scales` package gives you a few more options for labeling with 
your `ggplot` scales. For example, if you wanted to change the notation
for the axes in the plot of state area versus number of storm events, 
you could use the `scales` package to add commas to the numeric axis values.

For the rest of these slides, I've saved the `ggplot` object with out plot to
the object named `storm_plot`, so we don't have to repeat that code every time.

```{r warning = FALSE, message = FALSE, fig.width = 6, fig.height = 3}
library(scales)
storm_plot + 
  scale_x_continuous(labels = comma) + 
  scale_y_continuous(labels = comma)
```

The `scales` package also includes labeling functions for: 

- dollars (`labels = dollar`)
- percent (`labels = percent`)

### `ggplot2` extensions

The `ggplot2` framework is set up so that others can create packages that "extend"
the system, creating functions that can be added on as layers to a ggplot object. 
Some of the types of extensions available include: 

- More themes
- Useful additions (things that you may be able to do without the package, but that
the package makes easier)
- Tools for plotting different types of data

There is a gallery with links to `ggplot2` extensions at https://www.ggplot2-exts.org.
This list may not be exhaustive---there may be other extensions on CRAN or on GitHub
that the package maintainer did not submit for this gallery. 

### More `ggplot2` themes

You have already played around a lot with using `ggplot` themes to change how your
graphs look. 
Several people have created packages with additional themes: 

- `ggthemes`
- `ggthemr` 
- `ggtech`
- `ggsci`

```{r fig.width = 12, fig.height = 7, message = FALSE, warning = FALSE}
library(ggthemes)
library(gridExtra)

a <- storm_plot +
  theme_fivethirtyeight() + 
  ggtitle("Five Thirty Eight")
b <- storm_plot +
  theme_economist() + 
  ggtitle("Economist")
c <- storm_plot +
  theme_excel() + 
  ggtitle("Excel")
d <- storm_plot +
  theme_few() + 
  ggtitle("Stephen Few")

grid.arrange(a, b, c, d, ncol = 2)
``` 

### Other useful `ggplot2` extensions

Other `ggplot2` extensions do things you might have been able to figure out how
to do without the extension, but the extension makes it much easier to do. 
These tasks include: 

- Highlighting interesting points
- "Repelling" text labels
- Arranging plots

#### Repelling / highlighting with text labels

The first is repelling text labels.
When you add labels to points on a plot, they often overlap:

```{r fig.width = 10, fig.height = 6}
storm_plot + facet_wrap(~ region) + 
  geom_label(aes(label = state))
```

The `ggrepel` package helps make sure that these labels don't overlap:

```{r fig.width = 10, fig.height = 6}
library(ggrepel)
storm_plot + facet_wrap(~ region) + 
  geom_label_repel(aes(label = state))
```

It may be too much to label every point. Instead, you may just want to 
highlight notable point. 
You can use the `gghighlight` package to do that.

```{r warning = FALSE, message = FALSE, fig.width = 8, fig.height = 3}
library(gghighlight)
storm_plot + facet_wrap(~ region) + 
  gghighlight(area > 150000 | n > 1500, label_key = state)
```

The `gghighlight` package also works for things like histograms. For example, you could
create a dataset with the count by day-of-year of certain types of events:

```{r warning = FALSE, message = FALSE}
storms_by_month <- storms_2017 %>% 
  filter(event_type %in% c("Flood", "Flash Flood", "Heavy Rain")) %>% 
  mutate(month = month(begin_date_time, label = TRUE)) %>% 
  group_by(month, event_type) %>% 
  count() %>% 
  ungroup()
storms_by_month %>% 
  slice(1:4)
```

```{r fig.width = 7, fig.height = 3}
ggplot(storms_by_month, aes(x = month, y = n, group = event_type)) +  
  geom_bar(stat = "identity") + 
  labs(x = "Month", y = "# of events") + 
  gghighlight(max(n) > 400, label_key = event_type) +
  facet_wrap(~ event_type, ncol = 1)
```

#### Arranging plots

You may have multiple related plots you want to have as multiple panels of a
single figure. 
There are a few packages that help with this. One very good one is `patchwork`.
You need to install this from GitHub: 

```{r eval = FALSE}
devtools::install_github("thomasp85/patchwork")
```

Find out more: https://github.com/thomasp85/patchwork#patchwork

Say we want to plot seasonal patterns in events in the five counties with
the highest number of events in 2017. We can use `dplyr` to figure out these
counties:

```{r}
top_counties <- storms_2017 %>% 
  group_by(fips, state, cz_name) %>% 
  count() %>% 
  ungroup() %>% 
  top_n(5, wt = n) 
```

Then create a plot with the time patterns:

```{r}
library(forcats)
top_counties_month <- storms_2017 %>% 
  semi_join(top_counties, by = "fips") %>% 
  mutate(month = month(begin_date_time),
         county = paste(cz_name, " County, ", state, sep = "")) %>% 
  count(county, month) %>% 
  ggplot(aes(x = month, y = n)) + 
  geom_bar(stat = "identity") + 
  facet_wrap(~ fct_reorder(county, n, .fun = sum, .desc = TRUE), nrow = 2) +
  scale_x_continuous(name = "", breaks = c(1, 4, 7, 10),
                     labels = c("Jan", "Apr", "Jul", "Oct")) + 
  scale_y_continuous(name = "Frequency", breaks = c(0, 20))
```

Here's this plot: 

```{r fig.width = 6, fig.height = 3}
top_counties_month
```

Now that you have two ggplot objects (`storm_plot` and `top_counties_month`), you can use
`patchwork` to put them together:

```{r fig.height = 5, fig.width = 8, message = FALSE, warning = FALSE, fig.align = "center"}
library(patchwork)
storm_plot +
  top_counties_month + 
  plot_layout(ncol = 1, heights = c(2, 1))
```

A slightly fancier version:

```{r fig.height = 5, fig.width = 8, message = FALSE, warning = FALSE, fig.align = "center"}
(storm_plot + theme(legend.position = "top") + 
    gghighlight(n > 1500 | area > 200000, 
                label_key = state)) +
  top_counties_month + 
  plot_layout(ncol = 1, heights = c(2, 1))
```

Other packages for arranging `ggplot` objects include: 

- `gridExtra`
- `cowplot`

## In-course exercise

### Getting and cleaning the example data

This week, we'll be using some example data from NOAA's Storm Events Database.
This data lists major weather-related storm events during 2017.
For each event, it includes information like the start and end dates, where it
happened, associated deaths, injuries, and property damage, and some other
characteristics.
Each row is a separate **event**. However, often several events are grouped together
within the same **episode**. 
Some of the event types are listed by their county ID (FIPS code) ("C"), but some are listed
by a forecast zone ID ("Z"). Which ID is used is given in the column `CZ_TYPE`.

- Go to https://www1.ncdc.noaa.gov/pub/data/swdi/stormevents/csvfiles/ and download 
the bulk storm details data for 2017, in the file that starts
"StormEvents_details" and includes "d2017".
- Move this into a good directory for your current working directory and read it in
using `read_csv` from the `readr` package.
- Limit the dataframe to: the beginning and ending dates and times,
the episode ID, the event ID, the state name and
FIPS, the "CZ" name, type, and FIPS, the event type, the source, and the 
begining latitude and longitude and ending latitude and longitude
- Convert the beginning and ending dates to a "date-time" class (there should
be one column for the beginning date-time and one for the ending date-time)
- Change state and county names to title case (e.g., "New Jersey" instead of "NEW JERSEY")
- Limit to the events listed by county FIPS (`CZ_TYPE` of "C") and then remove the 
`CZ_TYPE` column
- Pad the state and county FIPS with a "0" at the beginning (hint: there's
a function in `stringr` to do this) and then unite the two columns to make
one `fips` column with the 5-digit county FIPS code
- Change all the column names to lower case (you may want to try the `rename_all`
function for this)
- There is data that comes with R on U.S. states (`data("state")`). Use that to create a dataframe
with the state name, area, and region
- Create a dataframe with the number of events per state in 2017. Merge in the state information
dataframe you just created. Remove any states that are not in the state information 
dataframe
- Create the following plot: 

```{r echo = FALSE, fig.width = 6, fig.height = 3}
library(ggplot2)
storm_plot <- ggplot(state_storms, aes(x = area, y = n)) + 
  geom_point(aes(color = region)) + 
  labs(x = "Land area (square miles)", 
       y = "# of storm events in 2017")
storm_plot
```

#### Example R code

Read in the data using `read_csv`. Here's the code I used. Yours might be a bit different, 
depending on the current name of the file and where you moved it.

```{r}
library(readr)
library(dplyr)

storms_2017 <- read_csv("data/StormEvents_details-ftp_v1.0_d2017_c20180918.csv")
```

Here's what the first few columns and rows should look like:

```{r}
storms_2017 %>% 
  select(1:3) %>% 
  slice(1:3)
```

Once you've read the data in, here's the code that I used to clean the data: 

```{r}
library(lubridate)
library(stringr)
library(tidyr)

storms_2017 <- storms_2017 %>% 
  select(BEGIN_DATE_TIME, END_DATE_TIME, 
         EPISODE_ID:STATE_FIPS, EVENT_TYPE:CZ_NAME, SOURCE,
         BEGIN_LAT:END_LON) %>% 
  mutate(BEGIN_DATE_TIME = dmy_hms(BEGIN_DATE_TIME),
         END_DATE_TIME = dmy_hms(END_DATE_TIME),
         STATE = str_to_title(STATE),
         CZ_NAME = str_to_title(CZ_NAME)) %>% 
  filter(CZ_TYPE == "C") %>% 
  select(-CZ_TYPE) %>% 
  mutate(STATE_FIPS = str_pad(STATE_FIPS, 2, side = "left", pad = "0"),
         CZ_FIPS = str_pad(CZ_FIPS, 3, side = "left", pad = "0")) %>% 
  unite(fips, STATE_FIPS, CZ_FIPS, sep = "") %>% 
  rename_all(funs(str_to_lower(.)))
```

Here's what the data looks like now: 

```{r}
storms_2017 %>% 
  slice(1:3)
```

There is data that comes with R on U.S. states (`data("state")`). Use that to create a dataframe
with the state name, area, and region:

```{r}
data("state")
us_state_info <- data_frame(state = state.name, 
                            area = state.area,
                            region = state.region)
```

Create a dataframe with the number of events per state in 2017. Merge in the state information
dataframe you just created. Remove any states that are not in the state information 
dataframe: 

```{r}
state_storms <- storms_2017 %>% 
  group_by(state) %>% 
  count() %>% 
  ungroup() %>% 
  right_join(us_state_info, by = "state")
```

To create the plot: 

Ultimately, in this group exercise, you will create a plot of state land area versus the number 
of storm events in the state:

```{r fig.width = 6, fig.height = 3}
library(ggplot2)
storm_plot <- ggplot(state_storms, aes(x = area, y = n)) + 
  geom_point(aes(color = region)) + 
  labs(x = "Land area (square miles)", 
       y = "# of storm events in 2017")
storm_plot
```

### Trying out `ggplot2` extensions

- Go back through the notes so far for this week. Pick your favorite plot that's been shown 
so far and recreate it. All the code is in the notes, but you'll need to work through it 
closely to make sure that you understand how to add code from the extension into the 
rest of the `ggplot2` code.
