# Appendix B: Homework

The following are six homework assignments for the course. 

## Homework #1

**Due date: Sept. 14**

For your first homework assignment, you'll be working through a few [swirl](http://swirlstats.com/) lessons that are relevant to the material we've covered so far. Swirl is a platform that helps you learn R **in** R - you can complete the lessons right in your R console. 

### Getting started  

First, you'll need to install the swirl package: 

```{r eval=FALSE}
install.packages("swirl")
```

Next, load the swirl package. We're going to download a course from swirl's [course repository](https://github.com/swirldev/swirl_courses) called R Programming E using the function `install_course_github`. Then call the `swirl()` function to enter the interactive platform: 

```{r eval=FALSE}
library(swirl)
uninstall_course("R_Programming_E") # Only run if you have an old version of
                                    # R_Programming_E installed
install_course_github("swirldev", "R_Programming_E")
swirl()
```

```{block, type = 'rmdnote'}
After calling `swirl()`, you may be prompted to clear your workspace variables by running `rm=(list=ls())`. Running this code will clear any variables you already have saved in your global environment. While swirl recommends that you do this, it's not necessary. 
```

### Swirl lessons 

Sign in with your name, and choose *R Programming E* when swirl asks you to choose a course. For this homework, you will need to work through the following lessons in that course (the lesson number is in parentheses): 

- Basic Building Blocks (1)
- Vectors (4)
- Missing Values (5)
- Subsetting Vectors (6)
- Logic (8)
- Looking at Data (12)
- Dates and Times (14)

Each lesson should take about 10-15 minutes, but some are much shorter. You can complete the lessons in any order you want, but you may find it easiest to start with the lowest-numbered lessons and work your way up, in the order we've listed the lessons here. 

You'll be able to get started on some of these lessons after your first day in class (Basic Building Blocks, for example), but others cover topics that we'll get to in weeks 2 and 3. Whether or not we've covered a swirl topic in class, you should be able to successfully work through the lesson. At the end of each lesson, you'll be prompted to "`inform someone about your successful completion of this lesson via email`." after answering `2` for 'Yes,' enter your full name, and enter rachel.severson@colostate.edu as the email address of the person you'd like to notify. You should be sending 7 emails in total. 

```{block, type = 'rmdnote'}
After telling swirl that you would like to send a notification email, an already-populated email should pop up with the lesson you just completed in the subject line - you just need to push send. This might not happen if you access your email through a web browser instead of an app. In this case, just send an email manually with a screenshot of the end of the lesson, and the name of the lesson you just completed. 
```

### Special swirl commands 

In the swirl environment, knowing about the following commands will be helpful:

- Within each lesson, the prompt `...` indicates that you should hit Enter to move on to the next section. 
- `play()`: temporarily exit swirl. It can be useful during a swirl lesson to play around in the R console to try things out. 
- `nxt()`: regain swirl's attention after `play()`ing around in the console. 
- `main()`: return to swirl's main menu. 
- `bye()`: exit swirl. Swirl will save your progress if you exit in the middle of a lesson. You can also hit the Esc. key to exit. (To re-enter swirl, run `swirl()`. In a new R session you will have to first load the swirl library: `library(swirl)`.)

#### For fun 

While they aren't required for class, you should consider trying out some other swirl lessons later in the course. The `Functions` lesson, as well as `lapply and sapply` and `vapply and tapply` could be particularly useful. You can also look through the [course directory](https://github.com/swirldev/swirl_courses) to see what other courses and lessons are available.  

## Homework #2

**Due date: Sept. 28**

[R Markdown homework assignment]

## Homework #3

**Due date: Oct. 12**

[`dplyr`, tidy data homework assignment]

## Homework #4

**Due date: Oct. 26**

[Advanced R Markdown homework assignment]

## Homework #5

**Due date: Nov. 9**

[Homework on functions, regular expressions, and web data assignment]

## Homework #6

**Due date: Nov. 30**

[Mapping homework assignment]