# Appendix B: Homework

The following are six homework assignments for the course. 

## Homework #1

**Due date: Sept. 14**

For your first homework assignment, you'll be working through a few [swirl](http://swirlstats.com/) lessons that are relevant to the material we've covered so far. Swirl is a platform that helps you lean R **in** R - you can complete the lessons right in your R console. 

### Getting started  

First, you'll need to install the swirl package: 

```{r eval=FALSE}
install.packages("swirl")
```

Next, load the swirl package. We're going to download a course from swirl's [course repository](https://github.com/swirldev/swirl_courses) called R Programming Alt using the function `install_from_swirl()`. Then call the `swirl()` function to enter the interactive platform: 

```{r eval=FALSE}
library(swirl)
install_from_swirl("R Programming Alt")
swirl()
```

**Note:** after calling `swirl()`, you may be promted to clear your workspace variables by running `rm=(list=ls())`. Running this code will clear any variables you already have saved in your global environment. While swirl recommends that you do this, it's not necessary. 

### Swirl lessons 

Sign in with your name, and choose the R Programming Alt course. You will need to work through the following lessons: 

- Basic Building Blocks
- Vectors 
- Missing Values 
- Subsetting Vectors 
- Logic 
- Looking at Data 
- Dates and Times 

Each lesson should take about 5-10 minutes at most. 

You'll be able to get started on some of these lessons after your first day in class (Basic Building Blocks, for example), but others cover topics that we'll get to in weeks 2 and 3. Whether or not we've covered a swirl topic in class, you should be able to succesfully work through the lesson. At the end of each lesson, you'll be prompted to "`inform somewone about your successful completion of this lesson`." after answering `1` for 'Yes,' enter your full name, and enter rachel.severson@colostate.edu as the email address of the person you'd like to notify. You should be sending 7 emails in total. 
### Special swirl commands 

In the swirl environment, knowing about the following commands will be helpful:

- `play()`: temporarily exit swirl. It can be useful during a swirl lesson to play around in the R console to try things out. 
- `nxt()`: regain swirl's attention after `play()`ing around in the console. 
- `bye()`: exit swirl. Swirl will save your progress if you exit in the middle of a lesson. (To re-enter swirl, run `swirl()`. In a new session you will have to first load the swirl library: `library(swirl)`.)


#### For fun 

While they aren't required for class, you should consider trying out some other swirl lessons later in the course. The `Functions` lesson, as well as `lapply and sapply` and `vapply and tapply` could be particularly useful. You can also look through the [course directory](https://github.com/swirldev/swirl_courses) to see what other courses and lessons are available.  

## Homework #2

**Due date: Sept. 28**

[R Markdown homework assignment]

## Homework #3

**Due date: Oct. 12**

[`dplyr`, tidy data homework assignment]

## Homework #4

**Due date: Oct. 26**

[Advanced R Markdown homework assignment]

## Homework #5

**Due date: Nov. 9**

[Homework on functions, regular expressions, and web data assignment]

## Homework #6

**Due date: Nov. 30**

[Mapping homework assignment]