--- 
title: "R Programming for Research"
subtitle: "Colorado State University, ERHS 535"
author: "Brooke Anderson and Rachel Severson"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
knit: "bookdown::render_book"
output: bookdown::gitbook
documentclass: book
bibliography: ["book.bib", "packages.bib"]
biblio-style: apalike
link-citations: yes
github-repo: rstudio/bookdown-demo
description: "This is a bookdown version of the course notes for R Programming for Research, Colorado State University, Fall 2016."
---

# Online course book, ERHS 535 {-}

```{r}
fortunes::fortune("write a book")
```

This is the online book for Colorado State University's ERHS 535 *R Programming for Research* course. This book includes course information, course notes, links to download pdfs of lecture slides, in-course exercises, homework assignments, and vocabulary lists for quizzes for this course. Because this is my first semester teaching the course with this online book, it will be evolving throughout the semester, as we get to new material. 

This work is licensed under a [Creative Commons Attribution-NonCommercial 4.0 International License](https://creativecommons.org/licenses/by-nc/4.0/).

```{r echo = FALSE}
knitr::include_app("https://brookeanderson.shinyapps.io/attendance/",
                   height = "600px")
```



