---
title: "Common Applications"
header-includes:
   - \usepackage{hyperref}
   - \definecolor{links}{HTML}{2A1B81}
   - \hypersetup{colorlinks,linkcolor=,urlcolor=links}
output:
  beamer_presentation:
    theme: metropolis
fontsize: 10pt
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.path = 'figures/',
  warning = FALSE,
  message = FALSE)
```

```{r echo = FALSE, message = FALSE, warning = FALSE}
library(knitr)
library(ggplot2)
library(dplyr)
library(tidyr)
```

# Functions

## Writing your own functions

* When to write your own function?

## Writing your own functions


* when you are writing the same code over and over again
* when you are running the same code over and over again
* when you want to make your code available to others

## Writing your own functions

Nice column names!

```{r}
nice_name <- function(x){gsub(" ", "_", tolower(x))}
```

## Writing your own functions

Make a nice(er) column name

```{r}
evil_name <- "StuPid name"

nice_name(evil_name)
```

## Writing your own functions

We can use the function in to in a tidy pipeline, just like any of the built-in functions

```{r, echo=FALSE}
set.seed(13)
df <- tibble("A B C" = LETTERS[1:26], NORM = rnorm(26, mean = 5, sd = 1))
head(df, 5)
```

## Writing your own functions

We can use the function in to in a tidy pipeline, just like any of the built-in functions

```{r}
df %>% rename_all(nice_name) %>% slice(1:5)
```

# Mutate and factors

## if_else

The `if_else()` function returns one of two options based on the result of a logical test:

```{r, eval=FALSE}
if_else(condition [test with yes/no answer],
        true [do this],
        false [do this], ...)
```

## if_else

For example:

```{r, echo=FALSE}
df <- tibble(var = c("teaspoon", "Mr. Nick", "silverspoon", "just a spoon ok!", "servingspoon", "Mr. Spoon", "tablespoon"))
df
```

## if_else

Spoon or person?

```{r}
df_type <-
df %>%
mutate(type = if_else(condition = grepl(".*spoon", var),
                     true = "spoon",
                     false = "person"))
```

## combining factors

We can make factors with the `factor()` function.

```{r}
df_factor <- df_type %>%
mutate(type_fct = factor(type))
df_factor
```

## combining factors

A factor with 6 levels

```{r}
x <- factor(c("control", "lpg", "gasifier",
              "fan_rocket", "rocket_elbow", "three_stone"))
```

## combining factors

Combine 4 of the levels

```{r}
library(forcats)

fct_collapse(x, wood = c("gasifier", "fan_rocket",
                         "rocket_elbow", "three_stone"))
```

# Loading multiple files

## Loading multiple files

Common task you might be faced with is loading multiple files in the same format

* e.g. an instrument that creates a file every day
* how might we go about this?

## Loading multiple files

First we need to know the files we want to load

* The `list.files()` function can help us here:

```{r, eval = FALSE}
list.files(
  path = ["path to directory you wish to load"],
  pattern = ["regular expression?"],
  full.names = ["full path to each file?"],
  include.dirs = ["include files in sub-folders?"]
  ...)
```

## Loading multiple files

Let's list our files

```{r, eval = TRUE}
files <- list.files(path = ".")
files[1:5]
```

## Loading multiple files

Let's try again!

```{r, eval = TRUE}
files <- list.files(path = "../data/pax_data")
files[1:5]
```

## Loading multiple files

And again!

```{r, eval = TRUE}
files <- list.files(path = "../data/pax_data",
           pattern = "^PAX.*.txt$",
           full.names = TRUE)
files[1:5]
```

## Loading multiple files

Now that we know the file names, we need a function to load the data

```{r, eval = TRUE}
library(readr)
data_file <- read_delim(files[1],
             delim = "\t",
             col_type = cols())
data_file
```

## Loading multiple files

Can up `map()` over vector of file names to read in all the data.

```{r}
library(purrr)
library(dplyr)
data_pax <- map(files, read_delim,
                delim = "\t", col_type = cols()) %>%
bind_rows()
```

## Loading multiple files

Looking at what we loaded:

```{r}
head(data_pax, 3)
```

```{r, eval = TRUE}
nrow(data_pax)
```

## Loading multiple files

We don't have to use an existing function:

```{r}
read_pax <- function(file){
read_delim(file,
           delim = "\t",
           col_type = cols()) %>%
mutate(datetime = 
  as.POSIXct(strptime(x = current_fire_datetime,
                          format = "%Y/%m/%d %T"))) %>%
rename(babs = "current_fire_Babs_bkg_fit",
      bscat = "current_fire_Bscat_bkg_fit") %>%
select(datetime, babs, bscat)
}
```

## Loading multiple files

We can use our own function the same way we use an existing function

```{r}
data_pax <- map(files, read_pax) %>%
  bind_rows()
head(data_pax, 3)
```

# Exploring parameter space

## Exploring parameter space

The `cross_*` family of functions are useful for this

```{r, eval = FALSE}
library(purrr)
cross_df(.l, ...)
```

## Exploring parameter space

Start with a list of the the parameters

```{r}
var_list <- list (distance = seq(1, 5, 1),
                  time = seq(10, 1000, 5))
```

## Exploring parameter space

Use the `cross_df()` function to create a dataframe of all the combinations

```{r}
cross_df(var_list)
```

## Exploring parameter space

Create a function to run our analysis

```{r}
library(tidyr)

speed_f <- function(distance, time){distance * time}
```

## Exploring parameter space

Use `cross_df()` and `map2()` to run our function on every combination of variables

```{r}
df <- cross_df(var_list) %>%
mutate(speed = map2(.x = distance,
                    .y = time,
                    .f = speed_f)) %>%
unnest(speed)

head(df, 3)
```

## Exploring parameter space

```{r}
library(ggplot2)

p <- ggplot(df, aes(x = distance,
                    y = time,
                    fill = speed)) +
  geom_tile() +
  theme_bw() +
  scale_fill_continuous(type = "viridis") +
  theme(legend.position = "top")
```

## Exploring parameter space

```{r, echo=FALSE}
p 
```

# Deeply nested lists

## Deeply nested lists

* download the .rds file
* add it to your working directory

https://tinyurl.com/deeplynested-list

## Deeply nested lists

Read in data

```{r}
library(readr)
result <- read_rds("../data/result.rds")
```

## Deeply nested lists

Work out what the object is

```{r}
class(result)
```

## Deeply nested lists

Use the `httr` library to extract the `content of the object`

```{r}
library(httr)
content <- content(result)
class(content)
```

## Deeply nested lists

Start making it look more like a dataframe

```{r}
library(tibble)
tbl <- as_tibble(content)
```

## Deeply nested lists

```{r, echo=FALSE}
tbl
```

## Deeply nested lists

When you see lists in dataframes, think `nesting`

```{r}
library(tidyr)
unnested <- unnest_wider(tbl, data)
```

## Deeply nested lists

```{r, echo=FALSE}
unnested
```

## Deeply nested lists

Let's take a look at one of the `sensor` cells:

```{r}
sensors_1 <- unnested$sensors[[1]]
sensors_1
```

## Deeply nested lists

When we see lists of lists or a list column think `map`

Use `map_df()` to extract the valves to a dataframe

```{r}
library(purrr)
map_df(sensors_1, magrittr::extract, c("comp", "value"))
```

## Deeply nested lists

But!!! That's just one cell of a list column

```{r}
list_of_tbl <- unnested %>%
mutate(df = map(sensors,
                 ~ .x %>% map_df(magrittr::extract,
                                 c("comp", "value"))))
```

## Deeply nested lists

Which gives us:

```{r, echo=FALSE}
list_of_tbl %>%
select(-sensors, -indices)
```

## Deeply nested lists

Use `unnest()` to flatten the list-column back to a regular column

```{r}
long_df <- list_of_tbl %>%
select(-sensors, -indices) %>%
unnest(df)
```

## Deeply nested lists

```{r, echo=FALSE}
long_df
```

## Deeply nested lists

Finally we can `pivot` the long object back to a wider format:

```{r}
wider_df <- long_df %>%
pivot_wider(names_from = "comp", values_from = "value")
```

## Deeply nested lists

```{r, echo=FALSE}
wider_df
```

## Deeply nested lists

Fix the timestamp!

```{r}
p_data <- wider_df %>% 
  mutate(timestamp = as.POSIXct(strptime(timestamp,
                              format = "%FT%H:%M:%S",
                              tz = "GMT")),
         timestamp = format(timestamp,
                            tz = "US/Mountain",
                            usetz = TRUE),
         timestamp = as.POSIXct(timestamp))
```

## Deeply nested lists

```{r, echo=FALSE}
p_data
```

## Deeply nested lists

Finally, we can plot the data:

```{r fig_timeseries}
p_data <- p_data %>% 
 pivot_longer(-timestamp, names_to = "var", values_to = "val")

p <- ggplot(data = p_data,
       mapping = aes(x = timestamp, y = val, color = var)) +
  geom_point() +
  geom_line() +
  theme_bw() +
  facet_wrap(~var, ncol = 1, scales = "free_y") +
  theme(legend.position = "none") +
  xlab("") + ylab("")
```

## Deeply nested lists

```{r, echo=FALSE, fig.width = 8, fig.height = 5}
p
```

