---
title: "Reporting data results #3"
author: "Brooke Anderson"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r echo = FALSE}
library(ggplot2)
library(tidyverse)
library(faraway)
data(worldcup)
```

# Other htmlWidget packages 

## `rbokeh` package

The `rbokeh` package provides an R interface to a Python interactive visualization library, Bokeh.

There is a website with many more details on using the `rbokeh` package: https://hafen.github.io/rbokeh/ 

You can find out more about the original Python library, Bokeh, at http://bokeh.pydata.org/en/latest/. 

## `rbokeh` package

```{r fig.align = "center"}
library(rbokeh)
figure(width = 600, height = 300) %>%
  ly_points(Time, Shots, data = worldcup,
    color = Position, hover = list(Time, Shots))
```

## `rbokeh` package

```{r echo = FALSE}
# prepare data
elements <- subset(elements, !is.na(group))
elements$group <- as.character(elements$group)
elements$period <- as.character(elements$period)

# add colors for groups
metals <- c("alkali metal", "alkaline earth metal", "halogen",
  "metal", "metalloid", "noble gas", "nonmetal", "transition metal")
colors <- c("#a6cee3", "#1f78b4", "#fdbf6f", "#b2df8a", "#33a02c",
  "#bbbb88", "#baa2a6", "#e08e79")
elements$color <- colors[match(elements$metal, metals)]
elements$type <- elements$metal

# make coordinates for labels
elements$symx <- paste(elements$group, ":0.1", sep = "")
elements$numbery <- paste(elements$period, ":0.8", sep = "")
elements$massy <- paste(elements$period, ":0.15", sep = "")
elements$namey <- paste(elements$period, ":0.3", sep = "")

# create figure
p <- figure(title = "Periodic Table", tools = c("resize", "hover"),
  ylim = as.character(c(7:1)), xlim = as.character(1:18),
  xgrid = FALSE, ygrid = FALSE, xlab = "", ylab = "",
  height = 445, width = 800) %>%

# plot rectangles
ly_crect(group, period, data = elements, 0.9, 0.9,
  fill_color = color, line_color = color, fill_alpha = 0.6,
  hover = list(name, atomic.number, type, atomic.mass,
    electronic.configuration)) %>%

# add symbol text
ly_text(symx, period, text = symbol, data = elements,
  font_style = "bold", font_size = "10pt",
  align = "left", baseline = "middle") %>%

# add atomic number text
ly_text(symx, numbery, text = atomic.number, data = elements,
  font_size = "6pt", align = "left", baseline = "middle") %>%

# add name text
ly_text(symx, namey, text = name, data = elements,
  font_size = "4pt", align = "left", baseline = "middle") %>%

# add atomic mass text
ly_text(symx, massy, text = atomic.mass, data = elements,
  font_size = "4pt", align = "left", baseline = "middle")

p
```

## `dygraphs` package

The `dygraphs` package lets you create interactive time series plots from R using the `dygraphs` JavaScript library. \bigskip

The main function syntax is fairly straightforward. Like many of these packages, it allows piping. \bigskip

There is a website with more information on using `dygraphs` available at http://rstudio.github.io/dygraphs/index.html. 

## `dygraphs` package

For example, here is the code to plot monthly deaths from lung diseases in the UK in the 1970s. 

```{r eval = FALSE}
library(dygraphs)
lungDeaths <- cbind(mdeaths, fdeaths)
dygraph(lungDeaths) %>%
  dySeries("mdeaths", label = "Male") %>%
  dySeries("fdeaths", label = "Female")
```

## `dygraphs` package

For example, here is the code to plot monthly deaths from lung diseases in the UK in the 1970s. 

```{r echo = FALSE, fig.height = 3.5}
library(dygraphs)
lungDeaths <- cbind(mdeaths, fdeaths)
dygraph(lungDeaths) %>%
  dySeries("mdeaths", label = "Male") %>%
  dySeries("fdeaths", label = "Female")
```

## `DT` package

The `DT` package provides a way to create interactive tables in R using the JavaScript `DataTables` library. \bigskip

We've already seen some examples of this output in some of the Shiny apps I showed last week. You can also use this package to include interactive tables in R Markdown documents you plan to render to HTML. \bigskip

There is a website with more information on this package at http://rstudio.github.io/DT/. 

## `DT` package

```{r out.height = 200, out.width = 400, fig.width = 3}
library(DT)
datatable(worldcup)
```


## `networkD3` package

The `networkd3` package allows you to create different networks. For example, a simple network: 

```{r eval = FALSE}
library(networkD3)
src <- c("A", "A", "A", "A",
        "B", "B", "C", "C", "D")
target <- c("B", "C", "D", "J",
            "E", "F", "G", "H", "I")
networkData <- data.frame(src, target)
simpleNetwork(networkData)
```

## `networkD3` package

```{r echo = FALSE, fig.height = 4}
library(networkD3)
networkData <- tibble(str = c("A", "A", "A", "A",
                              "B", "B", "C", "C", "D"),
                      target = c("B", "C", "D", "J",
                                 "E", "F", "G", "H", "I"))
simpleNetwork(networkData)
```

## `networkD3` package

A more complex network: 

```{r eval = FALSE}
data(MisLinks)
data(MisNodes)
forceNetwork(Links = MisLinks, Nodes = MisNodes,
            Source = "source", Target = "target",
            Value = "value", NodeID = "name",
            Group = "group", opacity = 0.8)
```

## `networkD3` package

```{r echo = FALSE, fig.height = 4}
# Load data
data(MisLinks)
data(MisNodes)

# Plot
forceNetwork(Links = MisLinks, Nodes = MisNodes,
            Source = "source", Target = "target",
            Value = "value", NodeID = "name",
            Group = "group", opacity = 0.8)
```

## `networkD3` package

A Sankey diagram: 

```{r eval = FALSE}
URL <- paste0(
        "https://cdn.rawgit.com/christophergandrud/networkD3/",
        "master/JSONdata/energy.json")
Energy <- jsonlite::fromJSON(URL)
sankeyNetwork(Links = Energy$links, Nodes = Energy$nodes, 
              Source = "source", Target = "target", 
              Value = "value", NodeID = "name",
              units = "TWh", fontSize = 12, 
              nodeWidth = 30)
```

## `networkD3` package

```{r echo = FALSE, fig.height = 4.5}
URL <- paste0(
        "https://cdn.rawgit.com/christophergandrud/networkD3/",
        "master/JSONdata/energy.json")
Energy <- jsonlite::fromJSON(URL)
sankeyNetwork(Links = Energy$links, Nodes = Energy$nodes, Source = "source",
             Target = "target", Value = "value", NodeID = "name",
             units = "TWh", fontSize = 12, nodeWidth = 30)
```


## Creating your own widget

If you find a JavaScript visualization library and would like to create bindings to R, you can create your own package for a new htmlWidget. \bigskip

There is advice on creating your own widget for R available at http://www.htmlwidgets.org/develop_intro.html.