---
title: 'Reporting data results #2'
output:
  beamer_presentation:
    theme: metropolis
  ioslides_presentation: default
  slidy_presentation: default
fontsize: 10pt
---

```{r echo = FALSE, message = FALSE, warning = FALSE}
library(knitr)
library(ggplot2)
library(dplyr)
library(tidyr)
```

# Example data

## Example data

This week, we'll be using some example data from NOAA's Storm Events Database. 

Go to https://www1.ncdc.noaa.gov/pub/data/swdi/stormevents/csvfiles/ and download 
the bulk storm details data for 2017, in the file that starts
"StormEvents_details" and includes "d2017".

Move this into a good directory for your current working directory and read it in
using `read_csv` from the `readr` package.


```{r echo = FALSE, message = FALSE, warning = FALSE}
library(readr)
library(dplyr)
storms_2017 <- read_csv("../data/StormEvents_details-ftp_v1.0_d2017_c20180918.csv")
```

## Example data

This data lists major weather-related storm events during 2017.

For each event, it includes information like the start and end dates, where it
happened, associated deaths, injuries, and property damage, and some other
characteristics.

## Example data

\small

```{r}
colnames(storms_2017)
```

## Example data

Each row is a separate **event**. However, often several events are grouped together
within the same **episode**. 

```{r}
nrow(storms_2017)
length(unique(storms_2017$EVENT_ID))
length(unique(storms_2017$EPISODE_ID))
```

## Example data

Some of the event types are listed by their county ID (FIPS code) ("C"), but some are listed
by a forecast zone ID ("Z"). Which ID is used is given in the column `CZ_TYPE`:

```{r}
summary(factor(storms_2017$CZ_TYPE))
```

## Example data

For the first in-course exercise, you will clean up this data a bit for us to 
use in the rest of the class. 

\footnotesize

- Download and read in the data
- Limit the dataframe to: the beginning and ending dates and times,
the episode ID, the event ID, the state name and
FIPS, the "CZ" name, type, and FIPS, the event type, the source, and the 
begining latitude and longitude and ending latitude and longitude
- Convert the beginning and ending dates to a "date-time" class (there should
be one column for the beginning date-time and one for the ending date-time)
- Change state and county names to title case (e.g., "New Jersey" instead of "NEW JERSEY")
- Limit to the events listed by county FIPS (`CZ_TYPE` of "C") and then remove the 
`CZ_TYPE` column
- Pad the state and county FIPS with a "0" at the beginning (hint: there's
a function in `stringr` to do this) and then unite the two columns to make
one `fips` column with the 5-digit county FIPS code
- Change all the column names to lower case (you may want to try the `rename_all`
function for this)

```{r echo = FALSE, message = FALSE, warning = FALSE}
library(lubridate)
library(stringr)
library(tidyr)

storms_2017 <- storms_2017 %>% 
  select(BEGIN_DATE_TIME, END_DATE_TIME, 
         EPISODE_ID:STATE_FIPS, EVENT_TYPE:CZ_NAME, SOURCE,
         BEGIN_LAT:END_LON) %>% 
  mutate(BEGIN_DATE_TIME = dmy_hms(BEGIN_DATE_TIME),
         END_DATE_TIME = dmy_hms(END_DATE_TIME),
         STATE = str_to_title(STATE),
         CZ_NAME = str_to_title(CZ_NAME)) %>% 
  filter(CZ_TYPE == "C") %>% 
  select(-CZ_TYPE) %>% 
  mutate(STATE_FIPS = str_pad(STATE_FIPS, 2, side = "left", pad = "0"),
         CZ_FIPS = str_pad(CZ_FIPS, 3, side = "left", pad = "0")) %>% 
  unite(fips, STATE_FIPS, CZ_FIPS, sep = "") %>% 
  rename_all(funs(str_to_lower(.)))
```

## Example data

\footnotesize

```{r}
storms_2017 %>% 
  slice(1:3)
```

## State data

There is data that comes with R on U.S. states. Use that to create a dataframe with the state name,
area, and region:

```{r}
data("state")
us_state_info <- data_frame(state = state.name, 
                            area = state.area,
                            region = state.region)
```


## Example data

Create a dataframe with the number of events per state in 2017. Merge in the state information
dataframe you just created. Remove any states that are not in the state information 
dataframe.

```{r}
state_storms <- storms_2017 %>% 
  group_by(state) %>% 
  count() %>% 
  ungroup() %>% 
  right_join(us_state_info, by = "state")
```

## Example data

Ultimately, in this group exercise, you will create a plot of state land area versus the number 
of storm events in the state:

```{r echo = FALSE, fig.width = 6, fig.height = 3}
library(ggplot2)
storm_plot <- ggplot(state_storms, aes(x = area, y = n)) + 
  geom_point(aes(color = region)) + 
  labs(x = "Land area (square miles)", 
       y = "# of storm events in 2017")
storm_plot
```

## In-course exercise

We'll now take a break to do the first part of the in-course exercise.

# `ggplot2` extras and extensions

## `scales` package

The `scales` package gives you a few more options for labeling with 
your `ggplot` scales. For example, if you wanted to change the notation
for the axes in the plot of state area versus number of storm events, 
you could use the `scales` package to add commas to the numeric axis values.

For the rest of these slides, I've saved the `ggplot` object with out plot to
the object named `storm_plot`, so we don't have to repeat that code every time.

## `scales` package

\small

```{r warning = FALSE, message = FALSE, fig.width = 6, fig.height = 3}
library(scales)
storm_plot + 
  scale_x_continuous(labels = comma) + 
  scale_y_continuous(labels = comma)
```

## `scales` package

The `scales` package also includes labeling functions for: 

- dollars (`labels = dollar`)
- percent (`labels = percent`)

## `ggplot2` extensions

The `ggplot2` framework is set up so that others can create packages that "extend"
the system, creating functions that can be added on as layers to a ggplot object. 

Some of the types of extensions available include: 

- More themes
- Useful additions (things that you may be able to do without the package, but that
the package makes easier)
- Tools for plotting different types of data

## Where to find `ggplot2` extensions

There is a gallery with links to `ggplot2` extensions at https://www.ggplot2-exts.org.

This list may not be exhaustive---there may be other extensions on CRAN or on GitHub
that the package maintainer did not submit for this gallery. 

## `ggplot2` extensions: Themes

You have already played around a lot with using `ggplot` themes to change how your
graphs look. 

Several people have created packages with additional themes: 

- `ggthemes`
- `ggthemr` 
- `ggtech`
- `ggsci`

## Some `ggthemes` themes

```{r fig.width = 12, fig.height = 7, echo = FALSE, message = FALSE, warning = FALSE}
library(ggthemes)
library(gridExtra)

a <- storm_plot +
  theme_fivethirtyeight() + 
  ggtitle("Five Thirty Eight")
b <- storm_plot +
  theme_economist() + 
  ggtitle("Economist")
c <- storm_plot +
  theme_excel() + 
  ggtitle("Excel")
d <- storm_plot +
  theme_few() + 
  ggtitle("Stephen Few")

grid.arrange(a, b, c, d, ncol = 2)
``` 

## `ggplot2` extensions: Useful additions

- Highlighting interesting points
- "Repelling" text labels
- Arranging plots

## `ggplot2` extensions: repelling text labels

When you add labels to points on a plot, they often overlap:

```{r fig.width = 10, fig.height = 6}
storm_plot + facet_wrap(~ region) + 
  geom_label(aes(label = state))
```

## `ggplot2` extensions: repelling text labels

```{r fig.width = 10, fig.height = 6}
library(ggrepel)
storm_plot + facet_wrap(~ region) + 
  geom_label_repel(aes(label = state))
```

## `gghighlight` package

It may be too much to label every point. Instead, you may just want to 
highlight notable point. 

You can use the `gghighlight` package to do that.

```{r warning = FALSE, message = FALSE, fig.width = 8, fig.height = 3}
library(gghighlight)
storm_plot + facet_wrap(~ region) + 
  gghighlight(area > 150000 | n > 1500, label_key = state)
```

## `gghighlight` package

The `gghighlight` package also works for things like histograms. For example, you could
create a dataset with the count by day-of-year of certain types of events:

\footnotesize

```{r warning = FALSE, message = FALSE}
storms_by_month <- storms_2017 %>% 
  filter(event_type %in% c("Flood", "Flash Flood", "Heavy Rain")) %>% 
  mutate(month = month(begin_date_time, label = TRUE)) %>% 
  group_by(month, event_type) %>% 
  count() %>% 
  ungroup()
storms_by_month %>% 
  slice(1:4)
```

## `gghighlight` package

\small

```{r fig.width = 7, fig.height = 3}
ggplot(storms_by_month, aes(x = month, y = n, group = event_type)) +  
  geom_bar(stat = "identity") + 
  labs(x = "Month", y = "# of events") + 
  gghighlight(max(n) > 400, label_key = event_type) +
  facet_wrap(~ event_type, ncol = 1)
```

## `ggplot2` extensions: Arranging plots

You may have multiple related plots you want to have as multiple panels of a
single figure. 

There are a few packages that help with this. One very good one is `patchwork`.

You need to install this from GitHub: 

```{r eval = FALSE}
devtools::install_github("thomasp85/patchwork")
```

Find out more: https://github.com/thomasp85/patchwork#patchwork


## `ggplot2` extensions: Arranging plots

Say we want to plot seasonal patterns in events in the five counties with
the highest number of events in 2017. We can use `dplyr` to figure out these
counties:

```{r}
top_counties <- storms_2017 %>% 
  group_by(fips, state, cz_name) %>% 
  count() %>% 
  ungroup() %>% 
  top_n(5, wt = n) 
```

## `ggplot2` extensions: Arranging plots

Then create a plot with the time patterns:

\footnotesize

```{r}
library(forcats)
top_counties_month <- storms_2017 %>% 
  semi_join(top_counties, by = "fips") %>% 
  mutate(month = month(begin_date_time),
         county = paste(cz_name, " County, ", state, sep = "")) %>% 
  count(county, month) %>% 
  ggplot(aes(x = month, y = n)) + 
  geom_bar(stat = "identity") + 
  facet_wrap(~ fct_reorder(county, n, .fun = sum, .desc = TRUE), nrow = 2) +
  scale_x_continuous(name = "", breaks = c(1, 4, 7, 10),
                     labels = c("Jan", "Apr", "Jul", "Oct")) + 
  scale_y_continuous(name = "Frequency", breaks = c(0, 20))
```

## `ggplot2` extensions: Arranging plots

Here's this plot: 

```{r fig.width = 6, fig.height = 3}
top_counties_month
```

## `ggplot2` extensions: Arranging plots

Now that you have two ggplot objects (`storm_plot` and `top_counties_month`), you can use
`patchwork` to put them together:

```{r eval = FALSE}
library(patchwork)
storm_plot +
  top_counties_month + 
  plot_layout(ncol = 1, heights = c(2, 1))
```

## `ggplot2` extensions: Arranging plots

```{r echo = FALSE, fig.height = 5, fig.width = 8, message = FALSE, warning = FALSE, fig.align = "center"}
library(patchwork)
storm_plot +
  top_counties_month + 
  plot_layout(ncol = 1, heights = c(2, 1))
```

## `ggplot2` extensions: Arranging plots

A slightly fancier version:

```{r eval = FALSE}
(storm_plot + theme(legend.position = "top") + 
    gghighlight(n > 1500 | area > 200000,
                label_key = state)) +
  top_counties_month + 
  plot_layout(ncol = 1, heights = c(2, 1))
```

## `ggplot2` extensions: Arranging plots

```{r echo = FALSE, fig.height = 5, fig.width = 8, message = FALSE, warning = FALSE, fig.align = "center"}
(storm_plot + theme(legend.position = "top") + 
    gghighlight(n > 1500 | area > 200000,  label_key = state)) +
  top_counties_month + 
  plot_layout(ncol = 1, heights = c(2, 1))
```

## `ggplot2` extensions: Arranging plots

Other packages for arranging `ggplot` objects: 

- `gridExtra`
- `cowplot`

## In-course exercise

We'll take a break now to do the second part of the in-course exercise.

## Heat maps

**Heat maps** are helpful for exploring the measurements for many variables, 
including how similar they are, by showing small boxes of color. 

In `ggplot2`, there is a `geom_tile` that is useful for creating heatmaps.

## Heat maps

For example, we might want to look at how many events of each type were reported
by each type of source. First, we can summarize the data to get this data:

\footnotesize

```{r, message = FALSE, warning = FALSE}
type_by_source <- storms_2017 %>% 
  filter(event_type %in% c("Thunderstorm Wind", "Heavy Rain", 
                           "Funnel Cloud", "Flash Flood",
                           "Flood", "Hail", "Lightning")) %>% 
  filter(source %in% c("911 Call Center", "Broadcast Media",
                       "Department of Highways",
                       "Emergency Manager", "Fire Department/Rescue",
                       "Law Enforcement", "NWS Employee", "Public",
                       "Social Media", "Trained Spotter")) %>% 
  group_by(event_type, source) %>% count() %>% ungroup()
type_by_source %>% slice(1:3)
```

## Heat maps

\footnotesize

```{r fig.width = 8, fig.height = 3.5, message = FALSE, warning = FALSE}
library(viridis); library(forcats)
ggplot(type_by_source, 
       aes(x = fct_reorder(event_type, n, .fun = sum),
           y = fct_reorder(source, n, .fun =  sum))) + 
  geom_tile(aes(fill = n)) + 
  scale_fill_viridis() + 
  labs(x = "", y = "", fill = "# of events") + 
  theme(legend.position = "bottom")
```

## Alluvial plots

The package `ggalluvial` allows you to make **alluvial plots**. 

\footnotesize

```{r fig.width = 9, fig.height = 3.5}
library(ggalluvial)
ggplot(data = type_by_source,
       aes(axis1 = event_type, axis2 = source, y = n)) + 
  scale_x_discrete(limits = c("Type", "Source"), expand = c(.1, .05)) +
  geom_alluvium(aes(fill = event_type %in% c("Flash Flood", "Flood"))) + 
  theme_minimal() + geom_stratum() + 
  geom_text(stat = "stratum", label.strata = TRUE) + 
  theme(legend.position = "top") + labs(fill = "Flood event: ")
```

## `ggridges` package

The `ggridges` package lets you plot the distribution of several levels of a factor
across another variable. 

## `ggridges` package

Create a dataframe with the timing versus the source of the storm:

\footnotesize

```{r message = FALSE, warning = FALSE}
library(ggridges)
library(forcats)
storms_by_yday <- storms_2017 %>% 
  filter(source %in% c("River/Stream Gages", "Insurance Company",
                       "Airline Pilot", "Trained Spotter",
                       "Storm Chaser", "Local Official",
                       "Official NWS Observations", "Utility Company",
                       "PostOffice", "911 Call Center")) %>%
  mutate(yday = yday(begin_date_time)) %>% 
  group_by(yday, source) %>% 
  count() %>% 
  ungroup() 
```

## `ggridges` package

\footnotesize

```{r message = FALSE, warning = FALSE, fig.height = 3, fig.width = 8}
library(viridis)
ggplot(storms_by_yday, aes(x = yday,
                            y = fct_reorder(source, yday, 
                                            .fun = median, .desc = TRUE),
                           fill = ..x..)) +
  geom_density_ridges_gradient() + 
  scale_fill_viridis(name = "Day in year", option = "B") + 
  xlim(c(0, 365)) + 
  labs(x = "Day in year", y = "")
```




# Mapping in the tidyverse

## Simple features

`sf` objects: "Simple features"

- R framework that is in active development
- There will likely be changes in the near future
- Plays very well with tidyverse functions, including `dplyr` and `ggplot2` tools

```{r message = FALSE, warning = FALSE}
library(sf)
```


## Simple features

To show simple features, we'll pull in the Colorado county boundaries from the U.S. Census. 

To do this, we'll use the `tigris` package, which accesses the Census API. We'll use the option
`class = "sf"` to read these spatial dataframes in as `sf` objects.

```{r message = FALSE, warning = FALSE, results = "hide"}
library(tigris)
co_counties <- counties(state = "CO", cb = TRUE, class = "sf")
class(co_counties)
```

## Simple features

You can think of an `sf` object as a dataframe, but with one special column called `geometry`.

\footnotesize
```{r}
co_counties %>% 
  slice(1:3)
```

## Simple features

The `geometry` column has a special class (sfc):

```{r}
class(co_counties$geometry)
```

## Simple features

You can add `sf` objects to `ggplot` objects using `geom_sf`:

\footnotesize

```{r fig.width = 6, fig.height = 4, out.width = "0.7\\textwidth", fig.align = "center"}
library(ggplot2)
ggplot() + 
  geom_sf(data = co_counties)
```


## Simple features

You can map one of the columns in the `sf` object to the fill aesthetic to make a **choropleth**:

\footnotesize

```{r fig.width = 6, fig.height = 4, out.width = "0.8\\textwidth", fig.align = "center"}
ggplot() + 
  geom_sf(data = co_counties, aes(fill = ALAND))
```

## Simple features

You can use all your usual `ggplot` tricks with this:

\footnotesize

```{r fig.width = 6, fig.height = 4, out.width = "0.7\\textwidth", fig.align = "center"}
ggplot() + 
  geom_sf(data = co_counties, aes(fill = ALAND)) + 
  scale_fill_viridis(name = "Land area", label = comma) + 
  ggtitle("Land areas of Colorado counties") + 
  theme_dark()
```

## Simple features

Because simple features are a special type of dataframe, you can also use a lot of `dplyr` tricks. 

For example, you could pull out just Larimer County, CO: 

```{r}
larimer <- co_counties %>% 
  filter(NAME == "Larimer")
larimer
```

## Simple features

\footnotesize

Note: You may need the development version of `ggplot2` for this to work.

```{r warning = FALSE, message = FALSE, fig.width = 6, fig.height = 4, out.width = "0.7\\textwidth", fig.align = "center"}
ggplot() + 
  geom_sf(data = co_counties, color = "lightgray") + 
  geom_sf(data = larimer, fill = "darkcyan") + 
  geom_sf_text(data = larimer, aes(label = NAME),
               color = "white") + 
  theme_dark() + labs(x = "", y = "")
```

## Simple features

This operability with tidyverse functions means that you should now be able to figure out how to 
create a map of the number o events listed in the NOAA Storm Events database (of those listed by
county) for each county in Colorado: 

```{r  echo = FALSE, warning = FALSE, message = FALSE, fig.width = 6, fig.height = 4, out.width = "0.7\\textwidth", fig.align = "center"}
co_event_counts <- storms_2017 %>% 
  filter(state == "Colorado") %>% 
  group_by(fips) %>% 
  count() %>% 
  ungroup()

co_county_events <- co_counties %>% 
  mutate(fips = paste(STATEFP, COUNTYFP, sep = "")) %>% 
  full_join(co_event_counts, by = "fips") %>% 
  mutate(n = ifelse(!is.na(n), n, 0))

ggplot() + 
  geom_sf(data = co_county_events, aes(fill = n)) + 
  scale_fill_viridis(name = "Number of events\n(2017)")
```


## In-course exercise

We'll now take a break for the next part of the in-course exercise. For this, create the map shown on 
the previous slide.

## In-course exercise

If you have time, try this one, too:

```{r echo = FALSE, fig.width = 8, fig.height = 3, out.width = "\\textwidth"}
co_event_counts <- storms_2017 %>% 
  filter(state == "Colorado") %>% 
  filter(event_type %in% c("Tornado", "Heavy Rain", "Hail")) %>% 
  group_by(fips, event_type) %>% 
  count() %>% 
  ungroup()

co_county_events <- co_counties %>% 
  mutate(fips = paste(STATEFP, COUNTYFP, sep = "")) %>% 
  right_join(co_event_counts, by = "fips")

ggplot() +
  geom_sf(data = co_counties, color = "lightgray") + 
  geom_sf(data = co_county_events, aes(fill = n)) + 
  scale_fill_viridis(name = "Number of events\n(2017)") + 
  theme(legend.position = "top") + 
  facet_wrap(~ event_type, ncol = 3) + 
  theme(axis.line = element_blank(),
        axis.text = element_blank(),
        axis.ticks = element_blank())
```


## Simple features

\small

```{r  eval = FALSE}
co_event_counts <- storms_2017 %>% 
  filter(state == "Colorado") %>% 
  group_by(fips) %>% 
  count() %>% 
  ungroup()

co_county_events <- co_counties %>% 
  mutate(fips = paste(STATEFP, COUNTYFP, sep = "")) %>% 
  full_join(co_event_counts, by = "fips") %>% 
  mutate(n = ifelse(!is.na(n), n, 0))

ggplot() + 
  geom_sf(data = co_county_events, aes(fill = n)) + 
  scale_fill_viridis(name = "Number of events\n(2017)")
```

## Simple features

\footnotesize

```{r eval = FALSE}
co_event_counts <- storms_2017 %>% 
  filter(state == "Colorado") %>% 
  filter(event_type %in% c("Tornado", "Heavy Rain", "Hail")) %>% 
  group_by(fips, event_type) %>% 
  count() %>% 
  ungroup()

co_county_events <- co_counties %>% 
  mutate(fips = paste(STATEFP, COUNTYFP, sep = "")) %>% 
  right_join(co_event_counts, by = "fips")

ggplot() +
  geom_sf(data = co_counties, color = "lightgray") + 
  geom_sf(data = co_county_events, aes(fill = n)) + 
  scale_fill_viridis(name = "Number of events\n(2017)") + 
  theme(legend.position = "top") + 
  facet_wrap(~ event_type, ncol = 3) + 
  theme(axis.line = element_blank(),
        axis.text = element_blank(),
        axis.ticks = element_blank())
```

## Coordinate reference system

Spatial objects can have different Coordinate Reference Systems (CRSs). CRSs can be *geographic* (e.g., WGS84, for longitude-latitude data) or *projected* (e.g., UTM, NADS83). 

There is a website that lists projection strings and can be useful in setting projection information or re-projecting data: http://www.spatialreference.org

Here is an excellent resource on projections and maps in R from Melanie Frazier: https://www.nceas.ucsb.edu/~frazier/RSpatialGuides/OverviewCoordinateReferenceSystems.pdf

## Basics of creating an `sf` object

You can create an `sf` object from a regular dataframe. 

You just need to specify: 

1. The coordinate information (which columns are longitudes and latitudes)
2. The CRS

For the CRS, if you are mapping the new `sf` object with other, existing `sf` objects, 
make sure that you use the same CRS for all `sf` objects.


## Basics of creating an `sf` object

Let's look at tornado tracks in Colorado. First, clean up the data:

```{r}
co_tornados <- storms_2017 %>% 
  filter(state == "Colorado" & 
           event_type == "Tornado") %>% 
  select(begin_date_time, event_id, begin_lat:end_lon) %>% 
  gather(key = "key", value = "value", -begin_date_time, -event_id) %>% 
  separate(key, c("time", "key")) %>% 
  spread(key = key, value = value)
```

## Basics of creating an `sf` object

```{r}
co_tornados %>% 
  slice(1:5)
```

## Basics of creating an `sf` object

Change to an `sf` object:

\small

```{r}
co_tornados <- st_as_sf(co_tornados, coords = c("lon", "lat"),
                        crs = 4269)
co_tornados %>% slice(1:3)
```

## Basics of creating an `sf` object

```{r out.width = "0.8\\textwidth", fig.width = 6}
ggplot() + 
  geom_sf(data = co_counties, color = "lightgray") + 
  geom_sf(data = co_tornados, aes(color = month(begin_date_time),
                                  shape = time)) + 
  scale_color_viridis(name = "Month") 
```

## Basics of creating an `sf` object

If you want to show lines instead of points, group by the appropriate ID and then 

```{r}
co_tornados <- co_tornados %>% 
  group_by(event_id) %>% 
  summarize(month = month(first(begin_date_time)),
            do_union=FALSE) %>% 
  st_cast("LINESTRING")
```
