---
title: "Reporting data results #1"
header-includes:
   - \usepackage{hyperref}
   - \definecolor{links}{HTML}{2A1B81}
   - \hypersetup{colorlinks,linkcolor=,urlcolor=links}
output: 
  beamer_presentation:
    theme: "metropolis"
fontsize: 10pt
---

```{r echo = FALSE, message = FALSE, warning = FALSE}
library("knitr")
```


```{r echo = FALSE, message = FALSE}
library("dplyr")

library("ggplot2")
library("gridExtra")
library("ggthemes")

library("faraway")
data("worldcup")

library("dlnm")
data("chicagoNMMAPS")
chic <- chicagoNMMAPS
chic_july <- chic %>%
  filter(month == 7 & year == 1995)
```

# Highlighting interesting aspects

## Highlighting

Guideline 4: **Highlight interesting aspects.** \bigskip

Consider adding elements to highlight noteworthy elements of the data. For example, in the graph on the right, the days of a major heat wave have been highlighted with a red line. \bigskip

```{r echo = FALSE, fig.height = 3, fig.width = 8}
chic_july <- subset(chic, month == 7 & year == 1995)
chic_july_ref <- filter(chic, month == 7 & 
                        year %in% c(1990:1994,
                                    1996:2000)) %>%
        summarise(mean = mean(death),
                  min = min(death),
                  max = max(death))
ref_points <- data.frame(date = c(-2, 33, 33, -2),
        death = c(rep(chic_july_ref$max, 2),
                  rep(chic_july_ref$min, 2)))
hw <- data.frame(date = c(12, 16, 16, 12),
                 death = c(425, 425, 0, 0))
        
a <- ggplot(chic_july, aes(x = as.POSIXlt(date)$mday,
                           y = death)) + 
        xlab("Day in July 1995") + 
        ylab("All-cause deaths") + 
        ylim(0, 450) + 
        theme_tufte() + 
        geom_polygon(aes(x = date, y = death), 
                     data = ref_points,
                     color = "lightgray", 
                     alpha = 0.1) + 
        geom_line() + 
        ggtitle("1. No highlighting")

b <- ggplot(chic_july, aes(x = as.POSIXlt(date)$mday,
                           y = death)) + 
        xlab("Day in July 1995") + 
        ylab("All-cause deaths") + 
        ylim(0, 450) + 
        theme_tufte() + 
        geom_polygon(aes(x = date, y = death), 
                     data = ref_points,
                     color = "lightgray", 
                     alpha = 0.1) + 
        geom_line(aes(x = date, y = death),
                  data = hw[1:2, ],
                     color = "red",
                  size = 2) +
        geom_line() + 
        ggtitle("2. With highlighting")

grid.arrange(a, b, ncol = 2)
```

## Highlighting

Guideline 4: **Highlight interesting aspects.** \bigskip

In the below graphs, the names of the players with the most shots and passes have been added to highlight these unusual points. \bigskip

```{r echo = FALSE, message = FALSE, fig.width = 8, fig.height = 4}
ex <- subset(worldcup, Position == "Forward")
a <- ggplot(ex, aes(x = Passes, y = Shots)) + 
        geom_point(size = 1.5, alpha = 0.5) + 
        theme_few()  + 
        ggtitle("1. No highlighting")

most_shots <- ex[which.max(ex$Shots), ]
most_passes <- ex[which.max(ex$Passes), ]
b <- ggplot(ex, aes(x = Passes, y = Shots)) + 
        geom_point(size = 1.5, alpha = 0.5) + 
        theme_few()  + 
        ggtitle("2. Highlighting") + 
        geom_text(data = most_shots,
           label = paste(rownames(most_shots), ",",
                                most_shots$Team, " "),
                  colour = "blue", size = 3,
                  hjust = 1, vjust = 0.4) + 
        geom_text(data = most_passes,
           label = paste(rownames(most_passes), ",",
                                most_passes$Team, " "),
                  colour = "blue", size = 3,
                  hjust = 1, vjust = 0.4) 

grid.arrange(a, b, ncol = 2)
```

## Highlighting

One helpful way to annotate is with text, using `geom_text()`. For this, you'll first need to create a dataframe with the hottest day in the data:

```{r}
hottest_day <- chic_july %>%
  filter(temp == max(temp))
hottest_day %>% select(date:dow)
```

## Highlighting

```{r fig.height = 3, fig.width = 4, out.width = "0.7\\textwidth", fig.align = "center"}
ggplot(data = chic_july, aes(x = date, y = death)) + 
  geom_point(color = "red") + 
  geom_text(data = hottest_day, 
            label = "Max", size = 3)
```

## Highlighting

With `geom_text`, you'll often want to use position adjustment (the `position` parameter) to move the text so it won't be right on top of the data points: 

\footnotesize

```{r fig.height = 3, fig.width = 4, out.width = "0.5\\textwidth", fig.align = "center", message = FALSE}
ggplot(data = chic_july, aes(x = date, y = death)) + 
  geom_point(color = "red") + 
  geom_text(data = hottest_day, 
            label = "Max", size = 3, 
            hjust = 0, vjust = -1)
```


## Highlighting

You can also use lines to highlight. For this, it is often useful to create a new dataframe with data for the reference. To add a line for the Chicago heat wave, I've added a dataframe called `hw` with the relevant date range. I'm setting the y-value to be high enough (425) to ensure the line will be placed above the mortality data. 

```{r message = FALSE, warning = FALSE}
library(lubridate)
hw <- data.frame(date = c(ymd("1995-07-12"), 
                          ymd("1995-07-16")),
                 death = c(425, 425))
        
b <- ggplot(data = chic_july, aes(x = date, y = death)) + 
        geom_point(color = "red") + 
        geom_line(data = hw,
                  aes(x = date, y = death),
                  size = 2)
```

-----

```{r fig.height = 3, fig.width = 4, out.width = "0.7\\textwidth", fig.align = "center"}
b
```

