---
title: "Reporting data results #1"
header-includes:
   - \usepackage{hyperref}
   - \definecolor{links}{HTML}{2A1B81}
   - \hypersetup{colorlinks,linkcolor=,urlcolor=links}
output: 
  beamer_presentation:
    theme: "metropolis"
fontsize: 10pt
---

```{r echo = FALSE, message = FALSE, warning = FALSE}
library("knitr")
```

# Plot guidelines

## Guidelines for good plots

There are a number of very thoughtful books and articles about creating graphics that effectively communicate information. \medskip

Some of the authors I highly recommend (and from whose work I've pulled the guidelines for good graphics we'll talk about this week) are: 

- Edward Tufte
- Howard Wainer
- Stephen Few
- Nathan Yau

You should plan, in particular, to read *The Visual Display of Quantitative Information* by Edward Tufte before you graduate.

## Guidelines for good plots

This week, we'll focus on six guidelines for good graphics, based on the writings of these and other specialists in data display. \medskip

The guidelines are: 

1. Aim for high data density. 
2. Use clear, meaningful labels.
3. Provide useful references.
4. Highlight interesting aspects of the data. 
5. Make order meaningful. 
6. When possible, use small multiples.

## Packages for examples

For the examples, I'll use `dplyr` for data cleaning and, for plotting, the packages `ggplot2`, `gridExtra`, and `ggthemes`.

```{r message = FALSE}
library("dplyr")

library("ggplot2")
library("gridExtra")
library("ggthemes")
```

## Example data

You can load the data for today's examples with the following code: 

```{r message = FALSE}
library("faraway")
data("worldcup")
```

## Example data

```{r message = FALSE}
library("dlnm")
data("chicagoNMMAPS")
chic <- chicagoNMMAPS
chic_july <- chic %>%
  filter(month == 7 & year == 1995)
```
\footnotesize
```{r}
slice(chic_july, 1:3)
```

