---
title: "Reporting data results #3"
author: "Brooke Anderson"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# More on Leaflet

```{r echo = FALSE, results = "hide"}
library(tigris)
denver_tracts <- tracts(state = "CO", county = 31, 
                        cb = TRUE, class = "sf")

load("../data/fars_colorado.RData") # Use relative path on your computer 
library(tidyverse)
accident_data <- driver_data %>%
  select(state, st_case, county, latitude, longitud,
                date, fatals, drunk_dr) %>%
  filter(county == 31 & longitud < -104.1) %>%
  distinct()

library(leaflet)
```

## Pop-ups

You can use the `popup` option to show information when the user clicks on a marker. 

It's easiest to do this if you have the information you want to show in the dataframe with the location data. For example, we have date-time, number of fatalities, and number of drunk drivers in this data: 

```{r}
accident_data %>% 
  select(date, fatals, drunk_dr) %>% 
  slice(1:3)
```

## Pop-ups

If we want to show day of the week, month, hour, and number of fatalities, go ahead and calculate any value not already in the dataset:

```{r}
library(lubridate)
accident_data <- accident_data %>%
  mutate(weekday = wday(date, label = TRUE, abbr = FALSE),
         month = month(date, label = TRUE, abbr = FALSE),
         hour = format(date, format = "%H:%M"))
accident_data %>% slice(1:2)
```

## Pop-ups

The popup text needs to be a character vector, written in HTML syntax. You can create that column in the dataframe first, and then pass it to the `popup` argument.

```{r}
accident_data <- accident_data %>% 
  mutate(popup_info = paste0("<b>Weekday:</b>  ", 
                             weekday, "<br/>",
                             "<b>Month:</b>  ",
                             month, "<br/>",
                             "<b>Hour:</b>  ",
                             hour, "<br/>",
                             "<b>Fatalities:</b>  ",
                             fatals))
accident_data$popup_info[1:3]
```

## Pop-ups

Now pass that vector to the `popup` argument for the layer you want to pair it with:

```{r eval = FALSE}
leaflet() %>%
  addTiles() %>%
  addCircleMarkers(data = accident_data, radius = 2,
                   lng = ~ longitud, lat = ~ latitude,
                   popup = ~ popup_info)
```

## Pop-ups

```{r echo = FALSE}
leaflet() %>%
  addTiles() %>%
  addCircleMarkers(data = accident_data, radius = 2,
                   lng = ~ longitud, lat = ~ latitude,
                   popup = ~ popup_info)
```

## Pop-ups

If you aren't familiar with HTML syntax, here's one cheatsheet: http://web.stanford.edu/group/csp/cs21/htmlcheatsheet.pdf

In the pop-ups, you can use HTML to format things like color, typeface, and size. You can also add links.

## Mapping values to color

To use color to show a value, you need to do a few things. First, you need to the the `colorFactor` function (or another in its family) to create a function for mapping from values to colors. Then, you need to use this within the call to add the markers.

```{r eval = FALSE}
library(viridisLite)
pal <- colorFactor(viridis(5), accident_data$drunk_dr)
leaflet() %>%
  addTiles() %>%
  addCircleMarkers(data = accident_data, radius = 2,
                   lng = ~ longitud, lat = ~ latitude,
                   popup = ~ popup_info, 
                   color = ~ pal(drunk_dr)) 
```

## Mapping values to color

```{r echo = FALSE}
library(viridis)
pal <- colorFactor(palette = viridis(n = 5), 
                   domain = accident_data$drunk_dr)
leaflet() %>%
  addTiles() %>% 
  addCircleMarkers(data = accident_data, radius = 2,
                   lng = ~ longitud, lat = ~ latitude,
                   popup = ~ popup_info, 
                   color = ~ pal(drunk_dr)) 
```

## Mapping values to color

The `colorFactor` function (and friends) are a pretty cool type of function that actually creates a new function: 

```{r}
pal <- colorFactor(palette = viridis(n = 5), 
                   domain = accident_data$drunk_dr)
class(pal)
head(pal)
```

## Adding a legend

Once you are showing something with color, you can add a legend to explain it. You can do that with the `addLegend` function, which must include values for the color palette and values for each point from this color palette.

```{r eval = FALSE}
leaflet() %>%
  addTiles() %>%
  addCircleMarkers(data = accident_data, radius = 2,
                   lng = ~ longitud, lat = ~ latitude,
                   popup = ~ popup_info, 
                   color = ~ pal(drunk_dr)) %>%
  addLegend(data = accident_data, 
            pal = pal, values = ~ drunk_dr, 
            title = "# drunk drivers")
```

## Adding a legend

```{r echo = FALSE}
leaflet() %>%
  addTiles() %>%
  addCircleMarkers(data = accident_data, radius = 2,
                   lng = ~ longitud, lat = ~ latitude,
                   popup = ~ popup_info, 
                   color = ~ pal(drunk_dr)) %>%
  addLegend(data = accident_data, 
            pal = pal, values = ~ drunk_dr, 
            title = "# drunk drivers")
```

## Adding polygons

You can add polygons with the `addPolygons` function.

```{r eval = FALSE}
leaflet() %>%
  addTiles() %>%
  addPolygons(data = denver_tracts)
```

## Adding polygons

```{r echo = FALSE}
leaflet() %>%
  addTiles() %>%
  addPolygons(data = denver_tracts)
```

## Adding polygons

You can add pop-ups for polygons, as well.

```{r eval = FALSE}
denver_tracts <- denver_tracts %>% 
  mutate(polygon_popup = paste0("Tract ID:  ", 
                                TRACTCE))

leaflet() %>%
  addTiles() %>%
  addPolygons(data = denver_tracts, popup = ~ polygon_popup)
```

## Adding polygons

```{r echo = FALSE}
denver_tracts <- denver_tracts %>% 
  mutate(polygon_popup = paste0("Tract ID:  ", 
                                TRACTCE))

leaflet() %>%
  addTiles() %>%
  addPolygons(data = denver_tracts, popup = ~ polygon_popup)
```

## Adding polygons

You can overlay different elements. For example, you can show both accidents and tracts:

```{r eval = FALSE}
leaflet() %>%
  addTiles() %>%
  addPolygons(data = denver_tracts, popup = ~ polygon_popup,
              color = "#000000", fillColor = "969696", 
              weight = 2) %>%
  addCircleMarkers(data = accident_data, radius = 2,
                   lng = ~ longitud, lat = ~ latitude,
                   popup = ~ popup_info, opacity = 0.9,
                   color = ~ pal(drunk_dr)) %>%
  addLegend(data = accident_data, 
            pal = pal, values = ~ drunk_dr, 
            title = "# drunk drivers", opacity = 0.9)
```

## Adding polygons

```{r echo = FALSE}
leaflet() %>%
  addTiles() %>%
  addPolygons(data = denver_tracts, popup = ~ polygon_popup,
              color = "#000000", fillColor = "969696", 
              weight = 2) %>%
  addCircleMarkers(data = accident_data, radius = 2,
                   lng = ~ longitud, lat = ~ latitude,
                   popup = ~ popup_info, opacity = 0.9,
                   color = ~ pal(drunk_dr)) %>%
  addLegend(data = accident_data, 
            pal = pal, values = ~ drunk_dr, 
            title = "# drunk drivers", opacity = 0.9)
```

## Adding polygons

You can add the ability for the user to pick which layers to see using `addLayersControls`. 

```{r eval = FALSE}
leaflet() %>%
  addTiles() %>%
  addPolygons(data = denver_tracts, popup = ~ polygon_popup,
              color = "#000000", fillColor = "969696", 
              weight = 2, group = "tracts") %>%
  addCircleMarkers(data = accident_data, radius = 2,
                   lng = ~ longitud, lat = ~ latitude,
                   popup = ~ popup_info, opacity = 0.9,
                   color = pal(accident_data$drunk_dr), 
                   group = "accidents") %>%
  addLegend(data = accident_data, 
            pal = pal, values = ~ drunk_dr, 
            title = "# drunk drivers", opacity = 0.9) %>% 
  addLayersControl(baseGroups = c("base map"), 
                   overlayGroups = c("tracts", "accidents"))
```

## Adding polygons

```{r echo = FALSE}
leaflet() %>%
  addTiles() %>%
  addPolygons(data = denver_tracts, popup = ~ polygon_popup,
              color = "#000000", fillColor = "969696", 
              weight = 2, group = "tracts") %>%
  addCircleMarkers(data = accident_data, radius = 2,
                   lng = ~ longitud, lat = ~ latitude,
                   popup = ~ popup_info, opacity = 0.9,
                   color = pal(accident_data$drunk_dr), 
                   group = "accidents") %>%
  addLegend(data = accident_data, 
            pal = pal, values = ~ drunk_dr, 
            title = "# drunk drivers", opacity = 0.9) %>% 
  addLayersControl(baseGroups = c("base map"), 
                   overlayGroups = c("tracts", "accidents"))
```

## Find out more

Here are some good tutorials for trying out other examples of `leaflet` in R:

http://zevross.com/blog/2015/10/14/manipulating-and-mapping-us-census-data-in-r-using-the-acs-tigris-and-leaflet-packages-3/


