---
title: "Exploring data #1"
header-includes:
   - \usepackage{hyperref}
   - \definecolor{links}{HTML}{2A1B81}
   - \hypersetup{colorlinks,linkcolor=,urlcolor=links}
output: 
  beamer_presentation:
    theme: "metropolis"
fontsize: 10pt
---

```{r echo = FALSE}
options(width = 50)
```

```{r echo = FALSE, message = FALSE, warning = FALSE}
library(readr)
library(dplyr)
```

```{r echo = FALSE}
beijing_pm <- read_rds("../data/beijing_pm.rds")
```

# Dates in R

## `Date` class

A common task when changing or adding columns is to change the class of some of
the columns. This is especially common for dates, which will often be read in as
a character vector when reading data into R.

## `lubridate` package 

The `lubridate` package is helpful for working with vectors of dates or date-times.

You will see dates represented in many different ways. For example, October
might be included in data as "October", "Oct", or "10". Further, the way the
elements are separated can vary.

Computers are very literal, so this ambiguity can be confusing for them.

## `lubridate` package

The `lubridate` package has a number of functions for converting character
strings into dates (or date-times). To decide which one to use, you just need to
know the order of the elements of the date in the character string.

For example, here are some commonly-used `lubridate` functions:

`lubridate` function | Order of date elements
---------------------|-----------------------
`ymd`                | year-month-day
`dmy`                | day-month-year
`mdy_hm`             | month-day-year-hour-minute
`ymd_hms`            | year-month-day-hour-minute-second

(Remember, you can use `vignette("lubridate")` and `?lubridate` to get help with
the `lubridate` package.)

## `lubridate` package

In many cases you can use functions from the `lubridate` package to parse dates
pretty easily. \medskip

For example, if you have a character string with the date in the order of
*year-month-day*, you can use the `ymd` function from `lubridate` to convert the
character string to the `Date` class. For example:

```{r message = FALSE}
library("lubridate")
my_date <- ymd("2008-10-13")
class(my_date)
```

## `lubridate` package

The functions in `lubridate` are pretty good at working with different ways of
expressing date and time elements intelligently:

```{r}
mdy("10-31-2017")
dmy("31 October 2017")
```

## `lubridate` package

There are lubridate functions that can parse date-times, too:

```{r}
ymd_hms("2017/10/31--17:33:10")
mdy_hm("Oct. 31, 2017 5:33PM", tz = "MST")
```

## Converting to `Date` class

We can use the `mdy_hms` function from `lubridate` to convert the `sample_time`
column in the `beijing_pm` dataset to a date-time class ("POSIXct"):

```{r}
beijing_pm <- beijing_pm %>% 
  mutate(sample_time = mdy_hm(sample_time))

head(beijing_pm, 3)
```

## Converting to `Date` class

Once you have an object in a date or date-time class, you can do things like plot by
date, calculate the range of dates, and calculate the total number of days the
dataset covers:

```{r}
range(beijing_pm$sample_time)
diff(range(beijing_pm$sample_time))
```

## `lubridate` package

The `lubridate` package also includes functions to pull out certain elements of
a date. For example, we could use `wday` to create a new column with the weekday
of each show:

```{r}
beijing_pm <- mutate(beijing_pm, 
                     sample_weekday = wday(sample_time, 
                                           label = TRUE))
```

## `lubridate` package

```{r}
beijing_pm %>% 
  select(sample_time, sample_weekday) %>% 
  sample_n(size = 3)
```

## `lubridate` package

The `wday` function created an **ordered factor** ("ord" below the
column name in the tibble print-out). You can use this like other
factors. 

## `lubridate` package

```{r}
beijing_pm %>% 
  group_by(sample_weekday) %>% 
  summarize(mean_pm = mean(value))
```

## `lubridate` package

Other functions in `lubridate` for pulling elements from a date include: 

- `mday`: Day of the month
- `yday`: Day of the year
- `month`: Month
- `quarter`: Fiscal quarter
- `year`: Year

