---
title: "Reporting data results #3"
author: "Brooke Anderson"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r echo = FALSE}
library(ggplot2)
```

# Introduction to plotly 


## `plotly` package

From the package documentation: 

> "Easily translate `ggplot2` graphs to an interactive web-based version and / or create custom web-based visualizations directly from R."

- Like many of the packages today, draws on functionality external to R, but within a package that allows you to work exclusively within R.
- Allows you to create interactive graphs from R. The functions extend much of the `ggplot2` code you've learned.
- Interactivity will only work within RStudio or on documents rendered to HTML.

## `plotly` package

The `plotly` package allows an interface to let you work with `plotly.js` code directly using R code. \bigskip

`plotly.js` is an open source library for creating interactive graphs in JavaScript. This JavaScript library is built on `d3.js` (Data-Driven Documents), which is a key driver in interactive web-based data graphics today.

## `plotly` package

There are two main ways of create plots within `plotly`: 

- Use one of the functions to create a customized interactive graphic: 
    + `plot_ly`: Workhorse of `plotly`, renders most non-map types of graphs
    + `plot_geo`, `plot_mapbax`: Specific functions for creating `plotly` maps. 
- Create a `ggplot` object and then convert it to a `plotly` object using the `ggplotly` function.

## `plotly` package

```{r fig.width = 4, fig.height = 3.5, fig.align = "center"}
library(faraway); data(worldcup); library(dplyr)
library(plotly)
a <- worldcup %>% ggplot(aes(x = Time, y = Shots)) + geom_point()
ggplotly(a)
```

## `plotly` package

```{r fig.width = 5, fig.height = 3.5, fig.align = "center"}
a <- worldcup %>% ggplot(aes(x = Time, y = Shots, color = Position)) + 
  geom_point()
ggplotly(a)
```

## `plotly` package

You can also use this with other `ggplot2` functionality, like faceting: 

```{r eval = FALSE}
shots_vs_time <- worldcup %>%
  mutate(Name = rownames(worldcup)) %>%
  filter(Team %in% c("Netherlands", "Germany", "Spain", "Uruguay")) %>%
  ggplot(aes(x = Time, y = Shots, color = Position, group = Name)) + 
  geom_point() + 
  facet_wrap(~ Team)
ggplotly(shots_vs_time)
```

## `plotly` package

```{r echo = FALSE}
shots_vs_time <- worldcup %>%
  mutate(Name = rownames(worldcup)) %>%
  filter(Team %in% c("Netherlands", "Germany", "Spain", "Uruguay")) %>%
  ggplot(aes(x = Time, y = Shots, color = Position, group = Name)) + 
  geom_point() + 
  facet_wrap(~ Team)
ggplotly(shots_vs_time)
```


## `plotly` package

If you pipe to the `rangeslider` function, it allows the viewer to zoom in on part of the x range. (This can be particularly nice for time series.) \bigbreak

You should have a dataset available through your R session named `USAccDeaths`. This gives a monthly county of accidental deaths in the US for 1973 to 1978. This code will plot it and add a range slider on the lower x-axis.

```{r eval = FALSE}
plot_ly(x = time(USAccDeaths), y = USAccDeaths) %>% 
  add_lines() %>% rangeslider()
```

## `plotly` package

```{r echo = FALSE}
plot_ly(x = time(USAccDeaths), y = USAccDeaths) %>% 
  add_lines() %>%
  rangeslider()
```

## `plotly` package

For a 3-D scatterplot, add a mapping to the `z` variable:

```{r eval = FALSE}
worldcup %>%
  plot_ly(x = ~ Time, y = ~ Shots, z = ~ Passes,
          color = ~ Position) %>%
  add_markers()
```

## `plotly` package

```{r echo = FALSE}
worldcup %>%
  plot_ly(x = ~ Time, y = ~ Shots, z = ~ Passes,
          color = ~ Position) %>%
  add_markers()
```

## `plotly` package

The `volcano` data comes with R and is in a matrix format. Each value gives the elevation for a particular pair of x- and y-coordinates.

```{r}
dim(volcano)
volcano[1:4, 1:4]
```

## `plotly` package

```{r}
plot_ly(z = ~ volcano, type = "surface")
```

## `plotly` package

Mapping with `plotly` can build on some data that comes with base R or other packages you've likely added (or can add easily, as with the `map_data` function from `ggplot2`). For example, we can map state capitals and cities with > 40,000 people using data in the `us.cities` dataframe in the `maps` package:

```{r}
head(maps::us.cities, 3)
```


## `plotly` package

Here is code you can use to map all of these cities on a US map:

```{r eval = FALSE}
ggplot2::map_data("state") %>%
   group_by(group) %>% 
   plot_ly(x = ~long, y = ~lat) %>%
   add_polygons(hoverinfo = "none") %>%
   add_markers(text = ~paste(name, "<br />", pop), hoverinfo = "text",
               alpha = 0.25,
     data = filter(maps::us.cities, -125 < long & long < -60 &
                     25 < lat & lat < 52)) %>%
   layout(showlegend = FALSE)
```

## `plotly` package

```{r echo = FALSE}
ggplot2::map_data("world", "usa") %>%
   group_by(group) %>% filter(-125 < long & long < -60 &
                                25 < lat & lat < 52) %>%
   plot_ly(x = ~long, y = ~lat) %>%
   add_polygons(hoverinfo = "none") %>%
   add_markers(text = ~paste(name, "<br />", pop), hoverinfo = "text",
               alpha = 0.25,
     data = filter(maps::us.cities, -125 < long & long < -60 &
                     25 < lat & lat < 52)) %>%
   layout(showlegend = FALSE)
```


## `plotly` package

The creator of the R `plotly` package has written a bookdown book on the package that you can read [here](https://cpsievert.github.io/plotly_book/). It provides extensive details and examples for using `plotly`.
