---
title: "Getting / cleaning data 2"
header-includes:
   - \usepackage{hyperref}
   - \definecolor{links}{HTML}{2A1B81}
   - \hypersetup{colorlinks,linkcolor=,urlcolor=links}
output: 
  beamer_presentation:
    theme: "metropolis"
fontsize: 10pt
---

```{r echo = FALSE, message = FALSE, warning = FALSE}
library(tidyverse)
library(knitr)
library(faraway)
data(worldcup)

library("titanic")
data("titanic_train")
```

# Selecting columns using regular expressions

## Tidy select

There are [`tidyverse`](https://dplyr.tidyverse.org/reference/select.html) functions to make selecting variables more straightforwards. You can call these functions as arguments of the `select` function to streamline variable selection. Examples include: `starts_with()`, `ends_with()`, and `contains()`.

## Tidy select (helpers)

Here we use `starts_with("t")` to select all variables that begin with `t`.

```{r}
titanic_train %>%
  select(starts_with("t")) %>%
  slice(1:3)
```

## Tidy select (helpers)

Here we use `contains("ss")` to select all variables that contain `ss`.

```{r}
titanic_train %>%
  select(contains("ss")) %>%
  slice(1:3)
```

## Tidy select


The are also tidyverse functions that allow us to easily operate on a selection of variables. These functions are called [`scoped varients`](https://dplyr.tidyverse.org/reference/scoped.html). You can identity these functions by these `_all`, `_at`, and `_if` suffixes.

## Tidy select (*_at)

Here we use `select_at` to select all the variables that contain `ss` in their name:

```{r}
titanic_train %>%
  select_at(vars(contains("ss"))) %>%
  slice(1:3)
```

## Tidy select (*_at)

Here we use `select_at` to select all the variables that contain `ss` in their name and then convert their names to lower case (a handy function to tidy the variable names).

```{r}
titanic_train %>%
  select_at(vars(contains("ss")), .funs = str_to_lower) %>%
  slice(1:3)
```

## Tidy select (*_if)

Here we use `select_if` to select all the numeric variables in a dataframe:

```{r}
titanic_train %>%
  select_if(is.numeric) %>%
  slice(1:3)
```

## Tidy select (*_if)

We can also add on a function with `.funs = ` to convert their names to lower case (a handy function to tidy the variable names).

```{r}
titanic_train %>%
  select_if(is.numeric, .funs = str_to_lower) %>%
  slice(1:3)
```

## Tidy select (*_if)

The `select_if` function takes the following form.

```{r eval = FALSE}
## Generic code
new_df <- select_if(old_df,
                    .predicate [selects the variable to keep], 
                    .funs = [the function to apply to 
                             the selected column names])
```



