---
title: "Reporting data results #3"
author: "Brooke Anderson"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r echo = FALSE}
library(ggplot2)
```

# Introduction to htmlWidgets

## `htmlWidgets`

Very smart people have been working on creating interactive graphics in R for a long time. So far, nothing "native" to" R has taken off in a big way (although keep an eye out for `ggvis` and "brushing"). \bigskip

There is now a series of R packages that allow you to create plots from these JavaScript libraries from within R. \bigskip

There is a website with much more on these `htmlWidgets` at http://www.htmlwidgets.org. 

## `htmlWidgets`

JavaScript has developed a number of interactive graphics libraries that can be for documents viewed in a web browser. These work by binding data to support vector graphics (SVGs). \bigskip

They allow you to do things like zoom and pan. The graphics can also "react" to certain events. For example, they can show a pop-up when you hover over or click on a point. 


## `htmlWidgets`

Some of the packages availabe to help you create interactive graphics from R using JavaScript graphics libraries: 

- `leaflet`: Mapping 
- `dygraphs`: Time series
- `plotly`: A variety of plots, including maps
- `rbokeh`: A variety of plots, including maps
- `networkD3`: Network data
- `d3heatmap`: Heatmaps
- `DT`: Data tables
- `DiagrammeR`: Diagrams and flowcharts

## `htmlWidgets`

These packages can be used to make some pretty cool interactive visualizations for HTML output from R Markdown or Shiny (you can also render any of theme in RStudio). 

There are, however, a few limitations: 

- Written by different people. The different packages have different styles as well as different interfaces. Learning how to use one package may not help you much with other of these packages.
- Many are still in development, so some of the syntax is subject to change. 



