---
title: "Reproducible research 1"
header-includes:
   - \usepackage{hyperref}
   - \definecolor{links}{HTML}{2A1B81}
   - \hypersetup{colorlinks,linkcolor=,urlcolor=links}
output: 
  beamer_presentation:
    theme: "metropolis"
fontsize: 10pt
---

```{r echo = FALSE, message = FALSE, warning = FALSE}
knitr::knit_hooks$set(source = function(x, options){
  if (!is.null(options$verbatim) && options$verbatim){
    opts = gsub(",\\s*verbatim\\s*=\\s*TRUE\\s*", "", options$params.src)
    bef = sprintf('\n\n    ```{r %s}\n', opts, "\n")
    stringr::str_c(
      bef, 
      knitr:::indent_block(paste(x, collapse = '\n'), "    "), 
      "\n    ```\n"
    )
  } else {
    stringr::str_c("\n\n```", tolower(options$engine), "\n", 
      paste(x, collapse = '\n'), "\n```\n\n"
    )
  }
})
library(faraway)
data(nepali)
```

# R Style

## Why is style important?

**R style guidelines** provide rules for how to format code in an R script. \medskip

Some people develop their own style as they learn to code. However, it is easy to get in the habit of following style guidelines, and they offer some important advantages: \medskip

- Clean code is easier to read and interpret later.
- It's easier to catch and fix mistakes when code is clear.
- Others can more easily follow and adapt your code if it's clean.
- Some style guidelines will help prevent possible problems (e.g., avoiding `.` in function names).

## Style guidelines

For this course, we will use R style guidelines from two sources: 

- [Hadley Wickham's R style guidelines](http://adv-r.had.co.nz/Style.html)
- [Google's R style guidelines](https://google.github.io/styleguide/Rguide.xml)


## Style guideline review

Hear are a few guidelines we've already covered in class:

- Use `<-`, not `=`, for assignment.
- Guidelines for naming objects:
    + All lowercase letters or numbers
    + Use underscore (`_`) to separate words, not camelCase or a dot (`.`) (this differs for Google and Wickham style guides)
    + Have some consistent names to use for "throw-away" objects (e.g., `df`, `ex`, `a`, `b`)
- Make names meaningful
    + Descriptive names for R scripts ("random_group_assignment.R")
    + Nouns for objects (`todays_groups` for an object with group assignments)
    + Verbs for functions (`make_groups` for the function to assign groups)

## Line length

Google: **Keep lines to 80 characters or less** \medskip

To set your script pane to be limited to 80 characters, go to "RStudio" -> "Preferences" -> "Code" -> "Display", and set "Margin Column" to 80.

```{r eval = FALSE}
# Do
my_df <- data.frame(n = 1:3,
                    letter = c("a", "b", "c"),
                    cap_letter = c("A", "B", "C"))

# Don't
my_df <- data.frame(n = 1:3, letter = c("a", "b", "c"), cap_letter = c("A", "B", "C"))
```

This guideline helps ensure that your code is formatted in a way that you can see all of the code without scrolling horizontally (left and right).

## Spacing

- Binary operators (e.g., `<-`, `+`, `-`) should have a space on either side
- A comma should have a space after it, but not before.
- Colons should not have a space on either side. 
- Put spaces before and after `=` when assigning parameter arguments

```{r eval = FALSE}
# Do
shots_per_min <- worldcup$Shots / worldcup$Time
#Don't 
shots_per_min<-worldcup$Shots/worldcup$Time

#Do
ave_time <- mean(worldcup[1:10, "Time"])
#Don't
ave_time<-mean(worldcup[1 : 10 ,"Time"])
```

## Semicolons

Although you can use a semicolon to put two lines of code on the same line, you should avoid it. 

```{r eval = FALSE}
# Do
a <- 1:10
b <- 3

# Don't
a <- 1:10; b <- 3
```

## Commenting

- For a comment on its own line, use `#`. Follow with a space, then the comment. 
- You can put a short comment at the end of a line of R code. In this case, put two spaces after the end of the code, one `#`, and one more space before the comment.
- If it helps make it easier to read your code, separate sections using a comment character followed by many hyphens (e.g., `#------------`). Anything after the comment character is "muted". 

```{r eval = FALSE}
# Read in health data ---------------------------

# Clean exposure data ---------------------------
```

## Indentation

Google:

- Within function calls, line up new lines with first letter after opening parenthesis for parameters to function calls:

Example:

```{r eval = FALSE}
# Relabel sex variable
nepali$sex <- factor(nepali$sex, 
                     levels = c(1, 2),
                     labels = c("Male", "Female"))
```

## Code grouping

- Group related pieces of code together. 
- Separate blocks of code by empty spaces.

```{r eval = TRUE}
# Load data
library(faraway)
data(nepali)

# Relabel sex variable
nepali$sex <- factor(nepali$sex, 
                     levels = c(1, 2),
                     labels = c("Male", "Female"))
```

Note that this grouping often happens naturally when using tidyverse functions, since they encourage piping (`%>%` and `+`).

## Broader guidelines

- Omit needless code.
- Don't repeat yourself.

We'll learn more about satisfying these guidelines when we talk about writing your own functions in the next part of the class.

