---
title: "Exploring data 2"
header-includes:
   - \usepackage{hyperref}
   - \definecolor{links}{HTML}{2A1B81}
   - \hypersetup{colorlinks,linkcolor=,urlcolor=links}
output:
  beamer_presentation:
    theme: metropolis
fontsize: 10pt
---

```{r echo = FALSE, message = FALSE, warning = FALSE}
library(knitr)
library(ggplot2)
library(dplyr)
library(tidyr)
library(ggthemes)
library(faraway)
data(worldcup)
data(nepali)

library(tibble)
library(broom)
library("microbiome")
data("atlas1006")
tidy_samples <- atlas1006 %>% 
  get_sample() %>% 
  as.data.frame() %>% 
  rownames_to_column(var = "species") %>% 
  pivot_longer(-species, 
               names_to = "sample", 
               values_to = "prevalence")
```

# Functions

## Functions

As you move to larger projects, you will find yourself using the same code a lot. \bigskip

Examples include: 

- Reading in data from a specific type of equipment (air pollution monitor, accelerometer)
- Running a specific type of analysis (e.g., fitting the same model format to
many datasets)
- Creating a specific type of plot or map

\bigskip 

If you find yourself cutting and pasting a lot, convert the code to a function.

## Functions

Advantages of writing functions include: 

- Coding is more efficient
- Easier to change your code (if you've cut and paste code and you want to
change something, you have to change it everywhere)
- Easier to share code with others

## Functions

You can name a function anything you want, as long as you follow the naming
rules for all R objects (although try to avoid names of preexisting-existing
functions). You then specify any inputs (arguments; separate multiple arguments
with commas) and put the code to run in braces. You **define** a function as an
R object just like you do with other R objects (`<-`).

Here is the basic structure of "where things go" in an R function definition. 

```{r, eval = FALSE}
## Note: this code will not run
[function name] <- function([any arguments]){
        [code to run]
}
```

## Functions

Here is an example of a very basic function. This function takes a number as
the input (`number`) and adds 1 to that number. An R function will only return one R object. By
default, that object will be the last line of code in the function body.

```{r}
add_one <- function(number){
        number + 1 # Value returned by the function
}

add_one(number = 1:3)
add_one(number = -1)
```

## Functions

```{r eval = FALSE}
add_one <- function(number){
        number + 1 # Value returned by the function
}
```

- I picked the name of the function (`add_one`) (just like you pick what name
you want to use with any R object)
- The only input is a numeric vector. I pick the name I want to use for the
vector that is input to the function. I picked `number`.
- Within the code inside the function, the `number` refers to the numeric vector
object that the user passed into the function.

## Functions

As another example, you could write a small function to fit a specific model to
a dataframe you input and return the model object:

```{r}
fit_time_pos_mod <- function(df){
  lm(Tackles ~ Time + Position, 
     data = df) # Returns result from this call
}
```

- I picked the name of the function (`fit_time_pos_mod`) (just like you pick
what name you want to use with any R object)
- The only input is a dataframe. I pick the name I want to use for the dataframe
that is input to the function. I picked `df` (I often use this as a default
parameter name for a dataframe).
- Within the code inside the function, the `df` refers to the dataframe object
that the user passed into the function.

## Functions

Now you can apply that function within a tidy pipeline, for example to fit the
model to a specific subset of the data (the top four teams):

```{r}
data(worldcup)
worldcup %>%
  filter(Team %in% c("Spain", "Netherlands", 
                     "Uruguay", "Germany")) %>% 
  fit_time_pos_mod() %>% 
  tidy()
```

## Functions 

- Functions can input any type of R object (for example, vectors, data frames,
even other functions and ggplot objects)
- Similarly, functions can output any type of R object
- However, functions can only output one R object. If you have complex things
you want to output, a list might be a good choice for the output object type.
- Functions can have "side effects". Examples include printing something or
drawing a plot. Any action that a function takes *besides returning an R object*
is a "side effect".
