---
title: "Exploring data #1"
header-includes:
   - \usepackage{hyperref}
   - \definecolor{links}{HTML}{2A1B81}
   - \hypersetup{colorlinks,linkcolor=,urlcolor=links}
output: 
  beamer_presentation:
    theme: "metropolis"
fontsize: 10pt
---

```{r echo = FALSE}
options(width = 50)
```

```{r echo = FALSE, message = FALSE, warning = FALSE}
library(readr)
library(dplyr)
```

```{r echo = FALSE}
beijing_pm <- read_rds("../data/beijing_pm.rds")
```

# Data from R packages

## Data from R packages

So far we've covered two ways to get data into R:

1. From flat files (either on your computer or online)
2. From files like SAS and Excel

Many R packages come with their own data, which is very easy to load and use.

## Data from R packages

For example, the `faraway` package has a dataset called `worldcup` that you'll
use today. To load it, use the `data()` function once you've loaded the package
with the name of the dataset as its argument:

```{r}
library("faraway")
data("worldcup")
```

## Data from R packages

Unlike most data objects you'll work with, the data that comes with an R package
will often have its own help file. You can access this using the `?` operator:

```{r, eval = FALSE}
?worldcup
```

## Data from R packages

To find out all the datasets that are available in the packages you currently
have loaded, run `data()` without an option inside the parentheses:

```{r, eval = FALSE}
data()
```

To find out all of the datasets available within a certain package, run `data`
with the argument `package`:

```{r eval = FALSE}
data(package = "faraway")
```

As a note, you can similarly use `library()`, without the name of a package, to
list all of the packages you have installed that you could call with
`library()`:

```{r, eval = FALSE}
library()
```

