---
title: "Getting / cleaning data 2"
header-includes:
   - \usepackage{hyperref}
   - \definecolor{links}{HTML}{2A1B81}
   - \hypersetup{colorlinks,linkcolor=,urlcolor=links}
output: 
  beamer_presentation:
    theme: "metropolis"
fontsize: 10pt
---

```{r echo = FALSE, message = FALSE, warning = FALSE}
library(tidyverse)
library(knitr)
library(faraway)
data(worldcup)
```

# Tidying with `dplyr` 

## `VADeaths` data

For this example, I'll use the `VADeaths` dataset that comes with R. \bigskip

This dataset gives the death rates per 1,000 people in Virginia in 1940. It gives death rates by age, gender, and rural / urban:

```{r}
data("VADeaths")
VADeaths
```

## `VADeaths` data

There are a few things that make this data untidy: 

- One variable (age category) is saved as row names, rather than a column.
- Other variables (gender, rural / urban) are in column names.
- Once you gather the data, you will have two variables (gender, rural / urban) in the same column.

In the following slides, we'll walk through how to tidy this data. 

## Tidying the `VADeaths` data

(1) One variable (age category) is saved as row names, rather than a column. \bigskip

To fix this, we need to convert the row names into a new column. We can do this using `mutate` (load `tibble` if needed):

```{r echo = 2:4}
library(tibble)
VADeaths %>%
  as.data.frame() %>% ## Convert from matrix to dataframe
  rownames_to_column(var = "age")
```

## Tidying the `VADeaths` data

(2) Two variables (gender, rural / urban) are in column names. \bigskip

Gather the data to convert column names to a new column:

```{r}
VADeaths %>%
  as.data.frame() %>%
  rownames_to_column(var = "age") %>%
  pivot_longer(- age, names_to = "gender_loc", values_to = "mort_rate") %>%
  slice(1:4)
```

## Tidying the `VADeaths` data

(3) Two variables (gender, rural / urban) in the same column. \bigskip

Separate the column into two separate columns for "gender" and "loc" (rural / urban):

```{r}
VADeaths %>%
  as.data.frame() %>%
  rownames_to_column(var = "age") %>%
  pivot_longer(- age, names_to = "gender_loc", values_to = "mort_rate") %>%
  separate(col = gender_loc, into = c("gender", "loc"),
           sep = " ") %>%
  slice(1)
```

## Tidying the `VADeaths` data

Now that the data is tidy, it's much easier to plot: 

```{r echo = FALSE}
VADeaths <- VADeaths %>%
  as.data.frame() %>%
  rownames_to_column(var = "age") %>%
  pivot_longer(- age, names_to = "gender_loc", values_to = "mort_rate") %>%
  separate(col = gender_loc, into = c("loc", "gender"),
           sep = " ")
```

```{r fig.width = 8, fig.height = 3}
ggplot(VADeaths, aes(x = age, y = mort_rate,
                     color = gender)) + 
  geom_point() + 
  facet_wrap( ~ loc, ncol = 2) + 
  xlab("Age category") + ylab("Death rate (per 1,000)") + 
  theme_minimal()
```

