---
title: "Reporting data results #1"
header-includes:
   - \usepackage{hyperref}
   - \definecolor{links}{HTML}{2A1B81}
   - \hypersetup{colorlinks,linkcolor=,urlcolor=links}
output: 
  beamer_presentation:
    theme: "metropolis"
fontsize: 10pt
---

```{r echo = FALSE, message = FALSE, warning = FALSE}
library("knitr")
```


```{r echo = FALSE, message = FALSE}
library("dplyr")

library("ggplot2")
library("gridExtra")
library("ggthemes")

library("faraway")
data("worldcup")

library("dlnm")
data("chicagoNMMAPS")
chic <- chicagoNMMAPS
chic_july <- chic %>%
  filter(month == 7 & year == 1995)
```

# Scales

## Scales

There are a number of different functions for adjusting scales. These follow the following convention: 

```{r eval = FALSE}
## Generic code
scale_[aesthetic]_[vector type]
```

For example, to adjust the x-axis scale for a continuous variable, you'd use `scale_x_continuous`. \medskip

You can use a `scale` function for an axis to change things like the axis label (which you could also change with `xlab` or `ylab`) as well as position and labeling of breaks.

## Scales

For example, here is the default for plotting time versus passes for the `worldcup` dataset, with the number of shots taken shown by size and position shown by color:

```{r fig.width = 7, fig.height = 4, out.width = "0.8\\textwidth", fig.align = "center"}
ggplot(worldcup, aes(x = Time, y = Passes,
                     color = Position, size = Shots)) + 
  geom_point(alpha = 0.5)
```

## Scales

```{r fig.width = 7, fig.height = 4, out.width = "0.8\\textwidth", fig.align = "center"}
ggplot(worldcup, aes(x = Time, y = Passes,
                     color = Position, size = Shots)) + 
  geom_point(alpha = 0.5) + 
  scale_x_continuous(name = "Time played (minutes)", 
                     breaks = 90 * c(2, 4, 6),
                     minor_breaks = 90 * c(1, 3, 5))
```

## Scales

Parameters you might find useful in `scale` functions include: 

```{r echo = FALSE}
scale_params <- data.frame(param = c("name",
                                     "breaks",
                                     "minor_breaks",
                                     "labels",
                                     "limits"),
                           desc = c("Label or legend name",
                                    "Vector of break points",
                                    "Vector of minor break points",
                                    "Labels to use for each break",
                                    "Limits to the range of the axis"))
knitr::kable(scale_params, col.names = c("Parameter", "Description"))
```

## Scales

For dates, you can use `scale` functions like `scale_x_date` and `scale_x_datetime`. For example, here's a plot of deaths in Chicago in July 1995 using default values for the x-axis: 

```{r fig.width = 5, fig.height = 2, out.width = "0.9\\textwidth", fig.align = "center"}
ggplot(chic_july, aes(x = date, y = death)) + 
  geom_line() 
```

## Scales

And here's an example of changing the formatting and name of the x-axis:

```{r fig.width = 5, fig.height = 2, out.width = "0.9\\textwidth", fig.align = "center"}
ggplot(chic_july, aes(x = date, y = death)) + 
  geom_line() + 
  scale_x_date(name = "Date in July 1995",
               date_labels = "%m-%d")
```

## Scales

You can also use the `scale` functions to transform an axis. For example, to show the Chicago plot with "deaths" on a log scale, you can run:

```{r fig.width = 5, fig.height = 2, out.width = "0.9\\textwidth", fig.align = "center"}
ggplot(chic_july, aes(x = date, y = death)) + 
  geom_line() +
  scale_y_log10()
```

## Scales

For colors and fills, the conventions for the names of the `scale` functions can vary. \medskip 

For example, to adjust the color scale when you're mapping a discrete variable (i.e., categorical, like gender or animal breed) to color, you'd use `scale_color_hue`. To adjust the color scale for a continuous variable, like age, you'll use `scale_color_gradient`.

## Scales

For any color scales, consider starting with `brewer` first (e.g., `scale_color_brewer`).  \medskip

Scale functions from `brewer` allow you to set colors using different palettes. You can explore these palettes at http://colorbrewer2.org/.

## Scales

The Brewer palettes fall into three categories: sequential, divergent, and qualitative. You should use sequential or divergent for continuous data and qualitative for categorical data. Use `display.brewer.pal` to show the palette for a given number of colors.

```{r out.width = "0.32\\textwidth", fig.show='hold', fig.height = 3, fig.width = 4}
library("RColorBrewer")
display.brewer.pal(name = "Set1", n = 8)
display.brewer.pal(name = "PRGn", n = 8)
display.brewer.pal(name = "PuBuGn", n = 8)
```


## Scales

Use the `palette` argument within a `scales` function to customize the palette:

```{r fig.width = 8, fig.height = 2, out.width = "\\textwidth"}
a <- ggplot(data.frame(x = 1:5, y = rnorm(5),
                       group = letters[1:5]),
            aes(x = x, y = y, color = group)) + 
  geom_point()
b <- a + scale_color_brewer(palette = "Set1")
c <- a + scale_color_brewer(palette = "Pastel2") + 
  theme_dark()
grid.arrange(a, b, c, ncol = 3)
```

## Scales

```{r fig.width = 7, fig.height = 4, out.width = "0.8\\textwidth", fig.align = "center"}
ggplot(worldcup, aes(x = Time, y = Passes,
                     color = Position, size = Shots)) + 
  geom_point(alpha = 0.5) + 
  scale_color_brewer(palette = "Dark2",
                     name = "Player position")
```

## Scales

You can also set colors manually:

```{r fig.width = 7, fig.height = 4, out.width = "0.8\\textwidth", fig.align = "center"}
ggplot(worldcup, aes(x = Time, y = Passes,
                     color = Position, size = Shots)) + 
  geom_point(alpha = 0.5) + 
  scale_color_manual(values = c("blue", "red", 
                                "darkgreen", "darkgray"))
```

## Excellent references



Some excellent further references for plotting are:

- Chapter 3 of *R for Data Science* by Garrett Grolemund and Hadley Wickham (http://r4ds.had.co.nz/)
- *Data Visualization* by Kieran Healy (https://socviz.co/)
- *R Graphics Cookbook* by Winston Chang (https://r-graphics.org/)
- Google images

\bigskip

For more technical details about plotting in R:

- *ggplot2: Elegant Graphics for Data Analysis* by Hadley Wickham
- *R Graphics* by Paul Murrell
