---
title: 'Reporting data results #2'
header-includes:
   - \usepackage{hyperref}
   - \definecolor{links}{HTML}{2A1B81}
   - \hypersetup{colorlinks,linkcolor=,urlcolor=links}
output: 
  beamer_presentation:
    theme: "metropolis"
fontsize: 10pt
---

```{r echo = FALSE, message = FALSE, warning = FALSE}
library(knitr)
library(ggplot2)
library(dplyr)
library(tidyr)
```

# Example data

## Example data

This week, we'll be using some example data from NOAA's Storm Events Database. 

Go to https://www1.ncdc.noaa.gov/pub/data/swdi/stormevents/csvfiles/ and download 
the bulk storm details data for 2017, in the file that starts
"StormEvents_details" and includes "d2017".

Move this into a good directory for your current working directory and read it in
using `read_csv` from the `readr` package.


```{r echo = FALSE, message = FALSE, warning = FALSE}
library(readr)
library(dplyr)
storms_2017 <- read_csv("../data/StormEvents_details-ftp_v1.0_d2017_c20180918.csv")
```

## Example data

This data lists major weather-related storm events during 2017.

For each event, it includes information like the start and end dates, where it
happened, associated deaths, injuries, and property damage, and some other
characteristics.

## Example data

\small

```{r}
colnames(storms_2017)
```

## Example data

Each row is a separate **event**. However, often several events are grouped together
within the same **episode**. 

```{r}
storms_2017 %>% nrow()
storms_2017 %>% 
  select(EVENT_ID) %>% 
  distinct() %>% 
  nrow()
storms_2017 %>% 
  select(EPISODE_ID) %>% 
  distinct() %>% 
  nrow()
```

## Example data

Some of the event types are listed by their county ID (FIPS code) ("C"), but some are listed
by a forecast zone ID ("Z"). Which ID is used is given in the column `CZ_TYPE`:

```{r}
storms_2017 %>% 
  group_by(CZ_TYPE) %>% 
  count()
```

## Example data

For the first in-course exercise, you will clean up this data a bit for us to 
use in the rest of the class. 

\footnotesize

- Download and read in the data
- Limit the dataframe to: the beginning and ending dates and times,
the episode ID, the event ID, the state name and
FIPS, the "CZ" name, type, and FIPS, the event type, the source, and the 
begining latitude and longitude and ending latitude and longitude
- Convert the beginning and ending dates to a "date-time" class (there should
be one column for the beginning date-time and one for the ending date-time)
- Change state and county names to title case (e.g., "New Jersey" instead of "NEW JERSEY")
- Limit to the events listed by county FIPS (`CZ_TYPE` of "C") and then remove the 
`CZ_TYPE` column
- Pad the state and county FIPS with a "0" at the beginning (hint: there's
a function in `stringr` to do this) and then unite the two columns to make
one `fips` column with the 5-digit county FIPS code
- Change all the column names to lower case (you may want to try the `rename_all`
function for this)

```{r echo = FALSE, message = FALSE, warning = FALSE}
library(lubridate)
library(stringr)
library(tidyr)

storms_2017 <- storms_2017 %>% 
  select(BEGIN_DATE_TIME, END_DATE_TIME, 
         EPISODE_ID:STATE_FIPS, EVENT_TYPE:CZ_NAME, SOURCE,
         BEGIN_LAT:END_LON) %>% 
  mutate(BEGIN_DATE_TIME = dmy_hms(BEGIN_DATE_TIME),
         END_DATE_TIME = dmy_hms(END_DATE_TIME),
         STATE = str_to_title(STATE),
         CZ_NAME = str_to_title(CZ_NAME)) %>% 
  filter(CZ_TYPE == "C") %>% 
  select(-CZ_TYPE) %>% 
  mutate(STATE_FIPS = str_pad(STATE_FIPS, 2, side = "left", pad = "0"),
         CZ_FIPS = str_pad(CZ_FIPS, 3, side = "left", pad = "0")) %>% 
  unite(fips, STATE_FIPS, CZ_FIPS, sep = "") %>% 
  rename_all(funs(str_to_lower(.)))
```

## Example data

\footnotesize

```{r}
storms_2017 %>% 
  slice(1:3)
```

## State data

There is data that comes with R on U.S. states. Use that to create a dataframe with the state name,
area, and region:

```{r}
data("state")
us_state_info <- tibble(state = state.name, 
                        area = state.area,
                        region = state.region)
```


## Example data

Create a dataframe with the number of events per state in 2017. Merge in the state information
dataframe you just created. Remove any states that are not in the state information 
dataframe.

```{r}
state_storms <- storms_2017 %>% 
  group_by(state) %>% 
  count() %>% 
  ungroup() %>% 
  right_join(us_state_info, by = "state")
```

## Example data

Ultimately, in this group exercise, you will create a plot of state land area versus the number 
of storm events in the state:

```{r echo = FALSE, fig.width = 6, fig.height = 3}
library(ggplot2)
storm_plot <- ggplot(state_storms, aes(x = area, y = n)) + 
  geom_point(aes(color = region)) + 
  labs(x = "Land area (square miles)", 
       y = "# of storm events in 2017")
storm_plot
```

## In-course exercise

We'll now take a break to do the first part of the in-course exercise.

# `ggplot2` extras and extensions

## `ggplot2` extensions

The `ggplot2` framework is set up so that others can create packages that "extend"
the system, creating functions that can be added on as layers to a ggplot object. 

Some of the types of extensions available include: 

- More themes
- Useful additions (things that you may be able to do without the package, but that
the package makes easier)
- Tools for plotting different types of data

## Where to find `ggplot2` extensions

There is a gallery with links to `ggplot2` extensions at https://exts.ggplot2.tidyverse.org/gallery/.

This list may not be exhaustive---there may be other extensions on CRAN or on GitHub
that the package maintainer did not submit for this gallery. 

## `ggplot2` extensions: Themes

You have already played around a lot with using `ggplot` themes to change how your
graphs look. 

Several people have created packages with additional themes: 

- `ggthemes`
- `ggthemr` 
- `ggtech`
- `ggsci`

## Some `ggthemes` themes

```{r fig.width = 12, fig.height = 7, echo = FALSE, message = FALSE, warning = FALSE}
library(ggthemes)
library(gridExtra)

a <- storm_plot +
  theme_fivethirtyeight() + 
  ggtitle("Five Thirty Eight")
b <- storm_plot +
  theme_economist() + 
  ggtitle("Economist")
c <- storm_plot +
  theme_excel() + 
  ggtitle("Excel")
d <- storm_plot +
  theme_few() + 
  ggtitle("Stephen Few")

grid.arrange(a, b, c, d, ncol = 2)
``` 

## `scales` package

The `scales` package gives you a few more options for labeling with 
your `ggplot` scales. For example, if you wanted to change the notation
for the axes in the plot of state area versus number of storm events, 
you could use the `scales` package to add commas to the numeric axis values.

For the rest of these slides, I've saved the `ggplot` object with out plot to
the object named `storm_plot`, so we don't have to repeat that code every time.

## `scales` package

\small

```{r warning = FALSE, message = FALSE, fig.width = 6, fig.height = 3}
library(scales)
storm_plot + 
  scale_x_continuous(labels = comma) + 
  scale_y_continuous(labels = comma)
```

## `scales` package

The `scales` package also includes labeling functions for: 

- dollars (`labels = dollar`)
- percent (`labels = percent`)

## `viridis` package

The `viridis` package can be used to change to a better color scale 
(see https://www.youtube.com/watch?v=xAoljeRJ3lU&feature=youtu.be for 
more on this color scale).

From the `viridis` helpfile: 

\medskip

> "This color map is designed in such a way that it will analytically be
perfectly perceptually-uniform, both in regular form and also when converted to
black-and-white. It is also designed to be perceived by readers with the most
common form of color blindness."

Also see the `viridis` package introduction vignette: 
https://cran.r-project.org/web/packages/viridis/vignettes/intro-to-viridis.html

## `viridis` package

```{r echo = FALSE, message = FALSE, fig.width = 8, fig.height = 2.2, fig.align = "center"}
library(viridis)
# Adapted from viridis helpfile
n <- 200
image(
  1:n, 1, as.matrix(1:n),
  col = viridis(n, option = "D"),
  xlab = "", ylab = "", xaxt = "n", yaxt = "n", bty = "n", 
  main = "Viridis"
) 
```
```{r echo = FALSE, message = FALSE, fig.width = 8, fig.height = 2.2, fig.align = "center"}
image(
  1:n, 1, as.matrix(1:n),
  col = viridis(n, option = "A"),
  xlab = "", ylab = "", xaxt = "n", yaxt = "n", bty = "n",
  main = "Magma"
)
```

```{r echo = FALSE, message = FALSE, fig.width = 8, fig.height = 2.2, fig.align = "center"}
image(
  1:n, 1, as.matrix(1:n),
  col = viridis(n, option = "B"),
  xlab = "", ylab = "", xaxt = "n", yaxt = "n", bty = "n",
  main = "Inferno"
)
```

```{r echo = FALSE, message = FALSE, fig.width = 8, fig.height = 2.2, fig.align = "center"}
image(
  1:n, 1, as.matrix(1:n),
  col = viridis(n, option = "C"),
  xlab = "", ylab = "", xaxt = "n", yaxt = "n", bty = "n",
  main = "Plasma"
)
```


## `viridis` package

For example, here is a hexagonal heatmap showing the two-dimensional distribution 
of state land area versus number of reported storm events:

```{r out.height = "0.5\\textheight", fig.width = 6, fig.height = 4, fig.align = "center"}
state_storms %>% 
  ggplot(aes(x = area, y = n)) + 
  geom_hex(bins = 10)
```

## `viridis` package

Change to a viridis color scale with `scale_fill_viridis`:

```{r out.height = "0.5\\textheight", message = FALSE, fig.width = 6, fig.height = 4, fig.align = "center"}
library(viridis)
state_storms %>% 
  ggplot(aes(x = area, y = n)) + 
  geom_hex(bins = 10) + 
  scale_fill_viridis()
```

## `ggplot2` extensions: Useful additions

- Highlighting interesting points
- "Repelling" text labels
- Arranging plots

## `ggplot2` extensions: repelling text labels

When you add labels to points on a plot, they often overlap:

```{r fig.width = 10, fig.height = 6}
storm_plot + facet_wrap(~ region) + 
  geom_label(aes(label = state))
```

## `ggplot2` extensions: repelling text labels

```{r fig.width = 10, fig.height = 6}
library(ggrepel)
storm_plot + facet_wrap(~ region) + 
  geom_label_repel(aes(label = state))
```

## `gghighlight` package

It may be too much to label every point. Instead, you may just want to 
highlight notable point. 

You can use the `gghighlight` package to do that.

```{r warning = FALSE, message = FALSE, fig.width = 8, fig.height = 3}
library(gghighlight)
storm_plot + facet_wrap(~ region) + 
  gghighlight(area > 150000 | n > 1500, label_key = state)
```

<!-- ## `gghighlight` package -->

<!-- The `gghighlight` package also works for things like histograms. For example, you could -->
<!-- create a dataset with the count by day-of-year of certain types of events: -->

<!-- \footnotesize -->

<!-- ```{r warning = FALSE, message = FALSE} -->
<!-- storms_by_month <- storms_2017 %>%  -->
<!--   filter(event_type %in% c("Flood", "Flash Flood", "Heavy Rain")) %>%  -->
<!--   mutate(month = month(begin_date_time, label = TRUE)) %>%  -->
<!--   group_by(month, event_type) %>%  -->
<!--   count() %>%  -->
<!--   ungroup() -->
<!-- storms_by_month %>%  -->
<!--   slice(1:4) -->
<!-- ``` -->

<!-- ## `gghighlight` package -->

<!-- \small -->

<!-- ```{r fig.width = 7, fig.height = 3} -->
<!-- ggplot(storms_by_month, aes(x = month, y = n, group = event_type)) +   -->
<!--   geom_bar(stat = "identity") +  -->
<!--   labs(x = "Month", y = "# of events") +  -->
<!--   gghighlight(max(n) > 400, label_key = event_type) + -->
<!--   facet_wrap(~ event_type, ncol = 1) -->
<!-- ``` -->

## `ggplot2` extensions: Arranging plots

You may have multiple related plots you want to have as multiple panels of a
single figure. 

There are a few packages that help with this. One very good one is `patchwork`.

You need to install this from GitHub: 

```{r eval = FALSE}
devtools::install_github("thomasp85/patchwork")
```

Find out more: https://github.com/thomasp85/patchwork#patchwork


## `ggplot2` extensions: Arranging plots

Say we want to plot seasonal patterns in events in the five counties with
the highest number of events in 2017. We can use `dplyr` to figure out these
counties:

```{r}
top_counties <- storms_2017 %>% 
  group_by(fips, state, cz_name) %>% 
  count() %>% 
  ungroup() %>% 
  top_n(5, wt = n) 
```

## `ggplot2` extensions: Arranging plots

Then create a plot with the time patterns:

\footnotesize

```{r}
library(forcats)
top_counties_month <- storms_2017 %>% 
  semi_join(top_counties, by = "fips") %>% 
  mutate(month = month(begin_date_time),
         county = paste(cz_name, " County, ", state, sep = "")) %>% 
  count(county, month) %>% 
  ggplot(aes(x = month, y = n)) + 
  geom_bar(stat = "identity") + 
  facet_wrap(~ fct_reorder(county, n, .fun = sum, .desc = TRUE), nrow = 2) +
  scale_x_continuous(name = "", breaks = c(1, 4, 7, 10),
                     labels = c("Jan", "Apr", "Jul", "Oct")) + 
  scale_y_continuous(name = "Frequency", breaks = c(0, 20))
```

## `ggplot2` extensions: Arranging plots

Here's this plot: 

```{r fig.width = 6, fig.height = 3}
top_counties_month
```

## `ggplot2` extensions: Arranging plots

Now that you have two ggplot objects (`storm_plot` and `top_counties_month`), you can use
`patchwork` to put them together:

```{r eval = FALSE}
library(patchwork)
storm_plot +
  top_counties_month + 
  plot_layout(ncol = 1, heights = c(2, 1))
```

## `ggplot2` extensions: Arranging plots

```{r echo = FALSE, fig.height = 5, fig.width = 8, message = FALSE, warning = FALSE, fig.align = "center"}
library(patchwork)
storm_plot +
  top_counties_month + 
  plot_layout(ncol = 1, heights = c(2, 1))
```

## `ggplot2` extensions: Arranging plots

A slightly fancier version:

```{r eval = FALSE}
(storm_plot + theme(legend.position = "top") + 
    gghighlight(n > 1500 | area > 200000,
                label_key = state)) +
  top_counties_month + 
  plot_layout(ncol = 1, heights = c(2, 1))
```

## `ggplot2` extensions: Arranging plots

```{r echo = FALSE, fig.height = 5, fig.width = 8, message = FALSE, warning = FALSE, fig.align = "center"}
(storm_plot + theme(legend.position = "top") + 
    gghighlight(n > 1500 | area > 200000,  label_key = state)) +
  top_counties_month + 
  plot_layout(ncol = 1, heights = c(2, 1))
```

## `ggplot2` extensions: Arranging plots

Other packages for arranging `ggplot` objects: 

- `gridExtra`
- `cowplot`

## In-course exercise

We'll take a break now to do the second part of the in-course exercise.

## Heat maps

**Heat maps** are helpful for exploring the measurements for many variables, 
including how similar they are, by showing small boxes of color. 

In `ggplot2`, there is a `geom_tile` that is useful for creating heatmaps.

## Heat maps

For example, we might want to look at how many events of each type were reported
by each type of source. First, we can summarize the data to get this data:

\footnotesize

```{r, message = FALSE, warning = FALSE}
type_by_source <- storms_2017 %>% 
  filter(event_type %in% c("Thunderstorm Wind", "Heavy Rain", 
                           "Funnel Cloud", "Flash Flood",
                           "Flood", "Hail", "Lightning")) %>% 
  filter(source %in% c("911 Call Center", "Broadcast Media",
                       "Department of Highways",
                       "Emergency Manager", "Fire Department/Rescue",
                       "Law Enforcement", "NWS Employee", "Public",
                       "Social Media", "Trained Spotter")) %>% 
  group_by(event_type, source) %>% count() %>% ungroup()
type_by_source %>% slice(1:3)
```

## Heat maps

\footnotesize

```{r fig.width = 8, fig.height = 3.5, message = FALSE, warning = FALSE}
library(viridis); library(forcats)
ggplot(type_by_source, 
       aes(x = fct_reorder(event_type, n, .fun = sum),
           y = fct_reorder(source, n, .fun =  sum))) + 
  geom_tile(aes(fill = n)) + 
  scale_fill_viridis() + 
  labs(x = "", y = "", fill = "# of events") + 
  theme(legend.position = "bottom")
```

## Alluvial plots

The package `ggalluvial` allows you to make **alluvial plots**. 

\footnotesize

```{r fig.width = 9, fig.height = 3.5}
library(ggalluvial)
ggplot(data = type_by_source,
       aes(axis1 = event_type, axis2 = source, y = n)) + 
  scale_x_discrete(limits = c("Type", "Source"), expand = c(.1, .05)) +
  geom_alluvium(aes(fill = event_type %in% c("Flash Flood", "Flood"))) + 
  theme_minimal() + geom_stratum() + 
  geom_text(stat = "stratum", label.strata = TRUE) + 
  theme(legend.position = "top") + labs(fill = "Flood event: ")
```

## `ggridges` package

The `ggridges` package lets you plot the distribution of several levels of a factor
across another variable. 

## `ggridges` package

Create a dataframe with the timing versus the source of the storm:

\footnotesize

```{r message = FALSE, warning = FALSE}
library(ggridges)
library(forcats)
storms_by_yday <- storms_2017 %>% 
  filter(source %in% c("River/Stream Gages", "Insurance Company",
                       "Airline Pilot", "Trained Spotter",
                       "Storm Chaser", "Local Official",
                       "Official NWS Observations", "Utility Company",
                       "PostOffice", "911 Call Center")) %>%
  mutate(yday = yday(begin_date_time)) %>% 
  group_by(yday, source) %>% 
  count() %>% 
  ungroup() 
```

## `ggridges` package

\footnotesize

```{r message = FALSE, warning = FALSE, fig.height = 3, fig.width = 8}
library(viridis)
ggplot(storms_by_yday, aes(x = yday,
                            y = fct_reorder(source, yday, 
                                            .fun = median, .desc = TRUE),
                           fill = ..x..)) +
  geom_density_ridges_gradient() + 
  scale_fill_viridis(name = "Day in year", option = "B") + 
  xlim(c(0, 365)) + 
  labs(x = "Day in year", y = "")
```

## Odds and ends

Before next week's class, be sure to do the following: 

- Register for GitHub: https://github.com
- Download and install git: https://git-scm.com 


# Mapping in the tidyverse

## Simple features

`sf` objects: "Simple features"

- R framework that is in active development
- There will likely be changes in the near future
- Plays very well with tidyverse functions, including `dplyr` and `ggplot2` tools

```{r message = FALSE, warning = FALSE}
library(sf)
```


## Simple features

To show simple features, we'll pull in the Colorado county boundaries from the U.S. Census. 

To do this, we'll use the `tigris` package, which accesses the U.S. Census API. It allows you
to pull geographic data for U.S. counties, states, tracts, voting districts, roads, rails, 
and a number of other geographies.

To learn more about the `tigris` package, check out this article: 
https://journal.r-project.org/archive/2016/RJ-2016-043/index.html

## Simple features

With `tigris`, you can read in data for county boundaries using the `counties`
function.

We'll use the option
`class = "sf"` to read these spatial dataframes in as `sf` objects.

```{r message = FALSE, warning = FALSE, results = "hide"}
library(tigris)
co_counties <- counties(state = "CO", cb = TRUE, class = "sf")
```

```{r}
class(co_counties)
```


## Simple features

You can think of an `sf` object as a dataframe, but with one special column called `geometry`.

\footnotesize
```{r}
co_counties %>% 
  slice(1:3)
```

## Simple features

The `geometry` column has a special class (sfc):

```{r}
class(co_counties$geometry)
```
## Simple features

You'll notice there's some extra stuff up at the top, too:

- **Geometry type**: Points, polygons, lines
- **Dimension**: Often two-dimensional, but can go up to four (if you have, for example, time
for each measurement and some measure of measurement error / uncertainty)
- **Bounding box (bbox)**: The x- and y-range of the data included
- **EPSG**: The EPSG Geodetic Parameter Dataset code for the Coordinate Reference Systems
- **Projection (proj4string)**: How the data is currently projected, includes projection ("+proj") and datum ("+datum")

## Simple features

You can pull some of this information out of the `geometry` column. For example, you can pull
out the coordinates of the bounding box: 

```{r}
st_bbox(co_counties$geometry)      # For all counties
st_bbox(co_counties$geometry[1]) # Just for first county
```

## Simple features

You can add `sf` objects to `ggplot` objects using `geom_sf`:

\footnotesize

```{r fig.width = 6, fig.height = 4, out.width = "0.7\\textwidth", fig.align = "center"}
library(ggplot2)
ggplot() + 
  geom_sf(data = co_counties)
```


## Simple features

You can map one of the columns in the `sf` object to the fill aesthetic to make a **choropleth**:

\footnotesize

```{r fig.width = 6, fig.height = 4, out.width = "0.8\\textwidth", fig.align = "center"}
ggplot() + 
  geom_sf(data = co_counties, aes(fill = ALAND))
```

## Simple features

You can use all your usual `ggplot` tricks with this:

\footnotesize

```{r fig.width = 6, fig.height = 4, out.width = "0.7\\textwidth", fig.align = "center"}
ggplot() + 
  geom_sf(data = co_counties, aes(fill = ALAND)) + 
  scale_fill_viridis(name = "Land area", label = comma) + 
  ggtitle("Land areas of Colorado counties") + 
  theme_dark()
```

## Simple features

Because simple features are a special type of dataframe, you can also use a lot of `dplyr` tricks. 

For example, you could pull out just Larimer County, CO: 

\footnotesize
```{r}
larimer <- co_counties %>% 
  filter(NAME == "Larimer")
larimer
```

## Simple features


```{r warning = FALSE, message = FALSE, fig.width = 6, fig.height = 4, out.width = "0.6\\textwidth", fig.align = "center"}
ggplot() + 
  geom_sf(data = co_counties, color = "lightgray") + 
  geom_sf(data = larimer, fill = "darkcyan") + 
  geom_sf_text(data = larimer, aes(label = NAME), 
               color = "white") + 
  theme_dark() + labs(x = "", y = "")
```

## Simple features

This operability with tidyverse functions means that you should now be able to figure out how to 
create a map of the number of events listed in the NOAA Storm Events database (of those listed by
county) for each county in Colorado: 

```{r  echo = FALSE, warning = FALSE, message = FALSE, fig.width = 6, fig.height = 4, out.width = "0.7\\textwidth", fig.align = "center"}
co_event_counts <- storms_2017 %>% 
  filter(state == "Colorado") %>% 
  group_by(fips) %>% 
  count() %>% 
  ungroup()

co_county_events <- co_counties %>% 
  mutate(fips = paste(STATEFP, COUNTYFP, sep = "")) %>% 
  full_join(co_event_counts, by = "fips") %>% 
  mutate(n = ifelse(!is.na(n), n, 0))

ggplot() + 
  geom_sf(data = co_county_events, aes(fill = n)) + 
  scale_fill_viridis(name = "Number of events\n(2017)")
```


## In-course exercise

We'll now take a break for the next part of the in-course exercise. For this, create the map shown on 
the previous slide.

## In-course exercise

If you have time, try this one, too. It shows the number of three certain types of events
by county. If a county had no events, it's shown in gray (as having a missing value when
you count up the events that did happen).

```{r echo = FALSE, fig.width = 8, fig.height = 3, out.width = "\\textwidth"}
co_event_counts <- storms_2017 %>% 
  filter(state == "Colorado") %>% 
  filter(event_type %in% c("Tornado", "Heavy Rain", "Hail")) %>% 
  group_by(fips, event_type) %>% 
  count() %>% 
  ungroup()

co_county_events <- co_counties %>% 
  mutate(fips = paste(STATEFP, COUNTYFP, sep = "")) %>% 
  right_join(co_event_counts, by = "fips")

ggplot() +
  geom_sf(data = co_counties, color = "lightgray") + 
  geom_sf(data = co_county_events, aes(fill = n)) + 
  scale_fill_viridis(name = "Number of events\n(2017)") + 
  theme(legend.position = "top") + 
  facet_wrap(~ event_type, ncol = 3) + 
  theme(axis.line = element_blank(),
        axis.text = element_blank(),
        axis.ticks = element_blank())
```


## Simple features

\small

```{r  eval = FALSE}
co_event_counts <- storms_2017 %>% 
  filter(state == "Colorado") %>% 
  group_by(fips) %>% 
  count() %>% 
  ungroup()

co_county_events <- co_counties %>% 
  mutate(fips = paste(STATEFP, COUNTYFP, sep = "")) %>% 
  full_join(co_event_counts, by = "fips") %>% 
  mutate(n = ifelse(!is.na(n), n, 0))

ggplot() + 
  geom_sf(data = co_county_events, aes(fill = n)) + 
  scale_fill_viridis(name = "Number of events\n(2017)")
```

## Simple features

\footnotesize

```{r eval = FALSE}
co_event_counts <- storms_2017 %>% 
  filter(state == "Colorado") %>% 
  filter(event_type %in% c("Tornado", "Heavy Rain", "Hail")) %>% 
  group_by(fips, event_type) %>% 
  count() %>% 
  ungroup()

co_county_events <- co_counties %>% 
  mutate(fips = paste(STATEFP, COUNTYFP, sep = "")) %>% 
  right_join(co_event_counts, by = "fips")

ggplot() +
  geom_sf(data = co_counties, color = "lightgray") + 
  geom_sf(data = co_county_events, aes(fill = n)) + 
  scale_fill_viridis(name = "Number of events\n(2017)") + 
  theme(legend.position = "top") + 
  facet_wrap(~ event_type, ncol = 3) + 
  theme(axis.line = element_blank(),
        axis.text = element_blank(),
        axis.ticks = element_blank())
```

## State boundaries

The `tigris` package allows you to pull state boundaries, as well, but on some computers
mapping these seems to take a really long time. 

Instead, for now I recommend that you pull the state boundaries using base R's `maps` 
package and convert that to an `sf` object: 

```{r}
library(maps)
us_states <- map("state", plot = FALSE, fill = TRUE) %>% 
  st_as_sf()
```

## State boundaries

You can see these borders include an `ID` column that you can use to join by state: 

\footnotesize
```{r}
head(us_states)
```

## State boundaries

As with other `sf` objects, you can map these state boundaries using `ggplot`: 
\footnotesize
```{r fig.width = 5, out.width = "0.6\\textwidth", fig.align = "center"}
ggplot() + 
  geom_sf(data = us_states, color = "white",
          fill = "darkcyan", alpha = 0.5)
```

## State boundaries

As a note, you can use `xlim` and `ylim` with these plots, but remember that the x-axis is longitude in degrees West, which are negative: 

\footnotesize
```{r fig.height = 3, fig.align = "center", out.width = "0.4\\textwidth"}
ggplot() + 
  geom_sf(data = us_states, color = "white", 
          fill = "darkcyan", alpha = 0.5) + 
  xlim(c(-90, -75)) + ylim(c(24, 38))
```



## Basics of creating an `sf` object

You can create an `sf` object from a regular dataframe. 

You just need to specify: 

1. The coordinate information (which columns are longitudes and latitudes)
2. The Coordinate Reference System (CRS) (how to translate your coordinates to places in the world)

For the CRS, if you are mapping the new `sf` object with other, existing `sf` objects, 
make sure that you use the same CRS for all `sf` objects.

## Coordinate reference system

Spatial objects can have different Coordinate Reference Systems (CRSs). CRSs can be *geographic* (e.g., WGS84, for longitude-latitude data) or *projected* (e.g., UTM, NADS83). 

There is a website that lists projection strings and can be useful in setting projection information or re-projecting data: http://www.spatialreference.org

Here is an excellent resource on projections and maps in R from Melanie Frazier: https://www.nceas.ucsb.edu/~frazier/RSpatialGuides/OverviewCoordinateReferenceSystems.pdf

## Basics of creating an `sf` object

Let's look at floods in Colorado. First, clean up the data:

```{r}
co_floods <- storms_2017 %>% 
  filter(state == "Colorado" & 
           event_type %in% c("Flood", "Flash Flood")) %>% 
  select(begin_date_time, event_id, begin_lat:end_lon) %>% 
  gather(key = "key", value = "value", 
         -begin_date_time, -event_id) %>% 
  separate(key, c("time", "key")) %>% 
  spread(key = key, value = value)
```

## Basics of creating an `sf` object

There are now two rows per event, one with the starting location and one with the ending location:

```{r}
co_floods %>% 
  slice(1:5)
```

## Basics of creating an `sf` object

Change to an `sf` object by saying which columns are the coordinates and setting a CRS:

\small

```{r}
co_floods <- st_as_sf(co_floods, coords = c("lon", "lat")) %>% 
  st_set_crs(4269)
co_floods %>% slice(1:3)
```



## Basics of creating an `sf` object

```{r out.width = "0.8\\textwidth", fig.width = 6}
ggplot() + 
  geom_sf(data = co_counties, color = "lightgray") + 
  geom_sf(data = co_floods, aes(color = month(begin_date_time),
                                  shape = time)) + 
  scale_color_viridis(name = "Month") 
```

## Changing from points to lines

If you want to show lines instead of points, group by the appropriate ID and then summarize
within each event to get a line:

```{r}
co_floods <- co_floods %>% 
  group_by(event_id) %>% 
  summarize(month = month(first(begin_date_time)),
            do_union = FALSE) %>% 
  st_cast("LINESTRING")
```

## Changing from points to lines

```{r}
head(co_floods)
```


## Changing from points to lines

\footnotesize
```{r out.width = "0.8\\textwidth", fig.width = 6, fig.align = "center"}
ggplot() + 
  geom_sf(data = co_counties, color = "lightgray") + 
  geom_sf(data = co_floods, 
          aes(color = factor(month), fill = factor(month))) + 
  scale_fill_viridis(name = "Month", discrete = TRUE) + 
  scale_color_viridis(name = "Month", discrete = TRUE) 
```


## Reading from GIS files

You can also create `sf` objects by reading in data from files you would normally use for GIS. 

For example, you can read in an `sf` object from a shapefile, which is a format often used for GIS in which a collection of several files jointly store geographic data. The files making up a shapefile can include:

- ".shp": The coordinates defining the shape of each geographic object. For a point, this
would be a single coordinate (e.g., latitude and longitude). For lines and polygons, there
will be multiple coordinates per geographic object.
- ".prf": Information on the projection of the data (how to get from the coordinates to 
a place in the world).
- ".dbf": Data that goes along with each geographical object. For example, earlier we looked
at data on counties, and one thing measured for each county was its land area. Characteristics
like that would be included in the ".dbf" file in a shapefile.


## Reading from GIS files

Often, with geographic data, you will be given the option to downloaded a compressed 
file (e.g., a zipped file). When you unzip the folder, it will include a number of 
files in these types of formats (".shp", "prf", ".dbf", etc.).

Sometimes, that single folder will include multiple files from each extension. For
example, it might have several files that end with ".shp". In this case, you have
multiple **layers** of geographic information you can read in.

## Reading in from GIS files

We've been looking at data on storms from NOAA for 2017. As an example, let's try to 
pair that data up with some from the National Hurricane Center for the same year. 

The National Hurricane Center allows you to access a variety of GIS data through 
the webpage https://www.nhc.noaa.gov/gis/?text. 

Let's pull some data on Hurricane Harvey in 2017 and map it with information from 
the NOAA Storm Events database. 

## Reading in from GIS files

On https://www.nhc.noaa.gov/gis/?text, go to the section called "Preliminary Best Track".
Select the year 2017. Then select "Hurricane Harvey" and download "al092017_best_track.zip".

Depending on your computer, you may then need to unzip this file (many computers will
unzip it automatically). Base R has a function called `unzip` that can help with this.

You'll then have a folder with a number of different files in it. Move this folder 
somewhere that is convenient for the working directory you use for class. For example,
I moved it into the "data" subdirectory of the working directory I use for the class.

## Reading in from GIS files

You can use `list.files` to see all the files in this unzipped folder:

\footnotesize
```{r}
list.files("../data/al092017_best_track/")
```

## Reading from GIS files

You can use `st_layers` to find out the available layers in a shapefile directory:

```{r}
st_layers("../data/al092017_best_track/")
```

## Reading from GIS files

Once you know which layer you want, you can use `read_sf` to read it in as an `sf` object: 

```{r}
harvey_track <- read_sf("../data/al092017_best_track/",
                        layer = "al092017_lin")
head(harvey_track)
```

## Reading from GIS files

```{r fig.width = 6, out.width = "0.8\\textwidth", fig.align = "center"}
ggplot() + 
  geom_sf(data = filter(us_states, ID %in% c("texas", "louisiana"))) + 
  geom_sf(data = harvey_track, aes(color = STORMTYPE)) + 
  xlim(c(-107, -89)) + ylim(c(25, 37)) 
```

## Reading from GIS files

You can read in other layers:

\footnotesize
```{r}
harvey_windswath <- read_sf("../data/al092017_best_track/",
                            layer = "al092017_windswath")
head(harvey_windswath)
```

## Reading from GIS files

\footnotesize
```{r fig.width = 6, out.width = "0.8\\textwidth", fig.align = "center"}
ggplot() + 
  geom_sf(data = filter(us_states, ID %in% c("texas", "louisiana"))) + 
  geom_sf(data = harvey_windswath, 
          aes(fill = factor(RADII)), alpha = 0.2) + 
  xlim(c(-107, -89)) + ylim(c(25, 37)) + 
  scale_fill_viridis(name = "Wind (kts)", discrete = TRUE, 
                     option = "B", begin = 0.6, direction = -1)
```

## Reading from GIS files

The `read_sf` function is very powerful and can read in data from lots of different 
formats. 

See Section 2 of the `sf` manual (https://cran.r-project.org/web/packages/sf/vignettes/sf2.html)
for more on this function.

## More on spatial data with R

You can find (much, much) more on working with spatial data in R online: 

- **R Spatial**: http://rspatial.org/index.html
- **Geocomputation with R**: https://geocompr.robinlovelace.net

## In-course exercise

For the in-course exercise, see if you can put everything we've talked about together
to create this map: 

```{r echo = FALSE, message = FALSE, fig.width = 7, out.width = "0.8\\textwidth", fig.align = "center"}
harvey_events <- storms_2017 %>% 
  filter(state %in% c("Texas", "Louisiana") & 
           ymd_hms("2017-08-25 00:00:00") < begin_date_time & 
           begin_date_time < ymd_hms("2017-10-05 00:00:00")) %>% 
  group_by(fips) %>% 
  count() %>% 
  ungroup()

tx_la_counties <- counties(state = c("TX", "LA"), cb = TRUE, class = "sf") %>% 
  mutate(fips = paste(STATEFP, COUNTYFP, sep = "")) %>% 
  full_join(harvey_events, by = "fips") %>% 
  mutate(n = ifelse(is.na(n), 0, n))

ggplot() + 
  geom_sf(data = tx_la_counties, color = NA, 
          aes(fill = n)) +
  geom_sf(data = filter(us_states, ID %in% c("texas", "louisiana")), 
          fill = NA, color = "lightgray") + 
  geom_sf(data = harvey_windswath, 
          aes(color = factor(RADII)), alpha = 0.4) + 
  geom_sf(data = harvey_track, color = "red", 
          alpha = 0.1, size = 1) + 
  xlim(c(-107, -89)) + ylim(c(25, 37)) + 
  scale_fill_viridis(name = "# of events", option = "A", direction = -1) + 
  scale_color_viridis(name = "Wind (kts)", discrete = TRUE, 
                      option = "B", direction = -1) + 
  theme_dark() + 
  ggtitle("Number of events near Harvey's landfall",
          subtitle = "# events starting Aug. 25 to Sept. 5, 2017")
```



