---
title: "Mapping in R"
author: "Brooke Anderson"
date: "November 16, 2016"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r echo = FALSE}
library(dplyr)
library(lubridate)
library(sp)
library(GISTools)
```


## Example data

```{r}
library(tigris)
denver_tracts <- tracts(state = "CO", county = 31, cb = TRUE)
load("../data/fars_colorado.RData")
accident_data <- driver_data %>%
  dplyr::select(state, st_case, county, latitude, longitud,
                date, fatals, drunk_dr) %>%
  dplyr::filter(county == 31 & longitud < -104.1) %>%
  dplyr::distinct()
```

## Leaflet

"Leaflet" is a JavaScript library for making interactive maps. You can find out more about the JavaScript version here: http://leafletjs.com

The `leaflet` package brings this functionality to R. The R Studio group has created a website on `leaflet`: http://rstudio.github.io/leaflet/. This website walks you through different options available with `leaflet`.

```{r}
library(leaflet)
```

## Add map background

If you just run `leaflet()`, you just get a blank leaflet area: 

```{r fig.width = 7, fig.height = 3}
leaflet()
```

## Add map background

In `leaflet`, the map background is composed of *tiles*. To get something more interesting, you'll need to add tiles to your leaflet map. If you don't include any other data, the leaflet map will include the world:

```{r eval = FALSE}
leaflet() %>%
  addTiles()
```

## Add map background

```{r echo = FALSE}
leaflet() %>%
  addTiles()
```

## Adding markers

For htmlWidgets, points are often referred to as *markers*.

Once you add these markers, the map will automatically scale to a reasonable size for their bounding box.

```{r eval = FALSE}
leaflet() %>%
  addTiles() %>%
  addMarkers(data = accident_data, lng = ~ longitud, lat = ~ latitude)
```

Use `lng` and `lat` to tell R which columns contain data on longitude and latitude for each point. This is not needed if you are using a spatial object (e.g., `SpatialPointsDataFrame`). Further, R will try to guess the columns in a regular dataframe.

## Adding markers

```{r echo = FALSE}
leaflet() %>%
  addTiles() %>%
  addMarkers(data = accident_data, lng = ~ longitud, lat = ~ latitude)
```

## Adding markers

You can use several types of R objects for your data for `leaflet`:

- Dataframe with columns for latitude and longitude
- Spatial objects (`SpatialPoints`, `SpatialLines`, etc.)
- Latitude-longitude matrix

## Adding markers

You can choose circles for your markers instead by using `addCircleMarkers`. You can adjust the circle size with `radius`. 

```{r eval = FALSE}
leaflet() %>%
  addTiles() %>%
  addCircleMarkers(data = accident_data, radius = 2,
                   lng = ~ longitud, lat = ~ latitude)
```

The `radius` argument specifies the size of the circle. For `CircleMarkers`, the size will reset as you zoom in and out. If you want something with a constant radius (e.g., in meters), you can add `Circles`.

## Adding markers

```{r echo = FALSE}
leaflet() %>%
  addTiles() %>%
  addCircleMarkers(data = accident_data, radius = 2,
                   lng = ~ longitud, lat = ~ latitude)
```

## Adding markers

If you have a lot of overlapping data, you can also use the `clusterOptions` argument to show the markers as clusters that group together when you zoom out but split up when you zoom in:

```{r eval = FALSE}
leaflet() %>%
  addTiles() %>%
  addMarkers(data = accident_data, 
                   lng = ~ longitud, lat = ~ latitude,
                   clusterOptions = markerClusterOptions())
```

## Adding markers

```{r echo = FALSE}
leaflet() %>%
  addTiles() %>%
  addMarkers(data = accident_data, 
                   lng = ~ longitud, lat = ~ latitude,
                   clusterOptions = markerClusterOptions())
```

## Add map background

For the background, the default is to use map tiles from OpenStreetMap. However, you can change the source of the tiles by using `addProviderTiles`. For example, to use Stamen Watercolor, you can call:

```{r eval = FALSE}
leaflet() %>%
  addProviderTiles("Stamen.Watercolor") %>%
  addCircleMarkers(data = accident_data, radius = 2,
                   lng = ~ longitud, lat = ~ latitude)
```

## Add map background

```{r echo = FALSE}
leaflet() %>%
  addProviderTiles("Stamen.Watercolor") %>%
  addCircleMarkers(data = accident_data, radius = 2,
                   lng = ~ longitud, lat = ~ latitude)
```

## Add map background

```{r fig.width = 7, fig.height = 3}
leaflet() %>%
  addProviderTiles("Thunderforest.TransportDark") %>%
  addCircleMarkers(data = accident_data, radius = 2,
                   lng = ~ longitud, lat = ~ latitude)
```

## Add map background

```{r fig.width = 7, fig.height = 3}
leaflet() %>%
  addProviderTiles("Esri.WorldStreetMap") %>%
  addCircleMarkers(data = accident_data, radius = 2,
                   lng = ~ longitud, lat = ~ latitude)
```

## Add map background

You can see previews of provider choices here: http://leaflet-extras.github.io/leaflet-providers/preview/index.html.

## Pop-ups

You can use the `popup` option to show information when the user clicks on a marker. 

It's easiest to do this if you have the information you want to show in the dataframe with the location data. For example, we have date-time, number of fatalities, and number of drunk drivers in this data: 

```{r}
accident_data %>% 
  dplyr::select(date, fatals, drunk_dr) %>% 
  dplyr::slice(1:3)
```

## Pop-ups

If we want to show day of the week, month, hour, and number of fatalities, go ahead and calculate any value not already in the dataset:

```{r}
library(lubridate)
accident_data <- accident_data %>%
  mutate(weekday = wday(date, label = TRUE, abbr = FALSE),
         month = month(date, label = TRUE, abbr = FALSE),
         hour = format(date, format = "%H:%M"))
```

## Pop-ups

The popup text needs to be a character vector, written in HTML syntax. You can create that vector first, and then pass it to the `popup` argument.

```{r}
popup_info <- paste0("<b>Weekday:</b>  ", 
                                  accident_data$weekday, "<br/>",
                                  "<b>Month:</b>  ",
                                  accident_data$month, "<br/>",
                                  "<b>Hour:</b>  ",
                                  accident_data$hour, "<br/>",
                                  "<b>Fatalities:</b>  ",
                                  accident_data$fatals)
popup_info[1:3]
```

## Pop-ups

Now pass that vector to the `popup` argument for the layer you want to pair it with:

```{r eval = FALSE}
leaflet() %>%
  addTiles() %>%
  addCircleMarkers(data = accident_data, radius = 2,
                   lng = ~ longitud, lat = ~ latitude,
                   popup = popup_info)
```

## Pop-ups

```{r echo = FALSE}
leaflet() %>%
  addTiles() %>%
  addCircleMarkers(data = accident_data, radius = 2,
                   lng = ~ longitud, lat = ~ latitude,
                   popup = popup_info)
```

## Pop-ups

If you aren't familiar with HTML syntax, here's one cheatsheet: http://web.stanford.edu/group/csp/cs21/htmlcheatsheet.pdf

In the popups, you can use HTML to format things like color, typeface, and size. You can also add links.

## Mapping values to color

To use color to show a value, you need to do a few things. First, you need to the the `colorFactor` function (or another in its family) to create a function for mapping from values to colors. Then, you need to use this within the call to add the markers.

```{r eval = FALSE}
library(viridisLite)
pal <- colorFactor(viridis(5), accident_data$drunk_dr)
leaflet() %>%
  addProviderTiles("OpenStreetMap.BlackAndWhite") %>%
  addCircleMarkers(data = accident_data, radius = 2,
                   lng = ~ longitud, lat = ~ latitude,
                   popup = popup_info, 
                   color = pal(accident_data$drunk_dr)) 
```

## Mapping values to color

```{r echo = FALSE}
library(viridisLite)
pal <- colorFactor(viridis(5), accident_data$drunk_dr)
leaflet() %>%
  addProviderTiles("OpenStreetMap.BlackAndWhite") %>%
  addCircleMarkers(data = accident_data, radius = 2,
                   lng = ~ longitud, lat = ~ latitude,
                   popup = popup_info, 
                   color = pal(accident_data$drunk_dr)) 
```

## Mapping values to color

The `colorFactor` function (and friends) are a pretty cool type of function that actually creates a new function: 

```{r}
pal <- colorFactor(viridis(5), accident_data$drunk_dr)
class(pal)
head(pal)
```

## Adding a legend

Once you are showing something with color, you can add a legend to explain it. You can do that with the `addLegend` function, which must include values for the color palette and values for each point from this color palette.

```{r eval = FALSE}
library(viridisLite)
pal <- colorFactor(viridis(5), accident_data$drunk_dr)
leaflet() %>%
  addProviderTiles("OpenStreetMap.BlackAndWhite") %>%
  addCircleMarkers(data = accident_data, radius = 2,
                   lng = ~ longitud, lat = ~ latitude,
                   popup = popup_info, 
                   color = pal(accident_data$drunk_dr)) %>%
  addLegend(pal = pal, values = accident_data$drunk_dr)
```

## Adding a legend

```{r echo = FALSE}
library(viridisLite)
pal <- colorFactor(viridis(5), accident_data$drunk_dr)
leaflet() %>%
  addProviderTiles("OpenStreetMap.BlackAndWhite") %>%
  addCircleMarkers(data = accident_data, radius = 2,
                   lng = ~ longitud, lat = ~ latitude,
                   popup = popup_info, 
                   color = pal(accident_data$drunk_dr)) %>%
  addLegend(pal = pal, values = accident_data$drunk_dr)
```

## Adding polygons

You can add polygons with the `addPolygons` function.

```{r eval = FALSE}
leaflet() %>%
  addProviderTiles("OpenStreetMap.BlackAndWhite") %>%
  addPolygons(data = denver_tracts)
```

## Adding polygons

```{r echo = FALSE}
leaflet() %>%
  addProviderTiles("OpenStreetMap.BlackAndWhite") %>%
  addPolygons(data = denver_tracts)
```

## Adding polygons

You can add popups for polygons, as well.

```{r eval = FALSE}
polygon_popup <- paste0("Tract ID:  ", 
                        denver_tracts@data$NAME)
leaflet() %>%
  addProviderTiles("OpenStreetMap.BlackAndWhite") %>%
  addPolygons(data = denver_tracts, popup = polygon_popup)
```

## Adding polygons

```{r echo = FALSE}
polygon_popup <- paste0("Tract ID:  ", 
                        denver_tracts@data$NAME)
leaflet() %>%
  addProviderTiles("OpenStreetMap.BlackAndWhite") %>%
  addPolygons(data = denver_tracts, popup = polygon_popup)
```

## Adding polygons

You can overlay different elements. For example, you can show both accidents and tracts:

```{r eval = FALSE}
polygon_popup <- paste0("Tract ID:  ", 
                        denver_tracts@data$NAME)
accident_data_sp <- accident_data
coordinates(accident_data_sp) <- c("longitud", "latitude")
proj4string(accident_data_sp) <- CRS(proj4string(denver_tracts))
leaflet() %>%
  addProviderTiles("Thunderforest.Transport") %>%
  addPolygons(data = denver_tracts, popup = polygon_popup,
              color = "#000000", fillColor = "969696", 
              weight = 2) %>%
  addCircleMarkers(data = accident_data_sp, radius = 2,
                   popup = popup_info, opacity = 0.9,
                   color = pal(accident_data$drunk_dr)) %>%
  addLegend(pal = pal, values = accident_data$drunk_dr, opacity = 0.9)
```

## Adding polygons

```{r echo = FALSE}
polygon_popup <- paste0("Tract ID:  ", 
                        denver_tracts@data$NAME)
accident_data_sp <- accident_data
coordinates(accident_data_sp) <- c("longitud", "latitude")
proj4string(accident_data_sp) <- CRS(proj4string(denver_tracts))
leaflet() %>%
  addProviderTiles("Thunderforest.Transport") %>%
  addPolygons(data = denver_tracts, popup = polygon_popup,
              color = "#000000", fillColor = "969696", 
              weight = 2) %>%
  addCircleMarkers(data = accident_data_sp, radius = 2,
                   popup = popup_info, opacity = 0.9,
                   color = pal(accident_data$drunk_dr)) %>%
  addLegend(pal = pal, values = accident_data$drunk_dr, opacity = 0.9)
```

## Adding polygons

You can add the ability for the user to pick which layers to see using `addLayersControls`. 

```{r eval = FALSE}
leaflet() %>%
  addProviderTiles("Thunderforest.Transport", group = "base map") %>%
  addPolygons(data = denver_tracts, popup = polygon_popup,
              color = "#000000", fillColor = "969696", 
              weight = 2, group = "tracts") %>%
  addCircleMarkers(data = accident_data_sp, radius = 2,
                   popup = popup_info, opacity = 0.9,
                   color = pal(accident_data$drunk_dr), 
                   group = "accidents") %>%
  addLegend(pal = pal, values = accident_data$drunk_dr, opacity = 0.9) %>%
  addLayersControl(baseGroups = c("base map"), 
                   overlayGroups = c("tracts", "accidents"))
```

## Adding polygons

```{r echo = FALSE}
leaflet() %>%
  addProviderTiles("Thunderforest.Transport", group = "base map") %>%
  addPolygons(data = denver_tracts, popup = polygon_popup,
              color = "#000000", fillColor = "969696", 
              weight = 2, group = "tracts") %>%
  addCircleMarkers(data = accident_data_sp, radius = 2,
                   popup = popup_info, opacity = 0.9,
                   color = pal(accident_data$drunk_dr), 
                   group = "accidents") %>%
  addLegend(pal = pal, values = accident_data$drunk_dr, opacity = 0.9) %>%
  addLayersControl(baseGroups = c("base map"), 
                   overlayGroups = c("tracts", "accidents"))
```

## Find out more

Here are some good tutorials for trying out other examples of `leaflet` in R:

http://zevross.com/blog/2015/10/14/manipulating-and-mapping-us-census-data-in-r-using-the-acs-tigris-and-leaflet-packages-3/

http://robinlovelace.net/r/2015/02/01/leaflet-r-package.html

http://trendct.org/2015/06/26/tutorial-how-to-put-dots-on-a-leaflet-map-with-r/