---
title: "Getting / cleaning data 2"
header-includes:
   - \usepackage{hyperref}
   - \definecolor{links}{HTML}{2A1B81}
   - \hypersetup{colorlinks,linkcolor=,urlcolor=links}
output: 
  beamer_presentation:
    theme: "metropolis"
fontsize: 10pt
---

```{r echo = FALSE, message = FALSE, warning = FALSE}
library(tidyverse)
library(knitr)
library(faraway)
data(worldcup)
```

# More with `dplyr`

## `dplyr`

So far, you've used several `dplyr` functions: 

- `rename`
- `filter`
- `select`
- `mutate`
- `group_by`
- `summarize`

Some other useful `dplyr` functions to add to your toolbox are: 

- `separate` and `unite`
- `mutate` and other `dplyr` functions with `group_by`
- `anti_join` and `semi_join`

## `separate`

Sometimes, you want to take one column and split it into two columns. For example, you may have information for two variables in one column: 

```{r echo = FALSE}
ebola <- tibble(ebola_key = c("Liberia_Cases", "Liberia_Deaths", 
                              "Spain_Cases", "Spain_Deaths"))
```

```{r}
ebola
```


## `separate`

If you have a consistent "split" character, you can use the `separate` function to split one column into two: 

```{r}
ebola %>%
  separate(col = ebola_key, into = c("country", "outcome"),
           sep = "_")
```

## `separate`

Here is the generic code for `separate`: 

```{r eval = FALSE}
## Generic code
separate([dataframe], 
         col = [name of the single column you want to split],
         into = [vector of the names of the columns 
                 you want to create],
         sep = [the character that designates where 
                you want to split])
```

## `separate`

The default is to drop the original column and only keep the columns into which it was split. However, you can use the argument `remove = FALSE` to keep the first column, as well: 

```{r}
ebola %>%
  separate(col = ebola_key, into = c("country", "outcome"),
           sep = "_", remove = FALSE)
```

## `separate`

You can use the `fill` argument (`fill = "right"` or `fill = "left"`) to control what happens when there are some observations that do not have the split character. \bigskip

For example, say your original column looked like this:

```{r echo = FALSE}
ebola <- tibble(ebola_key = c("Liberia_Cases", "Liberia", 
                              "Spain_Cases", "Spain_Deaths"))
ebola
```

## `separate`

You can use `fill = "right"` to set how to split observations like the second one, where there is no separator character ("_"):

```{r}
ebola %>%
  separate(col = ebola_key, into = c("country", "outcome"),
           sep = "_", fill = "right")
```

