---
title: "Reporting data results #3"
author: "Brooke Anderson"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Introduction to Leaflet

## Example data

Cleaned data on fatal automobile accident data for 2001--2010 [here](https://github.com/geanders/RProgrammingForResearch/blob/master/data/fars_colorado.RData) (click on the "Download" button on the page).

```{r message = FALSE, results='hide'}
library(tigris)
denver_tracts <- tracts(state = "CO", county = 31, 
                        cb = TRUE, class = "sf")

load("../data/fars_colorado.RData") # Use relative path on your computer 
library(tidyverse)
accident_data <- driver_data %>%
  select(state, st_case, county, latitude, longitud,
                date, fatals, drunk_dr) %>%
  filter(county == 31 & longitud < -104.1) %>%
  distinct()
```

## Leaflet

"Leaflet" is a JavaScript library for making interactive maps. You can find out more about the JavaScript version here: http://leafletjs.com

The `leaflet` package brings this functionality to R. The R Studio group has created a website on `leaflet`: http://rstudio.github.io/leaflet/. This website walks you through different options available with `leaflet`.

```{r}
library(leaflet)
```

## Add map background

If you just run `leaflet()`, you just get a blank leaflet area: 

```{r fig.width = 7, fig.height = 3}
leaflet()
```

## Add map background

In `leaflet`, the map background is composed of *tiles*. To get something more interesting, you'll need to add tiles to your leaflet map. If you don't include any other data, the leaflet map will include the world:

```{r eval = FALSE}
leaflet() %>%
  addTiles()
```

## Add map background

```{r echo = FALSE}
leaflet() %>%
  addTiles()
```

## Adding markers

For htmlWidgets, points are often referred to as *markers*.

Once you add these markers, the map will automatically scale to a reasonable size for their bounding box.

```{r eval = FALSE}
leaflet() %>%
  addTiles() %>%
  addMarkers(data = accident_data, lng = ~ longitud, lat = ~ latitude)
```

Use `lng` and `lat` to tell R which columns contain data on longitude and latitude for each point. This is not needed if you are using a spatial object (e.g., `SpatialPointsDataFrame`). Further, R will try to guess the columns in a regular dataframe.

## Adding markers

```{r echo = FALSE}
leaflet() %>%
  addTiles() %>%
  addMarkers(data = accident_data, lng = ~ longitud, lat = ~ latitude)
```

## Adding markers

You can use several types of R objects for your data for `leaflet`:

- Dataframe with columns for latitude and longitude
- Simple feature objects 
- Latitude-longitude matrix

## Adding markers

You can choose circles for your markers instead by using `addCircleMarkers`. You can adjust the circle size with `radius`. 

```{r eval = FALSE}
leaflet() %>%
  addTiles() %>%
  addCircleMarkers(data = accident_data, radius = 2,
                   lng = ~ longitud, lat = ~ latitude)
```

The `radius` argument specifies the size of the circle. For `CircleMarkers`, the size will reset as you zoom in and out. If you want something with a constant radius (e.g., in meters), you can add `Circles`.

## Adding markers

```{r echo = FALSE}
leaflet() %>%
  addTiles() %>%
  addCircleMarkers(data = accident_data, radius = 2,
                   lng = ~ longitud, lat = ~ latitude)
```

## Adding markers

If you have a lot of overlapping data, you can also use the `clusterOptions` argument to show the markers as clusters that group together when you zoom out but split up when you zoom in:

```{r eval = FALSE}
leaflet() %>%
  addTiles() %>%
  addMarkers(data = accident_data, 
                   lng = ~ longitud, lat = ~ latitude,
                   clusterOptions = markerClusterOptions())
```

## Adding markers

```{r echo = FALSE}
leaflet() %>%
  addTiles() %>%
  addMarkers(data = accident_data, 
                   lng = ~ longitud, lat = ~ latitude,
                   clusterOptions = markerClusterOptions())
```

## Add map background

For the background, the default is to use map tiles from OpenStreetMap. However, you can change the source of the tiles by using `addProviderTiles`. For example, to use Stamen Watercolor, you can call:

```{r eval = FALSE}
leaflet() %>%
  addProviderTiles("Stamen.Watercolor") %>%
  addCircleMarkers(data = accident_data, radius = 2,
                   lng = ~ longitud, lat = ~ latitude)
```

## Add map background

```{r echo = FALSE}
leaflet() %>%
  addProviderTiles("Stamen.Watercolor") %>%
  addCircleMarkers(data = accident_data, radius = 2,
                   lng = ~ longitud, lat = ~ latitude)
```


## Add map background

```{r fig.width = 7, fig.height = 3}
leaflet() %>%
  addProviderTiles("Esri.WorldStreetMap") %>%
  addCircleMarkers(data = accident_data, radius = 2,
                   lng = ~ longitud, lat = ~ latitude)
```

## Add map background

You can see previews of provider choices here: http://leaflet-extras.github.io/leaflet-providers/preview/index.html.

