---
title: "Exploring data #1"
header-includes:
   - \usepackage{hyperref}
   - \definecolor{links}{HTML}{2A1B81}
   - \hypersetup{colorlinks,linkcolor=,urlcolor=links}
output: 
  beamer_presentation:
    theme: "metropolis"
fontsize: 10pt
---

```{r echo = FALSE}
options(width = 50)
```

```{r, message = FALSE, echo = FALSE}
library("readr")
library("dplyr")
library("ggplot2")
```
```{r echo = FALSE, message = FALSE}
beijing_pm_raw <- read_csv("../data/Beijing_2017_HourlyPM25.csv", 
                       skip = 3, na = "-999")
```

```{r, message = FALSE, echo = FALSE}
library(lubridate)
beijing_pm <- beijing_pm_raw %>% 
  rename(sample_time = `Date (LST)`, 
         value = Value,
         qc = `QC Name`) %>% 
  select(sample_time, value, qc) %>% 
  mutate(aqi = cut(value, 
                   breaks = c(0, 50, 100, 150, 200, 
                              300, 500, Inf),
                   labels = c("Good", "Moderate",
                              "Unhealthy for Sensitive Groups",
                              "Unhealthy", "Very Unhealthy", 
                              "Hazardous", "Beyond Index"))) %>% 
  mutate(sample_time = mdy_hm(sample_time)) %>% 
  mutate(heating = sample_time < mdy("03/15/2017"))
```

# Aesthetics

## Plot aesthetics

Which aesthetics you must specify in the `aes` call depend on which geom you are
adding to the plot. \medskip

You can find out the aesthetics you can use for a geom in the "Aesthetics"
section of the geom's help file (e.g., `?geom_point`). \medskip

Required aesthetics are in bold in this section of the help file and optional
ones are not.

## Constant aesthetics

Instead of mapping an aesthetic to an element of your data, you can use a
constant value for the aesthetic. For example, you may want to make all the
points blue, rather than having color map to AQI:

```{r echo = FALSE, warning = FALSE, fig.align = "center", out.width = "0.6\\textwidth", message = FALSE, fig.width = 5, fig.height = 3}
beijing_pm %>%
  ggplot(aes(x = sample_time, y = value)) + 
  geom_point(color = "blue")
```

In this case, you can define that aesthetic as a constant for the geom,
**outside** of an `aes` statement.

## Constant aesthetics

For example, you may want to change the shape of the points in a scatterplot from their default shape, but not map them to a particular element of the data. \medskip

In R, you can specify point shape with a number. Here are the shapes that correspond to the numbers 1 to 25:

```{r echo = FALSE, out.width = "0.6\\textwidth", fig.width = 5, fig.height = 3, fig.align = "center"}
x <- rep(1:5, 5)
y <- rep(1:5, each = 5)
shape <- 1:25
to_plot <- tibble(x = x, y = y, shape = shape)
ggplot(to_plot, aes(x = x, y = y)) + 
  geom_point(shape = shape, size = 4, color = "black", fill = "red") + 
  geom_text(label = shape, nudge_x = -0.25) +
  xlim(c(0.5, 5.5)) + 
  theme_void() + 
  scale_y_reverse()
```

## Constant aesthetics

Here is an example of mapping point shape to a constant value other than the
default:

```{r warning = FALSE, fig.align = "center", out.width = "0.9\\textwidth", message = FALSE, fig.width = 8, fig.height = 3.5}
ggplot(data = beijing_pm) + 
  geom_point(mapping = aes(x = sample_time, y = value,
                           color = aqi),
             shape = 9)
```

## Constant aesthetics

R has character names for different colors. For example: 

```{r echo = FALSE, out.width = "0.18\\textwidth", fig.width = 5, fig.height = 3, fig.align = "center"}
x <- rep(0, 6)
y <- 1:6
color <- c("blue", "blue4", "darkorchid", "deepskyblue2", 
           "steelblue1", "dodgerblue3")
to_plot <- tibble(x = x, y = y, color = color)
ggplot(to_plot, aes(x = x, y = y)) + 
  geom_point(color = color, size = 2) + 
  geom_text(label = color, hjust = 0, nudge_x = 0.05) + 
  theme_void() + 
  xlim(c(-1, 1.5)) +
  scale_y_reverse()
```

Google "R colors" and search the images to find links to listings of different R colors.

## Constant aesthetics

Here is an example of mapping point shape and color to constant values other than the defaults:

```{r warning = FALSE, fig.align = "center", out.width = "0.9\\textwidth", message = FALSE, fig.width = 7, fig.height = 3}
ggplot(data = beijing_pm) + 
  geom_point(mapping = aes(x = sample_time, y = value),
             shape = 9, 
             color = "darkorchid")
```

## Useful plot additions

There are also a number of elements that you can add onto a `ggplot` object using `+`. A few very frequently used ones are: 

```{r echo = FALSE}
plot_adds <- data.frame(add = c("`ggtitle`",
                                "`xlab`, `ylab`, `labs`",
                                "`xlim`, `ylim`",
                                "`expand_limits`"),
                        descrip = c("Plot title",
                                    "x- and y-axis labels",
                                    "Limits of x- and y-axis",
                                    "Include a value in a range"))
knitr::kable(plot_adds, col.names = c("Element", "Description"))
```

## Useful plot additions

```{r warning = FALSE, fig.align = "center", out.width = "0.6\\textwidth", message = FALSE, fig.width = 5, fig.height = 3}
ggplot(data = beijing_pm) + 
  geom_point(mapping = aes(x = sample_time, y = value)) + 
  labs(x = "Sampling Date and Time", 
       y = "PM2.5 Concentration") + 
  ggtitle("Measurements of PM2.5 in Beijing, China, 2017",
          subtitle = "Based on U.S. Embassy Monitor") 
```

