---
title: "Entering / cleaning data 1"
header-includes:
   - \usepackage{hyperref}
   - \definecolor{links}{HTML}{2A1B81}
   - \hypersetup{colorlinks,linkcolor=,urlcolor=links}
output: 
  beamer_presentation:
    theme: "metropolis"
fontsize: 10pt
---

```{r echo = FALSE}
options(width = 50)
```

# Odds and ends

## `dplyr` versus base R

Just so you know, all of these actions also have alternatives in base R:

```{r echo = FALSE}
dplyr_vs_base <- data.frame(dplyr = c("`rename`",
                                      "`select`",
                                      "`slice`",
                                      "`filter`",
                                      "`mutate`"),
                            base = c("Reassign `colnames`",
                                     "Square bracket indexing (`[` , `]`)",
                                     "Square bracket indexing (`[` , `]`)",
                                     "`subset`",
                                     "Replacement assignment expression"))
knitr::kable(dplyr_vs_base, col.names = c("`dplyr`", 
                                          "Base R equivalent"))
```

You will see these alternatives used in older code examples.

## Alternative package::function notation

The `library` function is the most common way you'll access functions from R
packages, but it's not the only way to do this. There is another type of
notation that will allow you to use any external function from any package you
have installed on your computer: the `package::function` notation.

You can call functions by specifying the package name, a double colon, and then
the function name you want to use from that package. For example:

```{r}
stringr::str_to_upper("Astrophysicist")
```

## Alternative package::function notation

The `package::function` notation is not typically used because it substantially
increases how much you have to type in your code.

However, there can be cases where a function name is ambiguous. For example, you
might want to use functions from two different packages that have the same name.
In this case, using the `package::function` notation makes it crystal clear
which function you mean.

In practice, this problem is most likely to come up when you've loaded both
`plyr` and `dplyr`, which share several function names and are both popular
packages.

## Alternative package::function notation

There is another useful trick that you can do with the `package::function` notation.

RStudio has tab complemention, which means that once you start typing an object
name or function, if you press the Tab key, RStudio will give you suggestions
for how you can finish the name.

If you want to scroll through the names of all the external functions in a
package, you can do so by typing something like `?stringr::` in the console
(don't press Return) and then pressing the Tab key.






