---
title: "Exploring data 2"
header-includes:
   - \usepackage{hyperref}
   - \definecolor{links}{HTML}{2A1B81}
   - \hypersetup{colorlinks,linkcolor=,urlcolor=links}
output:
  beamer_presentation:
    theme: metropolis
fontsize: 10pt
---

```{r echo = FALSE, message = FALSE, warning = FALSE}
library(knitr)
library(ggplot2)
library(dplyr)
library(tidyr)
library(ggthemes)
library(faraway)
data(worldcup)
data(nepali)

library(tibble)
library(broom)
library("microbiome")
data("atlas1006")
tidy_samples <- atlas1006 %>% 
  get_sample() %>% 
  as.data.frame() %>% 
  rownames_to_column(var = "species") %>% 
  pivot_longer(-species, 
               names_to = "sample", 
               values_to = "prevalence")
```

# Functions

## Functions

As you move to larger projects, you will find yourself using the same code a lot. \bigskip

Examples include: 

- Reading in data from a specific type of equipment (air pollution monitor, accelerometer)
- Running a specific type of analysis (e.g., fitting the same model format to
many datasets)
- Creating a specific type of plot or map

\bigskip 

If you find yourself cutting and pasting a lot, convert the code to a function.

## Functions

Advantages of writing functions include: 

- Coding is more efficient
- Easier to change your code (if you've cut and paste code and you want to
change something, you have to change it everywhere)
- Easier to share code with others

## Functions

You can name a function anything you want, as long as you follow the naming
rules for all R objects (although try to avoid names of preexisting-existing
functions). You then specify any inputs (arguments; separate multiple arguments
with commas) and put the code to run in braces. You **define** a function as an
R object just like you do with other R objects (`<-`).

Here is the basic structure of "where things go" in an R function definition. 

```{r, eval = FALSE}
## Note: this code will not run
[function name] <- function([any arguments]){
        [code to run]
}
```

## Functions

Here is an example of a very basic function. This function takes a number as
the input (`number`) and adds 1 to that number. An R function will only return one R object. By
default, that object will be the last line of code in the function body.

```{r}
add_one <- function(number){
        number + 1 # Value returned by the function
}

add_one(number = 1:3)
add_one(number = -1)
```

## Functions

```{r eval = FALSE}
add_one <- function(number){
        number + 1 # Value returned by the function
}
```

- I picked the name of the function (`add_one`) (just like you pick what name
you want to use with any R object)
- The only input is a numeric vector. I pick the name I want to use for the
vector that is input to the function. I picked `number`.
- Within the code inside the function, the `number` refers to the numeric vector
object that the user passed into the function.

## Functions

As another example, you could write a small function to fit a specific model to
a dataframe you input and return the model object:

```{r}
fit_time_pos_mod <- function(df){
  lm(Tackles ~ Time + Position, 
     data = df) # Returns result from this call
}
```

- I picked the name of the function (`fit_time_pos_mod`) (just like you pick
what name you want to use with any R object)
- The only input is a dataframe. I pick the name I want to use for the dataframe
that is input to the function. I picked `df` (I often use this as a default
parameter name for a dataframe).
- Within the code inside the function, the `df` refers to the dataframe object
that the user passed into the function.

## Functions

Now you can apply that function within a tidy pipeline, for example to fit the
model to a specific subset of the data (the top four teams):

```{r}
data(worldcup)
worldcup %>%
  filter(Team %in% c("Spain", "Netherlands", 
                     "Uruguay", "Germany")) %>% 
  fit_time_pos_mod() %>% 
  tidy()
```

## Functions 

- Functions can input any type of R object (for example, vectors, data frames,
even other functions and ggplot objects)
- Similarly, functions can output any type of R object
- However, functions can only output one R object. If you have complex things
you want to output, a list might be a good choice for the output object type.
- Functions can have "side effects". Examples include printing something or
drawing a plot. Any action that a function takes *besides returning an R object*
is a "side effect".

## Functions---parameter defaults 

When defining a function, you can set default values for some of the parameters.
For example, in the `add_one` function, you can set the default value of the
`number` input to `0`.

```{r}
add_one <- function(number = 0){
        number + 1 # Value returned by the function
}
```

\small

Now, if someone runs the function without providing a value for `number`, the
function will use `0`. If they do provide a value for `number`, the function
will use that instead.

```{r}
add_one()    # Uses 0 for `number`
add_one(number = 3:5)   # Uses 5 for `number`
```

## Functions---parameters

You could write a function with no parameters:

```{r}
hello_world <- function(){
  print("Hello world!")
}

hello_world()
```

However, this will be pretty uncommon as you're first learning to write
functions.

## Functions---parameters

You can include multiple parameters, some with defaults and some without. For
example, you could write a function that inputs two numbers and adds them. If
you don't include a second value, `1` will be added as the second number:

```{r}
add_two_numbers <- function(first_number, second_number = 1){
  first_number + second_number
}
```

```{r}
add_two_numbers(first_number = 5:7, second_number = 5)
add_two_numbers(first_number = 5:7)
```

## Functions---the `return` function 

You can explicitly specify the value to return from the function (use `return`
function).

```{r eval = FALSE}
add_one <- function(number = 0){
        new_number <- number + 1 
        return(new_number)
}
```

If using `return` helps you think about what's happening with the code in your
function, you can use it. However, outside of a few exceptions, you usually
won't need to do it.

## `if` / `else`

In R, the `if` statement evaluates everything in the parentheses and, if that
evaluates to `TRUE`, runs everything in the braces. This means that you can
trigger code in an `if` statement with a single-value logical vector:

```{r}
tell_date <- function(){
  cat("Today's date is: ")
  cat(format(Sys.time(), "%b %d, %Y"))
  
  todays_wday <- lubridate::wday(Sys.time(),
                                 label = TRUE)
  if(todays_wday %in% c("Sat", "Sun")){
    cat("\n")
    cat("It's the weekend!")
  }
}
```

## `if` / `else`

```{r}
tell_date()
```

## `if` / `else`

You can add `else if` and `else` statements to tell R what to do if the
condition in the `if` statement isn't met.

For example, in the `tell_date` function, we might want to add some code so it
will print `"It's almost the weekend!"` on Fridays and how many days until
Saturday on other weekdays.

## `if` / `else`

```{r}
tell_date <- function(){
  # Print out today's date
  cat("Today's date is: ")
  cat(format(Sys.time(), "%b %d, %Y."), "\n")
  
  # Add something based on the weekday of today's date
  todays_wday <- lubridate::wday(Sys.time())
  
  if(todays_wday %in% c(1, 7)){      # What to do on Sat / Sun
    cat("It's the weekend!")
  } else if (todays_wday == c(6)) {  # What to do on Friday
    cat("It's almost the weekend!")
  } else {                           # What to do other days
    cat("It's ", 7 - todays_wday, "days until the weekend.")
  }
}
```

## `if` / `else`

```{r}
tell_date()
```

