---
title: 'Reporting data results #2'
output:
  slidy_presentation: default
  ioslides_presentation: default
  beamer_presentation:
    theme: metropolis
fontsize: 10pt
---

```{r echo = FALSE, message = FALSE, warning = FALSE}
library(knitr)
library(ggplot2)
library(dplyr)
library(tidyr)
library(ggthemes)
library(faraway)
data(worldcup)
library(gridExtra)
library(choroplethr)
library(choroplethrMaps)
library(purrr)
library(broom)

add_one <- function(number){
        number + 1 # Value returned by the function
}

fit_ht_wt_mod <- function(df){
  lm(wt ~ ht + sex, data = df) # Returns result from this call
}

nepali <- nepali %>%
  # Limit to certain columns
  select(id, sex, wt, ht, age) %>%
  # Convert id and sex to factors
  mutate(id = factor(id),
         sex = factor(sex, levels = c(1, 2),
                      labels = c("Male", "Female"))) %>%
  # Limit to first obs. per child
  distinct(id, .keep_all = TRUE)
```

# Functions

## Functions

As you move to larger projects, you will find yourself using the same code a lot. \bigskip

Examples include: 

- Reading in data from a specific type of equipment (air pollution monitor, accelerometer)
- Running a specific type of analysis (e.g., fitting the same model format to many datasets)
- Creating a specific type of plot or map

\bigskip 

If you find yourself cutting and pasting a lot, convert the code to a function.

## Functions

Advantages of writing functions include: 

- Coding is more efficient
- Easier to change your code (if you've cut and paste code and you want to change something, you have to change it everywhere)
- Easier to share code with others

## Functions

You can name a function anything you want, as long as you follow the naming rules for all R objects (although try to avoid names of preexisting-existing functions). You then specify any inputs (arguments; separate multiple arguments with commas) and put the code to run in braces. You **define** a function as an R object just like you do with other R objects (`<-`).

Here is the basic structure of "where things go" in an R function definition. 

```{r, eval = FALSE}
## Note: this code will not run
[function name] <- function([any arguments]){
        [code to run]
}
```

## Functions

Here is an example of a very basic function. This function takes a number as input and adds 1 to that number. An R function will only return one R object. By default, that object will be the last line of code in the function body.

```{r}
add_one <- function(number){
        number + 1 # Value returned by the function
}

add_one(number = 1:3)
add_one(number = -1)
```

## Functions

```{r eval = FALSE}
add_one <- function(number){
        number + 1 # Value returned by the function
}
```

- I picked the name of the function (`add_one`) (just like you pick what name you want to use with any R object)
- The only input is a numeric vector. I pick the name I want to use for the vector that is input to the function. I picked `number`.
- Within the code inside the function, the `number` refers to the numeric vector object that the user passed into the function.

## Functions

As another example, you could write a small function to fit a specific model to a dataframe you input and return the model object:

```{r}
fit_ht_wt_mod <- function(df){
  lm(wt ~ ht + sex, data = df) # Returns result from this call
}
```

- I picked the name of the function (`fit_ht_wt_mod`) (just like you pick what name you want to use with any R object)
- The only input is a dataframe. I pick the name I want to use for the dataframe that is input to the function. I picked `df` (I often use this as a default parameter name for a dataframe).
- Within the code inside the function, the `df` refers to the dataframe object that the user passed into the function.

## Functions

Now you can apply that function within a tidy pipeline, for example to fit the model to a specific subset of the data (all children with an age over 12 months):

```{r}
nepali %>%
  filter(age > 12) %>% 
  fit_ht_wt_mod() %>% 
  tidy()
```

## Functions 

- Functions can input any type of R object (for example, vectors, data frames, even other functions and ggplot objects)
- Similarly, functions can output any type of R object
- However, functions can only output one R object. If you have complex things you want to output, a list might be a good choice for the output object type.
- Functions can have "side effects". Examples include printing something or drawing a plot. Any action that a function takes *besides returning an R object* is a "side effect". 

## Functions-- parameter defaults 

When defining a function, you can set default values for some of the parameters. For example, in the `add_one` function, you can set the default value of the `number` input to `0`. 

```{r}
add_one <- function(number = 0){
        number + 1 # Value returned by the function
}
```

\small

Now, if someone runs the function without providing a value for `number`, the function will use `0`. If they do provide a value for `number`, the function will use that instead.

```{r}
add_one()    # Uses 0 for `number`
add_one(number = 3:5)   # Uses 5 for `number`
```

## Functions-- parameters

You could write a function with no parameters:

```{r}
hello_world <- function(){
  print("Hello world!")
}

hello_world()
```

However, this will be pretty uncommon as you're first learning to write functions. 

## Functions-- parameters

You can include multiple parameters, some with defaults and some without. For example, you could write a function that inputs two numbers and adds them. If you don't include a second value, `0` will be added as the second number:

```{r}
add_two_numbers <- function(first_number, second_number = 0){
  first_number + second_number
}
```

```{r}
add_two_numbers(first_number = 5:7, second_number = 0:2)
add_two_numbers(first_number = 5:7)
```

## Functions-- the `return` function 

You can explicitly specify the value to return from the function (use `return` function). 

```{r eval = FALSE}
add_one <- function(number = 0){
        new_number <- number + 1 
        return(new_number)
}
```

If using `return` helps you think about what's happening with the code in your function, you can use it. However, outside of a few exceptions, you usually won't need to do it.

## `if` / `else`

In R, the `if` statement evaluates everything in the parentheses and, if that evaluates to `TRUE`, runs everything in the braces. This means that you can trigger code in an `if` statement with a single-value logical vector: 

```{r}
tell_date <- function(){
  cat("Today's date is: ")
  cat(format(Sys.time(), "%b %d, %Y"))
  
  todays_wday <- lubridate::wday(Sys.time(),
                                 label = TRUE)
  if(todays_wday %in% c("Sat", "Sun")){
    cat("\n")
    cat("It's the weekend!")
  }
}
```

## `if` / `else`

```{r}
tell_date()
```

## `if` / `else`

You can add `else if` and `else` statements to tell R what to do if the condition in the `if` statement isn't met. 

For example, in the `tell_date` function, we might want to add some code so it will print `"It's almost the weekend!"` on Fridays and how many days until Saturday on other weekdays.

## `if` / `else`

```{r}
tell_date <- function(){
  # Print out today's date
  cat("Today's date is: ")
  cat(format(Sys.time(), "%b %d, %Y."), "\n")
  
  # Add something based on the weekday of today's date
  todays_wday <- lubridate::wday(Sys.time())
  
  if(todays_wday %in% c(1, 7)){      # What to do on Sat / Sun
    cat("It's the weekend!")
  } else if (todays_wday == c(6)) {  # What to do on Friday
    cat("It's almost the weekend!")
  } else {                           # What to do other days
    cat("It's ", 7 - todays_wday, "days until the weekend.")
  }
}
```

## `if` / `else`

```{r}
tell_date()
```

# Functional programming

## Applying a function repeatedly

You will often want to apply a function multiple times with different input. A powerful way to do this in R is using `lists`. We can use functions from the `purrr` library to `map` functions to each element of a list.

For example, you could use `map` to apply our `add_one` function separately to 1, 2, and 3 by using `map` on a vector with those values:

```{r message = FALSE}
library(purrr)
my_list <- list(a = 1:2, b = 3:5)
map(my_list, add_one)
```

## Applying a function repeatedly

One of the most common `repeated` tasks you might want to do is loading in multiple data files:

```{r}
library(readr)
library(dplyr)
library(purrr)
files <- list.files("../data/running", pattern = ".csv", full.names = TRUE)
df <- map(files, read_csv, col_types = cols(.default = "c")) %>%
      bind_rows()

head(df, 15)
```

## Applying a function repeatedly

We can use our own custom functions with `map()`:

```{r, include=FALSE}
clean_names <- function(x){gsub(" ", "_", tolower(x))}
```

```{r, warning=FALSE, message=FALSE}
library(lubridate)

read_run_csv <- function(file){
  read_csv(file, col_types = cols(.default = "c")) %>%
  rename_all(clean_names) %>%
  filter(distance == "1" & split != "Summary") %>%
  mutate_at(vars(matches("time|pace")), hms) %>%
  mutate_if(is.character, as.numeric) %>%
  mutate(file_name = sub(".csv", "", basename(file)))
  
}
read_run_csv(files[1])
```

## Applying a function repeatedly

Here we `map` our function `read_run_csv` over each element in the vector of `files`:

```{r, warning=FALSE}
rd <- suppressWarnings(map(files, read_run_csv))

class(rd)

length(rd)
```

## Applying a function repeatedly

We now have a list of data frames! We can `map` functions to each element in our lists of data frames:

```{r}
run_sum <- function(x){
  summarise(x, 
            distance = sum(distance, na.rm = TRUE),
            gain = sum(elevation_gain, na.rm = TRUE),
            gain_per_mile = mean(elevation_gain, na.rm = TRUE),
            loss_per_mile = mean(elev_loss, na.rm = TRUE),
            loss = sum(elev_loss, na.rm = TRUE),
            pace_mins = mean(as.numeric(avg_pace) / 60, na.rm = TRUE),
            t_min = min(avg_temperature, na.rm = TRUE),
            t_max = max(avg_temperature, na.rm = TRUE),
            t_mean = mean(avg_temperature, na.rm = TRUE),
            file_name = first(file_name))
}
```

```{r}
run_summary <- map_df(rd, run_sum)
run_summary
```

## Applying a function repeatedly

And just for fun we can do the same thing with a different dataset. Here we see that our `map` functions are useful for dealing with the objects created when loading `XML` files:

```{r, eval = FALSE}
library(XML)

read_tcx <- function(x){
  out <- suppressWarnings(getNodeSet(xmlParse(x), "//ns:Trackpoint", "ns") %>%
  map(xmlToList) %>%
  map_df(data.frame) %>%
  mutate(file_name = sub(".tcx", "", basename(x))))
  print(x)
  return(out)
}

files_xml <- list.files("../data/running", pattern = ".tcx", full.names = TRUE)

tcx_data <- map(files_xml, read_tcx)
```

## Applying a function repeatedly

This time we want to calculate statistics on the altitude:

```{r, eval = FALSE}
tcx_sum <- function(x){
  summarise(x, 
            e_min = min(as.numeric(AltitudeMeters), na.rm = TRUE),
            e_max = max(as.numeric(AltitudeMeters), na.rm = TRUE),
            e_mean = mean(as.numeric(AltitudeMeters), na.rm = TRUE),
            file_name = first(file_name))
}
```

We use `map_df()` to create a new data frame:

```{r, eval = FALSE}
tcx_summary <- map_df(tcx_data, tcx_sum)
head(tcx_summary,3)
```

```{r, eval=FALSE, include=FALSE}
  write_csv(tcx_summary, "../data/elevation.csv")
```

```{r, echo=FALSE}
  elevation <- read_csv("../data/elevation.csv", col_types = cols())
```

## Applying a function repeatedly

We'd like to see which variable best predicts pace. First let's organize our data into a suitable format:

```{r}
library(tidyr)
  run <- left_join(run_summary, elevation, by = "file_name") %>%
         select(-file_name) %>%
         gather("var", "value", -pace_mins)

head(run, 3)
```

## Applying a function repeatedly

We can create a model function and regress each variable using `nest()`:

```{r}
model_lm <- function(df) {
  lm(pace_mins ~ value, data = df)
}
```

Fit models:

```{r}
lm_models <- run %>%
             group_by(var) %>%
             nest() %>%
             mutate(mod = map(data, model_lm))

head(lm_models, 2)
```

## Applying a function repeatedly

We can use `map_df` and our friend the `broom` package to create tidy model output for **all** the models combined:

```{r}
library(broom)
results <- map_df(lm_models$mod, glance) %>%
           arrange(-r.squared) %>%
           mutate(variable = lm_models$var)

kable(results, digits = 2)
```

## Applying a function repeatedly

Finally, you might want to `map` over multiple inputs in parallel. You can do this using the `pmap()` function:

```{r}
txt_fun <- function(distance, gain, loss, pace_mins) {
  paste("Nick ran for", distance, "miles up", gain, "and down", loss, "feet in",
        round(distance * pace_mins, 0), "minutes")
}
```

```{r}
run_summary %>%
  select(distance, gain, loss, pace_mins) %>%
  pmap_chr(txt_fun) %>% 
  tail()
```

## Extra mapping stuff (to move to another week)


```{r eval = FALSE}
library(acs)
api.key.install('[your census api key]');
```

## Choropleths in R

You can pick from a large number of American Community Survey tables-- [see here](http://factfinder.census.gov/faces/affhelp/jsf/pages/metadata.xhtml?lang=en&type=dataset&id=dataset.en.ACS_12_5YR) for the list plus ID numbers. If the table has multiple columns, you will be prompted to select which one you want to plot. \bigskip

## Choropleths in R

For example, table B19301 gives per-capita income, so if you wanted to plot that, you could run:

```{r fig.width = 3, fig.height = 2.75, message = FALSE, eval = FALSE}
county_choropleth_acs(tableId = "B19301",
                       state_zoom = c("wyoming",
                                      "colorado"))

```

## Choropleths in R

```{r fig.width = 10, fig.height = 6, message = FALSE, echo = FALSE, fig.align = "center"}
county_choropleth_acs(tableId = "B19301",
                       state_zoom = c("wyoming",
                                      "colorado"))

```


## Google Maps API

The `ggmap` package allows you to use tools from Google Maps directly from R.

```{r message=FALSE, warning = FALSE}
## install.packages("ggmap")
library(ggmap)
```

This package uses the Google Maps API, so you should read their [terms of service](http://developers.google.com/maps/terms) and make sure you follow them. In particular, you are limited to just a certain number of queries per time. 

## Google Maps API

You can use the `get_map` function to get maps for different locations. \bigskip 

You can either use the longitude and latitude of the center point of the map, along with the `zoom` option to say how much to zoom in (3: continent to 20: building) or you can use a character string to specify a location. \bigskip

If you do the second, `get_map` will actually use the Google Maps API to geocode the string to a latitude and longitude and then get the map (you can imagine that this is like searching in Google Maps in the search box for a location).

```{r message = FALSE, warning = FALSE, fig.width = 3.5, fig.height = 3.5, fig.align = "center", eval = FALSE}
beijing <- get_map("Beijing", zoom = 12)
ggmap(beijing)
```

## Google Maps API

```{r message = FALSE, warning = FALSE, fig.width = 3.5, fig.height = 3.5, fig.align = "center", echo = FALSE}
beijing <- get_map("Beijing", zoom = 12)
ggmap(beijing)
```

## Google Maps API

With this package, you can get maps from the following different sources:

- Google Maps
- OpenStreetMap
- Stamen Maps
- CloudMade Maps (You may need a separate API key for this)

## Google Maps API

Here are different examples of Beijing using different map sources. (Also, note that I'm using the option `extent = "device"` to fill up the whole plot are with the map, instead of including axis labels and titles.)

```{r message = FALSE, warning = FALSE}
beijing_a <- get_map("Beijing", zoom = 12,
                     source = "stamen", maptype = "toner")
a <- ggmap(beijing_a, extent = "device")

beijing_b <- get_map("Beijing", zoom = 12,
                     source = "stamen", maptype = "watercolor")
b <- ggmap(beijing_b, extent = "device")

beijing_c <- get_map("Beijing", zoom = 12,
                     source = "google", maptype = "hybrid")
c <- ggmap(beijing_c, extent = "device")
```

## Google Maps API

```{r fig.width = 7}
grid.arrange(a, b, c, nrow = 1) 
```


## Google Maps API

As with the maps from `ggplot2`, you can add points to these maps:

```{r warning = FALSE, message = FALSE, echo = FALSE}
library(tidyr)
serial_phone <- read.csv("../data/serial_phone_data.csv") %>% 
        mutate(Cell_Site = substring(Cell_Site, 1, 4),
               Call_Time = as.POSIXct(Call_Time, format = "%d/%m/%y %H:%M",
                                      tz = "EST")) %>%
        left_join(serial, by = c("Cell_Site" = "Name")) %>%
        select(Person_Called, Call_Time, Duration, long, lat) %>%
        filter(!(Person_Called %in% c("incoming", "# + Adnan cell"))) %>%
        arrange(Call_Time)
```

```{r message = FALSE, warning = FALSE, fig.width = 5}
serial_map <- get_map(c(-76.7, 39.3), zoom = 12,
                      source = "stamen",
                      maptype = "toner")
serial_map <- ggmap(serial_map, extent = "device") + 
        geom_point(data = serial_phone,
                   aes(x = long, y = lat),
                   color = "red", size = 3,
                   alpha = 0.4) +
        geom_point(data = subset(serial,
                                 Type != "cell-site"),
                   aes(x = long, y = lat),
                   color = "darkgoldenrod1",
                   size = 2)
```

## Google Maps API

```{r warning = FALSE, message = FALSE, fig.width = 4, fig.height = 4, fig.align = "center", echo = FALSE}
serial_map
```


## Google Maps API

You can also use the Google Maps API, through the `geocode` function, to get the latitude and longitude of specific locations. Basically, if the string would give you the right location if you typed it in Google Maps, `geocode` should be able to geocode it.

For example, you can get the location of CSU:

```{r message = FALSE, warning = FALSE}
geocode("Colorado State University")
```

## Google Maps API

You can also get a location by address through this:

```{r message = FALSE, warning = FALSE}
geocode("1 First St NE, Washington, DC")
```

## Google Maps API

You can get distances, too, using the `mapdist` function with two locations. This will give you distance and also time. 

```{r message = FALSE, warning = FALSE}
mapdist("Fort Collins CO",
        "1 First St NE, Washington, DC") %>%
  select(from, miles, hours)
```
