---
title: 'Reporting data results #2'
output:
  ioslides_presentation: default
  beamer_presentation:
    theme: metropolis
fontsize: 10pt
---

```{r echo = FALSE, message = FALSE, warning = FALSE}
library(knitr)
library(ggplot2)
library(dplyr)
library(tidyr)
library(ggthemes)
library(faraway)
data(worldcup)
library(gridExtra)
library(choroplethr)
library(choroplethrMaps)
library(purrr)
library(broom)

add_one <- function(number){
        number + 1 # Value returned by the function
}

fit_ht_wt_mod <- function(df){
  lm(wt ~ ht + sex, data = df) # Returns result from this call
}

nepali <- nepali %>%
  # Limit to certain columns
  select(id, sex, wt, ht, age) %>%
  # Convert id and sex to factors
  mutate(id = factor(id),
         sex = factor(sex, levels = c(1, 2),
                      labels = c("Male", "Female"))) %>%
  # Limit to first obs. per child
  distinct(id, .keep_all = TRUE)
```

# Functions

## Functions

As you move to larger projects, you will find yourself using the same code a lot. \bigskip

Examples include: 

- Reading in data from a specific type of equipment (air pollution monitor, accelerometer)
- Running a specific type of analysis (e.g., fitting the same model format to many datasets)
- Creating a specific type of plot or map

\bigskip 

If you find yourself cutting and pasting a lot, convert the code to a function.

## Functions

Advantages of writing functions include: 

- Coding is more efficient
- Easier to change your code (if you've cut and paste code and you want to change something, you have to change it everywhere)
- Easier to share code with others

## Functions

You can name a function anything you want, as long as you follow the naming rules for all R objects (although try to avoid names of preexisting-existing functions). You then specify any inputs (arguments; separate multiple arguments with commas) and put the code to run in braces. You **define** a function as an R object just like you do with other R objects (`<-`).

Here is the basic structure of "where things go" in an R function definition. 

```{r, eval = FALSE}
## Note: this code will not run
[function name] <- function([any arguments]){
        [code to run]
}
```

## Functions

Here is an example of a very basic function. This function takes a number as input and adds 1 to that number. An R function will only return one R object. By default, that object will be the last line of code in the function body.

```{r}
add_one <- function(number){
        number + 1 # Value returned by the function
}

add_one(number = 1:3)
add_one(number = -1)
```

## Functions

```{r eval = FALSE}
add_one <- function(number){
        number + 1 # Value returned by the function
}
```

- I picked the name of the function (`add_one`) (just like you pick what name you want to use with any R object)
- The only input is a numeric vector. I pick the name I want to use for the vector that is input to the function. I picked `number`.
- Within the code inside the function, the `number` refers to the numeric vector object that the user passed into the function.

## Functions

As another example, you could write a small function to fit a specific model to a dataframe you input and return the model object:

```{r}
fit_ht_wt_mod <- function(df){
  lm(wt ~ ht + sex, data = df) # Returns result from this call
}
```

- I picked the name of the function (`fit_ht_wt_mod`) (just like you pick what name you want to use with any R object)
- The only input is a dataframe. I pick the name I want to use for the dataframe that is input to the function. I picked `df` (I often use this as a default parameter name for a dataframe).
- Within the code inside the function, the `df` refers to the dataframe object that the user passed into the function.

## Functions

Now you can apply that function within a tidy pipeline, for example to fit the model to a specific subset of the data (all children with an age over 12 months):

```{r}
nepali %>% 
  filter(age > 12) %>% 
  fit_ht_wt_mod() %>% 
  tidy()
```

## Functions 

- Functions can input any type of R object (for example, vectors, data frames, even other functions and ggplot objects)
- Similarly, functions can output any type of R object
- However, functions can only output one R object. If you have complex things you want to output, a list might be a good choice for the output object type.
- Functions can have "side effects". Examples include printing something or drawing a plot. Any action that a function takes *besides returning an R object* is a "side effect". 

## Functions-- parameter defaults 

When defining a function, you can set default values for some of the parameters. For example, in the `add_one` function, you can set the default value of the `number` input to `0`. 

```{r}
add_one <- function(number = 0){
        number + 1 # Value returned by the function
}
```

\small

Now, if someone runs the function without providing a value for `number`, the function will use `0`. If they do provide a value for `number`, the function will use that instead.

```{r}
add_one()    # Uses 0 for `number`
add_one(number = 3:5)   # Uses 5 for `number`
```

## Functions-- parameters

You could write a function with no parameters:

```{r}
hello_world <- function(){
  print("Hello world!")
}

hello_world()
```

However, this will be pretty uncommon as you're first learning to write functions. 

## Functions-- parameters

You can include multiple parameters, some with defaults and some without. For example, you could write a function that inputs two numbers and adds them. If you don't include a second value, `0` will be added as the second number:

```{r}
add_two_numbers <- function(first_number, second_number = 0){
  first_number + second_number
}
```

```{r}
add_two_numbers(first_number = 5:7, second_number = 0:2)
add_two_numbers(first_number = 5:7)
```

## Functions-- the `return` function 

You can explicitly specify the value to return from the function (use `return` function). 

```{r eval = FALSE}
add_one <- function(number = 0){
        new_number <- number + 1 
        return(new_number)
}
```

If using `return` helps you think about what's happening with the code in your function, you can use it. However, outside of a few exceptions, you usually won't need to do it.

## Functions-- Error checking

There are ways to check for errors in the arguments a user inputs to the function. One useful check is to see if user inputs are in the required class. 

The `assertive` package has some functions that you can use for common checks of the inputs to a function. If someone inputs something of the wrong class, it will give a useful error message. For example, the `assert_is_numeric` function will check if an object is in a numeric class. If so, it will do nothing. If not, it will return an error message:

```{r error = TRUE}
library(assertive)
assert_is_numeric(1:3)
assert_is_numeric(c("a", "b"))
```

## Functions-- Error checking

You could add this in the code for the `add_one` function, so a useful error message will be returned if a user tries to input something besides a numeric vector for `number`:

```{r}
add_one <- function(number){
  assert_is_numeric(number)
  number + 1 
}
```

```{r error = TRUE}
add_one(number = 1:3)
add_one(number = c("a", "b"))
```

## Functions-- Error checking

I would recommend that you not worry about this too much when you're learning to write functions for your own use. 

However, once you have mastered the basics of writing functions and start writing them for others to use, you'll want to start incorporating this. 

## `if` / `else`

In R, the `if` statement evaluates everything in the parentheses and, if that evaluates to `TRUE`, runs everything in the braces. This means that you can trigger code in an `if` statement with a single-value logical vector: 

```{r}
tell_date <- function(){
  cat("Today's date is: ")
  cat(format(Sys.time(), "%b %d, %Y"))
  
  todays_wday <- lubridate::wday(Sys.time(),
                                 label = TRUE)
  if(todays_wday %in% c("Sat", "Sun")){
    cat("\n")
    cat("It's the weekend!")
  }
}
```

## `if` / `else`

```{r}
tell_date()
```

## `if` / `else`

You can add `else if` and `else` statements to tell R what to do if the condition in the `if` statement isn't met. 

For example, in the `tell_date` function, we might want to add some code so it will print `"It's almost the weekend!"` on Fridays and how many days until Saturday on other weekdays.

## `if` / `else`

```{r}
tell_date <- function(){
  # Print out today's date
  cat("Today's date is: ")
  cat(format(Sys.time(), "%b %d, %Y."), "\n")
  
  # Add something based on the weekday of today's date
  todays_wday <- lubridate::wday(Sys.time())
  
  if(todays_wday %in% c(1, 7)){      # What to do on Sat / Sun
    cat("It's the weekend!")
  } else if (todays_wday == c(6)) {  # What to do on Friday
    cat("It's almost the weekend!")
  } else {                           # What to do other days
    cat("It's ", 7 - todays_wday, "days until the weekend.")
  }
}
```

## `if` / `else`

```{r}
tell_date()
```

# Functional programming

## Applying a function repeatedly 

One way that functions are really useful is that you can use the `map` family of functions from the `purrr` package to apply that function to all elements in a vector of a list (remember, a list could hold lots of similar dataframes). 

For example, you could use `map` to apply the `add_one` function separately to 1, 2, and 3 by using `map` on a vector with those values:

```{r message = FALSE}
library(purrr)
my_list <- list(a = 1:2, b = 3:5)
map(my_list, add_one)
```

## Applying a function repeatedly 

This can also be very useful if you have a dataframe for which you would like to apply the same function to subsets of the data. For example, for the `nepali` data you may want to apply the model of weight regressed on height and sex separately for children 12 months and younger versus older children. 

First, we can add a factor variable that specifies whether the child is younger or older than 12 months:

```{r}
nepali <- nepali %>% 
  mutate(young = age < 12,
         young = factor(young, levels = c(TRUE, FALSE), 
                        labels = c("younger", "older"))) 
```

## Applying a function repeatedly 

```{r}
nepali %>% 
  slice(1:3)
```


## Applying a function repeatedly 

Then, you can use the `nest` function to "nest" a dataframe by a factor variable. This function will create a column that actually stores its own dataframe:

```{r}
nested_nepali <- nepali %>% 
  group_by(young) %>% 
  nest()
nested_nepali
```

## Applying a function repeatedly 

Each element of the `data` column in the nested dataset is actually a full dataframe:

```{r}
nested_nepali$data[[1]] %>% slice(1:3)
```

## Applying a function repeatedly 

Now, you can use `map` to apply the modeling function to each of these subsets of the dataframe. Use `mutate` to add a column with the results. We can also add a column with the results of applying `augment` to each of the model results:

\small

```{r message = FALSE, warning = FALSE}
modeled_nepali <- nepali %>% 
  group_by(young) %>% 
  nest() %>% 
  mutate(mod_results = map(data, fit_ht_wt_mod),
         augmented_data = map(mod_results, augment))
modeled_nepali
```

## Applying a function repeatedly 

Each element of the `mod_results` column is the output from an `lm` model:

```{r}
modeled_nepali$mod_results[[1]]
```

## Applying a function repeatedly 

We can apply `tidy` and `glance` to this output, just like we did with the output from fitting a single model:

```{r}
tidy(modeled_nepali$mod_results[[1]])
```

## Applying a function repeatedly 

To get back to a regular dataframe, you can "unnest". Before you do this, you should limit the data to only columns that will "unnest" to the same number of rows (if you have multiple columns with dataframes or lists in them). For example, we could unnest the `augmented_data` column, so we can plot observed data versus the model fit:

```{r}
modeled_nepali %>% 
  select(young, augmented_data) %>% 
  unnest() %>% 
  slice(1:3) %>% select(1:7)
```

## Applying a function repeatedly 

\small

```{r out.width = "\\textwidth", fig.align = "center", fig.width = 6, fig.height = 2}
modeled_nepali %>% 
  select(young, augmented_data) %>% 
  unnest() %>% 
  ggplot(aes(x = ht, y = wt, group = sex, color = sex)) + 
  geom_point(size = 0.8) + 
  geom_line(aes(y = .fitted), size = 1.2) + 
  facet_wrap(~ young, ncol = 2)
```

## Applying a function repeatedly

The full pipe for creating this figure is: 

```{r eval = FALSE}
nepali %>%
  group_by(young) %>%    # Group by young and nest 
  nest() %>%             # to model each group of "young"
  mutate(mod_results = map(data, fit_ht_wt_mod), 
         augmented_data = map(mod_results, augment)) %>% 
  select(young, augmented_data) %>%  
  unnest %>%             # Unnest `augmented_data` (nested)
  ggplot(aes(x = ht, y = wt, group = sex, color = sex)) + 
  geom_point(size = 0.8) + 
  geom_line(aes(y = .fitted), size = 1.2) + 
  facet_wrap(~ young, ncol = 2)
```

## Extra mapping stuff (to move to another week)

```{r eval = FALSE}
library(acs)
api.key.install('[your census api key]');
```

## Choropleths in R

You can pick from a large number of American Community Survey tables-- [see here](http://factfinder.census.gov/faces/affhelp/jsf/pages/metadata.xhtml?lang=en&type=dataset&id=dataset.en.ACS_12_5YR) for the list plus ID numbers. If the table has multiple columns, you will be prompted to select which one you want to plot. \bigskip

## Choropleths in R

For example, table B19301 gives per-capita income, so if you wanted to plot that, you could run:

```{r fig.width = 3, fig.height = 2.75, message = FALSE, eval = FALSE}
county_choropleth_acs(tableId = "B19301",
                       state_zoom = c("wyoming",
                                      "colorado"))

```

## Choropleths in R

```{r fig.width = 10, fig.height = 6, message = FALSE, echo = FALSE, fig.align = "center"}
county_choropleth_acs(tableId = "B19301",
                       state_zoom = c("wyoming",
                                      "colorado"))

```


## Google Maps API

The `ggmap` package allows you to use tools from Google Maps directly from R.

```{r message=FALSE, warning = FALSE}
## install.packages("ggmap")
library(ggmap)
```

This package uses the Google Maps API, so you should read their [terms of service](http://developers.google.com/maps/terms) and make sure you follow them. In particular, you are limited to just a certain number of queries per time. 

## Google Maps API

You can use the `get_map` function to get maps for different locations. \bigskip 

You can either use the longitude and latitude of the center point of the map, along with the `zoom` option to say how much to zoom in (3: continent to 20: building) or you can use a character string to specify a location. \bigskip

If you do the second, `get_map` will actually use the Google Maps API to geocode the string to a latitude and longitude and then get the map (you can imagine that this is like searching in Google Maps in the search box for a location).

```{r message = FALSE, warning = FALSE, fig.width = 3.5, fig.height = 3.5, fig.align = "center", eval = FALSE}
beijing <- get_map("Beijing", zoom = 12)
ggmap(beijing)
```

## Google Maps API

```{r message = FALSE, warning = FALSE, fig.width = 3.5, fig.height = 3.5, fig.align = "center", echo = FALSE}
beijing <- get_map("Beijing", zoom = 12)
ggmap(beijing)
```

## Google Maps API

With this package, you can get maps from the following different sources:

- Google Maps
- OpenStreetMap
- Stamen Maps
- CloudMade Maps (You may need a separate API key for this)

## Google Maps API

Here are different examples of Beijing using different map sources. (Also, note that I'm using the option `extent = "device"` to fill up the whole plot are with the map, instead of including axis labels and titles.)

```{r message = FALSE, warning = FALSE}
beijing_a <- get_map("Beijing", zoom = 12,
                     source = "stamen", maptype = "toner")
a <- ggmap(beijing_a, extent = "device")

beijing_b <- get_map("Beijing", zoom = 12,
                     source = "stamen", maptype = "watercolor")
b <- ggmap(beijing_b, extent = "device")

beijing_c <- get_map("Beijing", zoom = 12,
                     source = "google", maptype = "hybrid")
c <- ggmap(beijing_c, extent = "device")
```

## Google Maps API

```{r fig.width = 7}
grid.arrange(a, b, c, nrow = 1) 
```


## Google Maps API

As with the maps from `ggplot2`, you can add points to these maps:

```{r warning = FALSE, message = FALSE, echo = FALSE}
library(tidyr)
serial_phone <- read.csv("../data/serial_phone_data.csv") %>% 
        mutate(Cell_Site = substring(Cell_Site, 1, 4),
               Call_Time = as.POSIXct(Call_Time, format = "%d/%m/%y %H:%M",
                                      tz = "EST")) %>%
        left_join(serial, by = c("Cell_Site" = "Name")) %>%
        select(Person_Called, Call_Time, Duration, long, lat) %>%
        filter(!(Person_Called %in% c("incoming", "# + Adnan cell"))) %>%
        arrange(Call_Time)
```

```{r message = FALSE, warning = FALSE, fig.width = 5}
serial_map <- get_map(c(-76.7, 39.3), zoom = 12,
                      source = "stamen",
                      maptype = "toner")
serial_map <- ggmap(serial_map, extent = "device") + 
        geom_point(data = serial_phone,
                   aes(x = long, y = lat),
                   color = "red", size = 3,
                   alpha = 0.4) +
        geom_point(data = subset(serial,
                                 Type != "cell-site"),
                   aes(x = long, y = lat),
                   color = "darkgoldenrod1",
                   size = 2)
```

## Google Maps API

```{r warning = FALSE, message = FALSE, fig.width = 4, fig.height = 4, fig.align = "center", echo = FALSE}
serial_map
```


## Google Maps API

You can also use the Google Maps API, through the `geocode` function, to get the latitude and longitude of specific locations. Basically, if the string would give you the right location if you typed it in Google Maps, `geocode` should be able to geocode it.

For example, you can get the location of CSU:

```{r message = FALSE, warning = FALSE}
geocode("Colorado State University")
```

## Google Maps API

You can also get a location by address through this:

```{r message = FALSE, warning = FALSE}
geocode("1 First St NE, Washington, DC")
```

## Google Maps API

You can get distances, too, using the `mapdist` function with two locations. This will give you distance and also time. 

```{r message = FALSE, warning = FALSE}
mapdist("Fort Collins CO",
        "1 First St NE, Washington, DC") %>%
  select(from, miles, hours)
```
