---
title: 'Reporting data results #2'
output:
  slidy_presentation: default
  ioslides_presentation: default
  beamer_presentation:
    theme: metropolis
fontsize: 10pt
---

```{r echo = FALSE, message = FALSE, warning = FALSE}
library(knitr)
library(ggplot2)
library(dplyr)
library(tidyr)
library(ggthemes)
library(faraway)
data(worldcup)
library(gridExtra)
library(choroplethr)
library(choroplethrMaps)
library(purrr)
library(broom)

add_one <- function(number){
        number + 1 # Value returned by the function
}

fit_ht_wt_mod <- function(df){
  lm(wt ~ ht + sex, data = df) # Returns result from this call
}

nepali <- nepali %>%
  # Limit to certain columns
  select(id, sex, wt, ht, age) %>%
  # Convert id and sex to factors
  mutate(id = factor(id),
         sex = factor(sex, levels = c(1, 2),
                      labels = c("Male", "Female"))) %>%
  # Limit to first obs. per child
  distinct(id, .keep_all = TRUE)
```

# Functions

## Functions

As you move to larger projects, you will find yourself using the same code a lot. \bigskip

Examples include: 

- Reading in data from a specific type of equipment (air pollution monitor, accelerometer)
- Running a specific type of analysis (e.g., fitting the same model format to many datasets)
- Creating a specific type of plot or map

\bigskip 

If you find yourself cutting and pasting a lot, convert the code to a function.

## Functions

Advantages of writing functions include: 

- Coding is more efficient
- Easier to change your code (if you've cut and paste code and you want to change something, you have to change it everywhere)
- Easier to share code with others

## Functions

You can name a function anything you want, as long as you follow the naming rules for all R objects (although try to avoid names of preexisting-existing functions). You then specify any inputs (arguments; separate multiple arguments with commas) and put the code to run in braces. You **define** a function as an R object just like you do with other R objects (`<-`).

Here is the basic structure of "where things go" in an R function definition. 

```{r, eval = FALSE}
## Note: this code will not run
[function name] <- function([any arguments]){
        [code to run]
}
```

## Functions

Here is an example of a very basic function. This function takes a number as input and adds 1 to that number. An R function will only return one R object. By default, that object will be the last line of code in the function body.

```{r}
add_one <- function(number){
        number + 1 # Value returned by the function
}

add_one(number = 1:3)
add_one(number = -1)
```

## Functions

```{r eval = FALSE}
add_one <- function(number){
        number + 1 # Value returned by the function
}
```

- I picked the name of the function (`add_one`) (just like you pick what name you want to use with any R object)
- The only input is a numeric vector. I pick the name I want to use for the vector that is input to the function. I picked `number`.
- Within the code inside the function, the `number` refers to the numeric vector object that the user passed into the function.

## Functions

As another example, you could write a small function to fit a specific model to a dataframe you input and return the model object:

```{r}
fit_ht_wt_mod <- function(df){
  lm(wt ~ ht + sex, data = df) # Returns result from this call
}
```

- I picked the name of the function (`fit_ht_wt_mod`) (just like you pick what name you want to use with any R object)
- The only input is a dataframe. I pick the name I want to use for the dataframe that is input to the function. I picked `df` (I often use this as a default parameter name for a dataframe).
- Within the code inside the function, the `df` refers to the dataframe object that the user passed into the function.

## Functions

Now you can apply that function within a tidy pipeline, for example to fit the model to a specific subset of the data (all children with an age over 12 months):

```{r}
nepali %>% 
  filter(age > 12) %>% 
  fit_ht_wt_mod() %>% 
  tidy()
```

## Functions 

- Functions can input any type of R object (for example, vectors, data frames, even other functions and ggplot objects)
- Similarly, functions can output any type of R object
- However, functions can only output one R object. If you have complex things you want to output, a list might be a good choice for the output object type.
- Functions can have "side effects". Examples include printing something or drawing a plot. Any action that a function takes *besides returning an R object* is a "side effect". 

## Functions-- parameter defaults 

When defining a function, you can set default values for some of the parameters. For example, in the `add_one` function, you can set the default value of the `number` input to `0`. 

```{r}
add_one <- function(number = 0){
        number + 1 # Value returned by the function
}
```

\small

Now, if someone runs the function without providing a value for `number`, the function will use `0`. If they do provide a value for `number`, the function will use that instead.

```{r}
add_one()    # Uses 0 for `number`
add_one(number = 3:5)   # Uses 5 for `number`
```

## Functions-- parameters

You could write a function with no parameters:

```{r}
hello_world <- function(){
  print("Hello world!")
}

hello_world()
```

However, this will be pretty uncommon as you're first learning to write functions. 

## Functions-- parameters

You can include multiple parameters, some with defaults and some without. For example, you could write a function that inputs two numbers and adds them. If you don't include a second value, `0` will be added as the second number:

```{r}
add_two_numbers <- function(first_number, second_number = 0){
  first_number + second_number
}
```

```{r}
add_two_numbers(first_number = 5:7, second_number = 0:2)
add_two_numbers(first_number = 5:7)
```

## Functions-- the `return` function 

You can explicitly specify the value to return from the function (use `return` function). 

```{r eval = FALSE}
add_one <- function(number = 0){
        new_number <- number + 1 
        return(new_number)
}
```

If using `return` helps you think about what's happening with the code in your function, you can use it. However, outside of a few exceptions, you usually won't need to do it.

## Functions-- Error checking

There are ways to check for errors in the arguments a user inputs to the function. One useful check is to see if user inputs are in the required class. 

The `assertive` package has some functions that you can use for common checks of the inputs to a function. If someone inputs something of the wrong class, it will give a useful error message. For example, the `assert_is_numeric` function will check if an object is in a numeric class. If so, it will do nothing. If not, it will return an error message:

```{r error = TRUE}
library(assertive)
assert_is_numeric(1:3)
assert_is_numeric(c("a", "b"))
```

## Functions-- Error checking

You could add this in the code for the `add_one` function, so a useful error message will be returned if a user tries to input something besides a numeric vector for `number`:

```{r}
add_one <- function(number){
  assert_is_numeric(number)
  number + 1 
}
```

```{r error = TRUE}
add_one(number = 1:3)
add_one(number = c("a", "b"))
```

## Functions-- Error checking

I would recommend that you not worry about this too much when you're learning to write functions for your own use. 

However, once you have mastered the basics of writing functions and start writing them for others to use, you'll want to start incorporating this. 

## `if` / `else`

In R, the `if` statement evaluates everything in the parentheses and, if that evaluates to `TRUE`, runs everything in the braces. This means that you can trigger code in an `if` statement with a single-value logical vector: 

```{r}
tell_date <- function(){
  cat("Today's date is: ")
  cat(format(Sys.time(), "%b %d, %Y"))
  
  todays_wday <- lubridate::wday(Sys.time(),
                                 label = TRUE)
  if(todays_wday %in% c("Sat", "Sun")){
    cat("\n")
    cat("It's the weekend!")
  }
}
```

## `if` / `else`

```{r}
tell_date()
```

## `if` / `else`

You can add `else if` and `else` statements to tell R what to do if the condition in the `if` statement isn't met. 

For example, in the `tell_date` function, we might want to add some code so it will print `"It's almost the weekend!"` on Fridays and how many days until Saturday on other weekdays.

## `if` / `else`

```{r}
tell_date <- function(){
  # Print out today's date
  cat("Today's date is: ")
  cat(format(Sys.time(), "%b %d, %Y."), "\n")
  
  # Add something based on the weekday of today's date
  todays_wday <- lubridate::wday(Sys.time())
  
  if(todays_wday %in% c(1, 7)){      # What to do on Sat / Sun
    cat("It's the weekend!")
  } else if (todays_wday == c(6)) {  # What to do on Friday
    cat("It's almost the weekend!")
  } else {                           # What to do other days
    cat("It's ", 7 - todays_wday, "days until the weekend.")
  }
}
```

## `if` / `else`

```{r}
tell_date()
```

# Functional programming

## Applying a function repeatedly

You will often want to apply a function multiple times with different input. A powerful way to do this in R is using `lists`. We can use functions from the `purrr` library to `map` functions to each element of a list.

For example, you could use `map` to apply our `add_one` function separately to 1, 2, and 3 by using `map` on a vector with those values:

```{r message = FALSE}
library(purrr)
my_list <- list(a = 1:2, b = 3:5)
map(my_list, add_one)
```

## Applying a function repeatedly

One of the most common `repeated` tasks you might want to do is load in multiple data files:

```{r}
library(readr)
library(dplyr)
library(purrr)
files <- list.files("../data/running", pattern = ".csv", full.names = TRUE)
df <- map(files, read_csv, col_types = cols(.default = "c")) %>%
      bind_rows()

head(df, 15)
```

## Applying a function repeatedly

We can use our own custom functions with the `map()` functions:

```{r, include=FALSE}
clean_names <- function(x){gsub(" ", "_", tolower(x))}
```

```{r, warning=FALSE, message=FALSE}
library(lubridate)

read_run_csv <- function(file){
  read_csv(file, col_types = cols(.default = "c")) %>%
  rename_all(clean_names) %>%
  filter(distance == "1" & split != "Summary") %>%
  mutate_at(vars(matches("time|pace")), hms) %>%
  mutate_if(is.character, as.numeric) %>%
  mutate(filename = sub(".csv", "", basename(file)))
  
}
read_run_csv(files[1])
```

## Applying a function repeatedly

We can use our own custom functions with the `map()` functions:

```{r, warning=FALSE}
rd <- suppressWarnings(map(files, read_run_csv))

class(rd)

length(rd)
```

## Applying a function repeatedly

We now have a list of dataframes! We can map functions to each element in our lists of dataframes:

```{r}
run_sum <- function(x){
  summarise(x, 
            distance = sum(distance, na.rm = TRUE),
            gain = sum(elevation_gain, na.rm = TRUE),
            loss = sum(elev_loss, na.rm = TRUE),
            pace = mean(avg_pace, na.rm = TRUE),
            t_min = min(avg_temperature, na.rm = TRUE),
            t_max = max(avg_temperature, na.rm = TRUE),
            t_mean = mean(avg_temperature, na.rm = TRUE))
}
```

```{r}
out <- map_df(rd, run_sum)
out
```

