---
title: "Reporting data results #2"
output: 
  beamer_presentation:
    theme: "Frankfurt"
    colortheme: "seagull"
    fonttheme: "structuresmallcapsserif"
fontsize: 10pt
---

```{r echo = FALSE, message = FALSE, warning = FALSE}
library(knitr)
library(ggplot2)
library(dplyr)
library(tidyr)
library(ggthemes)
library(faraway)
data(worldcup)
library(gridExtra)
library(choroplethr)
library(choroplethrMaps)
```

# Matrices and lists

## Matrices and lists

This week, we'll talk about the `apply` family of functions, which allow you to apply a function to all values in a vector, matrix, or list. \bigskip

First, you need to know about two more object types in R:

- `matrix`
- `list`

## `matrix`

A matrix is like a data frame, but all the values in all columns must be of the same class (e.g., numeric, character). \bigskip

R uses matrices a lot for its underlying math (e.g., for the linear algebra operations required for fitting regression models). R can do matrix operations quite quickly.

## `matrix`

You can create a matrix with the `matrix` function. Input a vector with the values to fill the matrix and `ncol` to set the number of columns:

```{r}
foo <- matrix(1:10, ncol = 5)
foo
```

## `matrix`

By default, the matrix will fill up by column. You can fill it by row with the `byrow` function: 

```{r}
foo <- matrix(1:10, ncol = 5, byrow = TRUE)
foo
```

## `matrix`

If you want to convert a data frame to a matrix, you can use the `as.matrix` function: 

```{r}
foo <- data.frame(col_1 = 1:2, col_2 = 3:4,
                  col_3 = 5:6, col_4 = 7:8,
                  col_5 = 9:10)
(foo <- as.matrix(foo))
```

## `matrix`

You can index matrices with square brackets, just like data frames: 

```{r}
foo[1, 1:2]
```

You cannot, however, use "tidyverse" functions with matrices: 

```{r eval = FALSE}
foo %>% filter(col_1 == 1)
```
```
Error in UseMethod("filter_") : 
  no applicable method for 'filter_' applied to
  an object of class "c('matrix', 'integer',
  'numeric')"
```

## `matrix`

All elements in a matrix must have the same class. \bigskip

The matrix will default to make all values the most general class of any of the values, in any column. For example, if we replaced one numeric value with the character "a", everything would turn into a character:

```{r}
foo[1, 1] <- "a"
foo
```

## `list`

A list has different elements, just like a data frame has different columns. However, the different elements of a list can have different lengths (unlike the columns of a data frame). The different elements can also have different classes.

```{r}
bar <- list(some_letters = letters[1:3],
            some_numbers = 1:5, 
            some_logical_values = c(TRUE, FALSE))
bar
```

## `list`

To index an element from a list, use double square brackets. You can use bracket indexing either with numbers (which element in the list?) or with names. You can also index lists with the `$` operator.

```{r}
bar[[1]]
bar[["some_numbers"]]
bar$some_logical_values
```

## `list`

Lists can be used to contain data with an unusual structure and / or lots of different components. For example, the information from fitting a regression is often stored as a list:

```{r}
my_mod <- glm(rnorm(10) ~ c(1:10))
is.list(my_mod)
```

## `list`

```{r}
head(names(my_mod), 3)
my_mod[["coefficients"]]
```

# `apply` functions

## `apply` family 

There is a whole family of `apply` functions, as part of base R. These include:

- `apply`: Apply a function over all the rows (`MARGIN = 1`) or columns (`MARGIN = 2`) of a matrix
- `lapply`: Apply a function over elements of a list.
- `sapply`: Like `lapply`, but returns a vector instead of a list.

## `apply`

Here is the syntax for `apply`: 

```{r, eval = FALSE}
## Generic code
apply([matrix], MARGIN = [margin (1: rows, 2: columns)],
      FUN =  [function])
```

I'll use the `worldcup` data as an example:

```{r}
ex <- worldcup[ , c("Shots", "Passes", "Tackles", "Saves")]
head(ex)
```

## `apply`

Take the mean of all columns:

```{r}
apply(ex, MARGIN = 2, mean)
```

Take the sum of all rows:

```{r}
head(apply(ex, MARGIN = 1, sum), 4)
```

## `apply`

You can use your own function with any of the `apply` functions. For example, if you wanted to calculate a value for each player that is a weighted mean of some of their statistics, you could run:

```{r}
weighted_mean <- function(soccer_stats,
                          weights = c(0.40, 0.01,
                                      0.25, 1.5)){
  out <- sum(weights * soccer_stats)
  return(out)
}

head(apply(ex, MARGIN = 1, weighted_mean), 4)
```


## `lapply`

`lapply()` will apply a function across a list. The different elements of the list do not have to be the same length (unlike a data frame, where the columns all have to have the same length).

```{r}
(ex <- list(a = c(1:5), b = rnorm(3), c = letters[1:4]))
```

## `lapply`

This call will calculate the mean of each function:

```{r warning = FALSE}
lapply(ex, FUN = mean)
```

## `lapply`

You can include arguments for the function that you specify with `FUN`, and they'll be passed to that function. For example, to get the first value of each element, you can run:

```{r warning = FALSE}
lapply(ex, FUN = head, n = 1)
```

## `sapply`

`sapply()` also applies a function over a list, but it returns a vector rather than a list:

```{r}
sapply(ex, FUN = head, n = 1)
```


## `apply` family 

In practice, I do use `apply()` some, but I can often find a way to do similar things to other `apply` family functions using the tools in `dplyr`. \bigskip

You should know that `apply` family functions take advantage of the matrix structure in R. This can be one of the fastest way to run code in R. It is usually a lot faster than doing the same things with loops. However, unless you are working with large data sets, you may not notice a difference, and "tidyverse" functions are usually comparable in speed. \bigskip

I would recommend using whichever method makes the most sense to you until you run into an analysis that takes a noticeable amount of time to run, and then you might want to work a bit more to optimize your code. \bigskip

# Point maps

## Point maps 

It is very easy to create point maps in R based on longitude and latitude values of specific locations. \bigskip

To get a base map, you can use the `map_data` function from the `ggplot2` package to pull data for maps at different levels ("usa", "state", "world", "county").

## Point maps 

The maps you pull using `map_data` are just data frames. They include the data you need to plot polygon shapes for areas like states and counties. 

```{r warning = FALSE, message = FALSE}
library(ggplot2)
us_map <- map_data("state")
head(us_map, 3)
```

You can add points to these based on latitude and longitude.

## Point maps 

Map choices with `map_data` are currently limited to: 

- `county`
- `state`
- `usa`
- `france`
- `italy`
- `nz`
- `world`
- `world2`

## Point maps 

Mapping uses the `long` and `lat` columns from this data for location:

```{r fig.width = 6, fig.height = 2.5, fig.align = "center"}
north_carolina <- us_map %>% 
  filter(region == "north carolina")
ggplot(north_carolina, aes(x = long, y = lat)) +
  geom_point()
```

## Point maps 

If you try to plot lines, however, you'll have a problem:

```{r fig.width = 3.5, fig.height = 2.25, fig.align = "center"}
carolinas <- us_map %>% 
  filter(str_detect(region, "carolina"))
ggplot(carolinas, aes(x = long, y = lat)) + 
  geom_path()
```

## Point maps 

The `group` column fixes this problem. It will plot a separate path or polygon for each separate group. For mapping, this gives separate groupings for mainland versus islands and for different states:

```{r}
carolinas %>%
  group_by(group) %>%
  slice(1)
```

## Point maps

Using `group = group` avoids the extra lines from the earlier map:

```{r fig.width = 3.5, fig.height = 2.25, fig.align = "center"}
ggplot(carolinas, aes(x = long, y = lat,
                      group = group)) + 
  geom_path()
```

## Point maps

To plot filled regions, use `geom_polygon` with `fill = region`. Also, the "void" theme is often useful when mapping:

```{r fig.width = 4.5, fig.height = 2, fig.align = "center"}
ggplot(carolinas, aes(x = long, y = lat,
                      group = group,
                      fill = region)) + 
  geom_polygon(color = "black") + 
  theme_void()
```

## Point maps 

Here is an example of plotting all of the US by state:

```{r}
map_1 <- ggplot(us_map, aes(x = long, y = lat,
                            group = group)) + 
        geom_polygon(fill = "dodgerblue",
                     color = "white") +
        theme_void()
```

## Point maps

```{r fig.width = 6.5, fig.height = 4, fig.align = "center"}
map_1
```

## Point maps 

To add points to these maps, you can use `geom_point`, again using longitude and latitude to define position. \bigskip

Here I'll use an example of data points related to the story told in last year's ["Serial" podcast](http://serialpodcast.org). 

```{r}
serial <- read.csv("../data/serial_map_data.csv")
head(serial, 3)
```

## Point maps 

[David Robinson](https://github.com/dgrtwo/serial-ggvis/blob/master/serial-preprocessing.Rmd) figured out a way to convert the x and y coordinates in this data to latitude and longitude coordinates. I'm also adding a column for whether of not the point is a cell tower.

```{r message = FALSE, warning = FALSE}
library(dplyr)
serial <- serial %>%
    mutate(long = -76.8854 + 0.00017022 * x,
           lat  = 39.23822 + 1.371014e-04 * y,
           tower = Type == "cell-site")
```

## Point maps 

```{r message = FALSE, warning = FALSE}
serial[c(1:2, (nrow(serial) - 1):nrow(serial)), 
       c("Type", "Name", "long", "lat", "tower")]
```

## Point maps 

Now I can map just Baltimore City and Baltimore County in Maryland and add these points. \bigskip

I used `map_data` to pull the "county" map and specified "region" as "maryland", to limit the map just to Maryland counties. 

```{r}
baltimore <- map_data('county', region = 'maryland')
head(baltimore, 3)
```

## Point maps

From that, I filter to just rows where the `subregion` column was "baltimore city" or "baltimore". 

```{r}
baltimore <- baltimore %>%
  filter(subregion %in% c("baltimore city",
                          "baltimore"))
head(baltimore, 3)
```

## Point maps

I used `geom_point` to plot the points. When you plot points, you need to "ungroup" the `group` column you used to plot the polygons for counties. To do that, set `group = NA` in the `geom_point` statement. 

```{r}
balt_plot <- ggplot(baltimore, 
                    aes(x = long, y = lat, group = group)) + 
        geom_polygon(fill = "lightblue", color = "black") +
        geom_point(data = serial, aes(x = long, y = lat,
                                      group = NA,
                                      color = tower)) + 
        theme_bw()
```

## Point maps

```{r fig.width = 7, fig.height = 4}
balt_plot
```

# Choropleths

## Choropleths in R

There's a fantastic new(-ish) package in R to plot choropleth maps. You could also plot choropleths using `ggplot` and other mapping functions, but I would strongly recommend this new package if you're mapping the US. \bigskip

You will need to install and load the `choroplethr` package in R to use the functions below.

```{r message = FALSE, warning = FALSE}
# install.packages("choroplethr")
library(choroplethr)

# install.packages("choroplethrMaps")
library(choroplethrMaps)
```

## Choropleths in R

At the most basic level, you can use this package to plot some data that comes automatically with the package (you'll just need to load the data using the `data` function). For example, if you wanted to plot state-by-state populations as of 2012, you could use:

```{r warning = FALSE, message=FALSE}
data(df_pop_state)
map_3 <- state_choropleth(df_pop_state)
```

## Choropleths in R

```{r fig.width=8, fig.height=3.5}
map_3
```


## Choropleths in R

You can find out more about the `df_pop_state` data if you type `?df_pop_state`. Notice that, for the data frame, the location is given in a column called `region` and the population size to plot is in a column called `value`. 

```{r}
head(df_pop_state, 3)
```


## Choropleths in R

You could use this function to create any state-level choropleth you wanted, as long as you could create a data frame with a column for states called `region` and a column with the value you want to show called `value`.

## Choropleths in R

You can run similar functions at different spatial resolutions (for example, county or zip code):

```{r}
data(df_pop_county)
head(df_pop_county, 3)
```

## Choropleths in R

You can plot choropleths at this level, as well:

```{r warning = FALSE, message = FALSE}
map_4 <- county_choropleth(df_pop_county)
```

## Choropleths in R

```{r fig.width=8, fig.height=4}
map_4
```


## Choropleths in R

You can even do this for countries of the world:

```{r fig.width = 8, fig.height = 3.5, warning=FALSE, message=FALSE}
data(df_pop_country)
country_choropleth(df_pop_country)
```

## Choropleths in R

You can zoom into states or counties. For example, to plot population by county in Colorado, you could run:

```{r fig.width = 6, fig.height = 3, message = FALSE, warning = FALSE}
county_choropleth(df_pop_county, state_zoom = "colorado")
```

## Choropleths in R

You can also use this package to map different tables from the US Census' American Community Survey. \bigskip 

The package includes the `choroplethr_acs()` function to do this, with an option for which level of map you want (`map = `, choices are "state", "country", and "zip"). If you want to map at the state level, for example, use `state_choroplethr_acs()` (other options are county level and zip code level). 

## Choropleths in R

These functions pull recent Census data directly from the US Census using its API, so they require you to get an API key, which you can get [here](http://api.census.gov/data/key_signup.html). \bigskip

Once you put in your request, they'll email you your key. Once they give you your API key, you'll need to install it on R:

```{r echo = FALSE, message = FALSE, warning = FALSE}
library(acs)
api.key.install(Sys.getenv("acskey"))
```

```{r eval = FALSE}
library(acs)
api.key.install('[your census api key]');
```

## Choropleths in R

You can pick from a large number of American Community Survey tables-- [see here](http://factfinder.census.gov/faces/affhelp/jsf/pages/metadata.xhtml?lang=en&type=dataset&id=dataset.en.ACS_12_5YR) for the list plus ID numbers. If the table has multiple columns, you will be prompted to select which one you want to plot. \bigskip

## Choropleths in R

For example, table B19301 gives per-capita income, so if you wanted to plot that, you could run:

```{r fig.width = 3, fig.height = 2.75, message = FALSE, eval = FALSE}
county_choropleth_acs(tableId = "B19301",
                       state_zoom = c("wyoming",
                                      "colorado"))

```

## Choropleths in R

```{r fig.width = 10, fig.height = 6, message = FALSE, echo = FALSE, fig.align = "center"}
county_choropleth_acs(tableId = "B19301",
                       state_zoom = c("wyoming",
                                      "colorado"))

```


## Google Maps API

The `ggmap` package allows you to use tools from Google Maps directly from R.

```{r message=FALSE, warning = FALSE}
## install.packages("ggmap")
library(ggmap)
```

This package uses the Google Maps API, so you should read their [terms of service](http://developers.google.com/maps/terms) and make sure you follow them. In particular, you are limited to just a certain number of queries per time. 

## Google Maps API

You can use the `get_map` function to get maps for different locations. \bigskip 

You can either use the longitude and latitude of the center point of the map, along with the `zoom` option to say how much to zoom in (3: continent to 20: building) or you can use a character string to specify a location. \bigskip

If you do the second, `get_map` will actually use the Google Maps API to geocode the string to a latitude and longitude and then get the map (you can imagine that this is like searching in Google Maps in the search box for a location).

```{r message = FALSE, warning = FALSE, fig.width = 3.5, fig.height = 3.5, fig.align = "center", eval = FALSE}
beijing <- get_map("Beijing", zoom = 12)
ggmap(beijing)
```

## Google Maps API

```{r message = FALSE, warning = FALSE, fig.width = 3.5, fig.height = 3.5, fig.align = "center", echo = FALSE}
beijing <- get_map("Beijing", zoom = 12)
ggmap(beijing)
```

## Google Maps API

With this package, you can get maps from the following different sources:

- Google Maps
- OpenStreetMap
- Stamen Maps
- CloudMade Maps (You may need a separate API key for this)

## Google Maps API

Here are different examples of Beijing using different map sources. (Also, note that I'm using the option `extent = "device"` to fill up the whole plot are with the map, instead of including axis labels and titles.)

```{r message = FALSE, warning = FALSE}
beijing_a <- get_map("Beijing", zoom = 12,
                     source = "stamen", maptype = "toner")
a <- ggmap(beijing_a, extent = "device")

beijing_b <- get_map("Beijing", zoom = 12,
                     source = "stamen", maptype = "watercolor")
b <- ggmap(beijing_b, extent = "device")

beijing_c <- get_map("Beijing", zoom = 12,
                     source = "google", maptype = "hybrid")
c <- ggmap(beijing_c, extent = "device")
```

## Google Maps API

```{r fig.width = 7}
grid.arrange(a, b, c, nrow = 1) 
```


## Google Maps API

As with the maps from `ggplot2`, you can add points to these maps:

```{r warning = FALSE, message = FALSE, echo = FALSE}
library(tidyr)
serial_phone <- read.csv("../data/serial_phone_data.csv") %>% 
        mutate(Cell_Site = substring(Cell_Site, 1, 4),
               Call_Time = as.POSIXct(Call_Time, format = "%d/%m/%y %H:%M",
                                      tz = "EST")) %>%
        left_join(serial, by = c("Cell_Site" = "Name")) %>%
        select(Person_Called, Call_Time, Duration, long, lat) %>%
        filter(!(Person_Called %in% c("incoming", "# + Adnan cell"))) %>%
        arrange(Call_Time)
```

```{r message = FALSE, warning = FALSE, fig.width = 5}
serial_map <- get_map(c(-76.7, 39.3), zoom = 12,
                      source = "stamen",
                      maptype = "toner")
serial_map <- ggmap(serial_map, extent = "device") + 
        geom_point(data = serial_phone,
                   aes(x = long, y = lat),
                   color = "red", size = 3,
                   alpha = 0.4) +
        geom_point(data = subset(serial,
                                 Type != "cell-site"),
                   aes(x = long, y = lat),
                   color = "darkgoldenrod1",
                   size = 2)
```

## Google Maps API

```{r warning = FALSE, message = FALSE, fig.width = 4, fig.height = 4, fig.align = "center", echo = FALSE}
serial_map
```


## Google Maps API

You can also use the Google Maps API, through the `geocode` function, to get the latitude and longitude of specific locations. Basically, if the string would give you the right location if you typed it in Google Maps, `geocode` should be able to geocode it.

For example, you can get the location of CSU:

```{r message = FALSE, warning = FALSE}
geocode("Colorado State University")
```

## Google Maps API

You can also get a location by address through this:

```{r message = FALSE, warning = FALSE}
geocode("1 First St NE, Washington, DC")
```

## Google Maps API

You can get distances, too, using the `mapdist` function with two locations. This will give you distance and also time. 

```{r message = FALSE, warning = FALSE}
mapdist("Fort Collins CO",
        "1 First St NE, Washington, DC") %>%
  select(from, miles, hours)
```

# String operations

## String operations

The `str_trim` function from the `stringr` package allows you to trim leading and trailing whitespace: 

```{r}
with_spaces <- c("    a ", "  bob", " gamma")
with_spaces
str_trim(with_spaces)
```

This is rarer, but if you ever want to, you can add leading and / or trailing whitespace to elements of a character vector with `str_pad` from the `stringr` package.

## String operations

There are also functions to change a full character string to uppercase, lowercase, or title case: 

```{r echo = FALSE}
library(titanic)
data(titanic_train)
```

```{r}
titanic_train$Name[1]
str_to_upper(titanic_train$Name[1])
str_to_lower(titanic_train$Name[1])
str_to_title(str_to_lower(titanic_train$Name[1]))
```

