---
title: "Reporting data results #3"
author: "Brooke Anderson"
date: "November 14, 2018"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r echo = FALSE}
library(ggplot2)
```

## `htmlWidgets`

Very smart people have been working on creating interactive graphics in R for a long time. So far, nothing "native" to" R has taken off in a big way (although keep an eye out for `ggvis`). \bigskip

There is now a series of R packages that allow you to create plots from these JavaScript libraries from within R. \bigskip

There is a website with much more on these `htmlWidgets` at http://www.htmlwidgets.org. 

## `htmlWidgets`

JavaScript has developed a number of interactive graphics libraries that can be for documents viewed in a web browser. These work by binding data to support vector graphics (SVGs). \bigskip

They allow you to do things like zoom and pan. The graphics can also "react" to certain events. For example, they can show a pop-up when you hover over or click on a point. 


## `htmlWidgets`

Some of the packages availabe to help you create interactive graphics from R using JavaScript graphics libraries: 

- `leaflet`: Mapping 
- `dygraphs`: Time series
- `plotly`: A variety of plots, including maps
- `rbokeh`: A variety of plots, including maps
- `networkD3`: Network data
- `d3heatmap`: Heatmaps
- `DT`: Data tables
- `DiagrammeR`: Diagrams and flowcharts

## `htmlWidgets`

These packages can be used to make some pretty cool interactive visualizations for HTML output from R Markdown or Shiny (you can also render any of theme in RStudio). 

There are, however, a few limitations: 

- Written by different people. The different packages have different styles as well as different interfaces. Learning how to use one package may not help you much with other of these packages.
- Many are still in development, often in early development. 


```{r echo = FALSE}
library(dplyr)
library(lubridate)
library(sp)
library(GISTools)
```


## Example data

```{r message = FALSE, results='hide'}
library(tigris)
denver_tracts <- tracts(state = "CO", county = 31, 
                        cb = TRUE, class = "sf")
load("../data/fars_colorado.RData")
accident_data <- driver_data %>%
  dplyr::select(state, st_case, county, latitude, longitud,
                date, fatals, drunk_dr) %>%
  dplyr::filter(county == 31 & longitud < -104.1) %>%
  dplyr::distinct()
```

## Leaflet

"Leaflet" is a JavaScript library for making interactive maps. You can find out more about the JavaScript version here: http://leafletjs.com

The `leaflet` package brings this functionality to R. The R Studio group has created a website on `leaflet`: http://rstudio.github.io/leaflet/. This website walks you through different options available with `leaflet`.

```{r}
library(leaflet)
```

## Add map background

If you just run `leaflet()`, you just get a blank leaflet area: 

```{r fig.width = 7, fig.height = 3}
leaflet()
```

## Add map background

In `leaflet`, the map background is composed of *tiles*. To get something more interesting, you'll need to add tiles to your leaflet map. If you don't include any other data, the leaflet map will include the world:

```{r eval = FALSE}
leaflet() %>%
  addTiles()
```

## Add map background

```{r echo = FALSE}
leaflet() %>%
  addTiles()
```

## Adding markers

For htmlWidgets, points are often referred to as *markers*.

Once you add these markers, the map will automatically scale to a reasonable size for their bounding box.

```{r eval = FALSE}
leaflet() %>%
  addTiles() %>%
  addMarkers(data = accident_data, lng = ~ longitud, lat = ~ latitude)
```

Use `lng` and `lat` to tell R which columns contain data on longitude and latitude for each point. This is not needed if you are using a spatial object (e.g., `SpatialPointsDataFrame`). Further, R will try to guess the columns in a regular dataframe.

## Adding markers

```{r echo = FALSE}
leaflet() %>%
  addTiles() %>%
  addMarkers(data = accident_data, lng = ~ longitud, lat = ~ latitude)
```

## Adding markers

You can use several types of R objects for your data for `leaflet`:

- Dataframe with columns for latitude and longitude
- Simple feature objects 
- Latitude-longitude matrix

## Adding markers

You can choose circles for your markers instead by using `addCircleMarkers`. You can adjust the circle size with `radius`. 

```{r eval = FALSE}
leaflet() %>%
  addTiles() %>%
  addCircleMarkers(data = accident_data, radius = 2,
                   lng = ~ longitud, lat = ~ latitude)
```

The `radius` argument specifies the size of the circle. For `CircleMarkers`, the size will reset as you zoom in and out. If you want something with a constant radius (e.g., in meters), you can add `Circles`.

## Adding markers

```{r echo = FALSE}
leaflet() %>%
  addTiles() %>%
  addCircleMarkers(data = accident_data, radius = 2,
                   lng = ~ longitud, lat = ~ latitude)
```

## Adding markers

If you have a lot of overlapping data, you can also use the `clusterOptions` argument to show the markers as clusters that group together when you zoom out but split up when you zoom in:

```{r eval = FALSE}
leaflet() %>%
  addTiles() %>%
  addMarkers(data = accident_data, 
                   lng = ~ longitud, lat = ~ latitude,
                   clusterOptions = markerClusterOptions())
```

## Adding markers

```{r echo = FALSE}
leaflet() %>%
  addTiles() %>%
  addMarkers(data = accident_data, 
                   lng = ~ longitud, lat = ~ latitude,
                   clusterOptions = markerClusterOptions())
```

## Add map background

For the background, the default is to use map tiles from OpenStreetMap. However, you can change the source of the tiles by using `addProviderTiles`. For example, to use Stamen Watercolor, you can call:

```{r eval = FALSE}
leaflet() %>%
  addProviderTiles("Stamen.Watercolor") %>%
  addCircleMarkers(data = accident_data, radius = 2,
                   lng = ~ longitud, lat = ~ latitude)
```

## Add map background

```{r echo = FALSE}
leaflet() %>%
  addProviderTiles("Stamen.Watercolor") %>%
  addCircleMarkers(data = accident_data, radius = 2,
                   lng = ~ longitud, lat = ~ latitude)
```


## Add map background

```{r fig.width = 7, fig.height = 3}
leaflet() %>%
  addProviderTiles("Esri.WorldStreetMap") %>%
  addCircleMarkers(data = accident_data, radius = 2,
                   lng = ~ longitud, lat = ~ latitude)
```

## Add map background

You can see previews of provider choices here: http://leaflet-extras.github.io/leaflet-providers/preview/index.html.

## Pop-ups

You can use the `popup` option to show information when the user clicks on a marker. 

It's easiest to do this if you have the information you want to show in the dataframe with the location data. For example, we have date-time, number of fatalities, and number of drunk drivers in this data: 

```{r}
accident_data %>% 
  dplyr::select(date, fatals, drunk_dr) %>% 
  dplyr::slice(1:3)
```

## Pop-ups

If we want to show day of the week, month, hour, and number of fatalities, go ahead and calculate any value not already in the dataset:

```{r}
library(lubridate)
accident_data <- accident_data %>%
  mutate(weekday = wday(date, label = TRUE, abbr = FALSE),
         month = month(date, label = TRUE, abbr = FALSE),
         hour = format(date, format = "%H:%M"))
```

## Pop-ups

The popup text needs to be a character vector, written in HTML syntax. You can create that vector first, and then pass it to the `popup` argument.

```{r}
popup_info <- paste0("<b>Weekday:</b>  ", 
                                  accident_data$weekday, "<br/>",
                                  "<b>Month:</b>  ",
                                  accident_data$month, "<br/>",
                                  "<b>Hour:</b>  ",
                                  accident_data$hour, "<br/>",
                                  "<b>Fatalities:</b>  ",
                                  accident_data$fatals)
popup_info[1:3]
```

## Pop-ups

Now pass that vector to the `popup` argument for the layer you want to pair it with:

```{r eval = FALSE}
leaflet() %>%
  addTiles() %>%
  addCircleMarkers(data = accident_data, radius = 2,
                   lng = ~ longitud, lat = ~ latitude,
                   popup = popup_info)
```

## Pop-ups

```{r echo = FALSE}
leaflet() %>%
  addTiles() %>%
  addCircleMarkers(data = accident_data, radius = 2,
                   lng = ~ longitud, lat = ~ latitude,
                   popup = popup_info)
```

## Pop-ups

If you aren't familiar with HTML syntax, here's one cheatsheet: http://web.stanford.edu/group/csp/cs21/htmlcheatsheet.pdf

In the popups, you can use HTML to format things like color, typeface, and size. You can also add links.

## Mapping values to color

To use color to show a value, you need to do a few things. First, you need to the the `colorFactor` function (or another in its family) to create a function for mapping from values to colors. Then, you need to use this within the call to add the markers.

```{r eval = FALSE}
library(viridisLite)
pal <- colorFactor(viridis(5), accident_data$drunk_dr)
leaflet() %>%
  addProviderTiles("OpenStreetMap.BlackAndWhite") %>%
  addCircleMarkers(data = accident_data, radius = 2,
                   lng = ~ longitud, lat = ~ latitude,
                   popup = popup_info, 
                   color = pal(accident_data$drunk_dr)) 
```

## Mapping values to color

```{r echo = FALSE}
library(viridisLite)
pal <- colorFactor(viridis(5), accident_data$drunk_dr)
leaflet() %>%
  addProviderTiles("OpenStreetMap.BlackAndWhite") %>%
  addCircleMarkers(data = accident_data, radius = 2,
                   lng = ~ longitud, lat = ~ latitude,
                   popup = popup_info, 
                   color = pal(accident_data$drunk_dr)) 
```

## Mapping values to color

The `colorFactor` function (and friends) are a pretty cool type of function that actually creates a new function: 

```{r}
pal <- colorFactor(viridis(5), accident_data$drunk_dr)
class(pal)
head(pal)
```

## Adding a legend

Once you are showing something with color, you can add a legend to explain it. You can do that with the `addLegend` function, which must include values for the color palette and values for each point from this color palette.

```{r eval = FALSE}
library(viridisLite)
pal <- colorFactor(viridis(5), accident_data$drunk_dr)
leaflet() %>%
  addProviderTiles("OpenStreetMap.BlackAndWhite") %>%
  addCircleMarkers(data = accident_data, radius = 2,
                   lng = ~ longitud, lat = ~ latitude,
                   popup = popup_info, 
                   color = pal(accident_data$drunk_dr)) %>%
  addLegend(pal = pal, values = accident_data$drunk_dr)
```

## Adding a legend

```{r echo = FALSE}
library(viridisLite)
pal <- colorFactor(viridis(5), accident_data$drunk_dr)
leaflet() %>%
  addProviderTiles("OpenStreetMap.BlackAndWhite") %>%
  addCircleMarkers(data = accident_data, radius = 2,
                   lng = ~ longitud, lat = ~ latitude,
                   popup = popup_info, 
                   color = pal(accident_data$drunk_dr)) %>%
  addLegend(pal = pal, values = accident_data$drunk_dr)
```

## Adding polygons

You can add polygons with the `addPolygons` function.

```{r eval = FALSE}
leaflet() %>%
  addProviderTiles("OpenStreetMap.BlackAndWhite") %>%
  addPolygons(data = denver_tracts)
```

## Adding polygons

```{r echo = FALSE}
leaflet() %>%
  addProviderTiles("OpenStreetMap.BlackAndWhite") %>%
  addPolygons(data = denver_tracts)
```

## Adding polygons

You can add popups for polygons, as well.

```{r eval = FALSE}
polygon_popup <- paste0("Tract ID:  ", 
                        denver_tracts$TRACTCE)
leaflet() %>%
  addProviderTiles("OpenStreetMap.BlackAndWhite") %>%
  addPolygons(data = denver_tracts, popup = polygon_popup)
```

## Adding polygons

```{r echo = FALSE}
polygon_popup <- paste0("Tract ID:  ", 
                        denver_tracts$TRACTCE)
leaflet() %>%
  addProviderTiles("OpenStreetMap.BlackAndWhite") %>%
  addPolygons(data = denver_tracts, popup = polygon_popup)
```

## Adding polygons

You can overlay different elements. For example, you can show both accidents and tracts:

```{r eval = FALSE}
polygon_popup <- paste0("Tract ID:  ", 
                        denver_tracts$TRACTCE)
leaflet() %>%
  addProviderTiles("OpenStreetMap") %>%
  addPolygons(data = denver_tracts, popup = polygon_popup,
              color = "#000000", fillColor = "969696", 
              weight = 2) %>%
  addCircleMarkers(data = accident_data, radius = 2,
                   lng = ~ longitud, lat = ~ latitude,
                   popup = popup_info, opacity = 0.9,
                   color = pal(accident_data$drunk_dr)) %>%
  addLegend(pal = pal, values = accident_data$drunk_dr, opacity = 0.9)
```

## Adding polygons

```{r echo = FALSE}
polygon_popup <- paste0("Tract ID:  ", 
                        denver_tracts$TRACTCE)
leaflet() %>%
  addProviderTiles("OpenStreetMap") %>%
  addPolygons(data = denver_tracts, popup = polygon_popup,
              color = "#000000", fillColor = "969696", 
              weight = 2) %>%
  addCircleMarkers(data = accident_data, radius = 2,
                   lng = ~ longitud, lat = ~ latitude,
                   popup = popup_info, opacity = 0.9,
                   color = pal(accident_data$drunk_dr)) %>%
  addLegend(pal = pal, values = accident_data$drunk_dr, opacity = 0.9)
```

## Adding polygons

You can add the ability for the user to pick which layers to see using `addLayersControls`. 

```{r eval = FALSE}
leaflet() %>%
  addProviderTiles("OpenStreetMap") %>%
  addPolygons(data = denver_tracts, popup = polygon_popup,
              color = "#000000", fillColor = "969696", 
              weight = 2, group = "tracts") %>%
  addCircleMarkers(data = accident_data, radius = 2,
                   lng = ~ longitud, lat = ~ latitude,
                   popup = popup_info, opacity = 0.9,
                   color = pal(accident_data$drunk_dr), 
                   group = "accidents") %>%
  addLegend(pal = pal, values = accident_data$drunk_dr, opacity = 0.9) %>%
  addLayersControl(baseGroups = c("base map"), 
                   overlayGroups = c("tracts", "accidents"))
```

## Adding polygons

```{r echo = FALSE}
leaflet() %>%
  addProviderTiles("OpenStreetMap") %>%
  addPolygons(data = denver_tracts, popup = polygon_popup,
              color = "#000000", fillColor = "969696", 
              weight = 2, group = "tracts") %>%
  addCircleMarkers(data = accident_data, radius = 2,
                   lng = ~ longitud, lat = ~ latitude,
                   popup = popup_info, opacity = 0.9,
                   color = pal(accident_data$drunk_dr), 
                   group = "accidents") %>%
  addLegend(pal = pal, values = accident_data$drunk_dr, opacity = 0.9) %>%
  addLayersControl(baseGroups = c("base map"), 
                   overlayGroups = c("tracts", "accidents"))
```

## Find out more

Here are some good tutorials for trying out other examples of `leaflet` in R:

http://zevross.com/blog/2015/10/14/manipulating-and-mapping-us-census-data-in-r-using-the-acs-tigris-and-leaflet-packages-3/

http://trendct.org/2015/06/26/tutorial-how-to-put-dots-on-a-leaflet-map-with-r/

## `plotly` package

From the package documentation: 

> "Easily translate `ggplot2` graphs to an interactive web-based version and / or create custom web-based visualizations directly from R."

- Like many of the packages today, draws on functionality external to R, but within a package that allows you to work exclusively within R.
- Allows you to create interactive graphs from R. The functions extend much of the `ggplot2` code you've learned.
- Interactivity will only work within RStudio or on documents rendered to HTML.

## `plotly` package

The `plotly` package allows an interface to let you work with `plotly.js` code directly using R code. \bigskip

`plotly.js` is an open source library for creating interactive graphs in JavaScript. This JavaScript library is built on `d3.js` (Data-Driven Documents), which is a key driver in interactive web-based data graphics today.

## `plotly` package

There are two main ways of create plots within `plotly`: 

- Use one of the functions to create a customized interactive graphic: 
    + `plot_ly`: Workhorse of `plotly`, renders most non-map types of graphs
    + `plot_geo`, `plot_mapbax`: Specific functions for creating `plotly` maps. 
- Create a `ggplot` object and then convert it to a `plotly` object using the `ggplotly` function.

## `plotly` package

```{r fig.width = 4, fig.height = 3.5, fig.align = "center"}
library(faraway); data(worldcup); library(dplyr)
library(plotly)
a <- worldcup %>% ggplot(aes(x = Time, y = Shots)) + geom_point()
ggplotly(a)
```

## `plotly` package

```{r fig.width = 5, fig.height = 3.5, fig.align = "center"}
a <- worldcup %>% ggplot(aes(x = Time, y = Shots, color = Position)) + 
  geom_point()
ggplotly(a)
```

## `plotly` package

You can also use this with other `ggplot2` functionality, like faceting: 

```{r eval = FALSE}
shots_vs_time <- worldcup %>%
  mutate(Name = rownames(worldcup)) %>%
  filter(Team %in% c("Netherlands", "Germany", "Spain", "Uruguay")) %>%
  ggplot(aes(x = Time, y = Shots, color = Position, group = Name)) + 
  geom_point() + 
  facet_wrap(~ Team)
ggplotly(shots_vs_time)
```

## `plotly` package

```{r echo = FALSE}
shots_vs_time <- worldcup %>%
  mutate(Name = rownames(worldcup)) %>%
  filter(Team %in% c("Netherlands", "Germany", "Spain", "Uruguay")) %>%
  ggplot(aes(x = Time, y = Shots, color = Position, group = Name)) + 
  geom_point() + 
  facet_wrap(~ Team)
ggplotly(shots_vs_time)
```


## `plotly` package

If you pipe to the `rangeslider` function, it allows the viewer to zoom in on part of the x range. (This can be particularly nice for time series.) \bigbreak

You should have a dataset available through your R session named `USAccDeaths`. This gives a monthly county of accidental deaths in the US for 1973 to 1978. This code will plot it and add a range slider on the lower x-axis.

```{r eval = FALSE}
plot_ly(x = time(USAccDeaths), y = USAccDeaths) %>% 
  add_lines() %>% rangeslider()
```

## `plotly` package

```{r echo = FALSE}
plot_ly(x = time(USAccDeaths), y = USAccDeaths) %>% 
  add_lines() %>%
  rangeslider()
```

## `plotly` package

For a 3-D scatterplot, add a mapping to the `z` variable:

```{r eval = FALSE}
worldcup %>%
  plot_ly(x = ~ Time, y = ~ Shots, z = ~ Passes,
          color = ~ Position, size = I(4)) %>%
  add_markers()
```

## `plotly` package

```{r echo = FALSE}
worldcup %>%
  plot_ly(x = ~ Time, y = ~ Shots, z = ~ Passes,
          color = ~ Position, size = I(3)) %>%
  add_markers()
```

## `plotly` package

The `volcano` data comes with R and is in a matrix format. Each value gives the elevation for a particular pair of x- and y-coordinates.

```{r}
dim(volcano)
volcano[1:4, 1:4]
```

## `plotly` package

```{r}
plot_ly(z = ~ volcano, type = "surface")
```

## `plotly` package

Mapping with `plotly` can build on some data that comes with base R or other packages you've likely added (or can add easily, as with the `map_data` function from `ggplot2`). For example, we can map state capitals and cities with > 40,000 people using data in the `us.cities` dataframe in the `maps` package:

```{r}
head(maps::us.cities, 3)
```


## `plotly` package

Here is code you can use to map all of these cities on a US map:

```{r eval = FALSE}
ggplot2::map_data("state") %>%
   group_by(group) %>% 
   plot_ly(x = ~long, y = ~lat) %>%
   add_polygons(hoverinfo = "none") %>%
   add_markers(text = ~paste(name, "<br />", pop), hoverinfo = "text",
               alpha = 0.25,
     data = filter(maps::us.cities, -125 < long & long < -60 &
                     25 < lat & lat < 52)) %>%
   layout(showlegend = FALSE)
```

## `plotly` package

```{r echo = FALSE}
ggplot2::map_data("world", "usa") %>%
   group_by(group) %>% filter(-125 < long & long < -60 &
                                25 < lat & lat < 52) %>%
   plot_ly(x = ~long, y = ~lat) %>%
   add_polygons(hoverinfo = "none") %>%
   add_markers(text = ~paste(name, "<br />", pop), hoverinfo = "text",
               alpha = 0.25,
     data = filter(maps::us.cities, -125 < long & long < -60 &
                     25 < lat & lat < 52)) %>%
   layout(showlegend = FALSE)
```


## `plotly` package

The creator of the R `plotly` package has written a bookdown book on the package that you can read [here](https://cpsievert.github.io/plotly_book/). It provides extensive details and examples for using `plotly`.

## `rbokeh` package

The `rbokeh` package provides an R interface to a Python interactive visualization library, Bokeh.

There is a website with many more details on using the `rbokeh` package: https://hafen.github.io/rbokeh/ 

You can find out more about the original Python library, Bokeh, at http://bokeh.pydata.org/en/latest/. 

## `rbokeh` package

```{r fig.align = "center"}
library(rbokeh)
figure(width = 600, height = 300) %>%
  ly_points(Time, Shots, data = worldcup,
    color = Position, hover = list(Time, Shots))
```

## `rbokeh` package

```{r echo = FALSE}
# prepare data
elements <- subset(elements, !is.na(group))
elements$group <- as.character(elements$group)
elements$period <- as.character(elements$period)

# add colors for groups
metals <- c("alkali metal", "alkaline earth metal", "halogen",
  "metal", "metalloid", "noble gas", "nonmetal", "transition metal")
colors <- c("#a6cee3", "#1f78b4", "#fdbf6f", "#b2df8a", "#33a02c",
  "#bbbb88", "#baa2a6", "#e08e79")
elements$color <- colors[match(elements$metal, metals)]
elements$type <- elements$metal

# make coordinates for labels
elements$symx <- paste(elements$group, ":0.1", sep = "")
elements$numbery <- paste(elements$period, ":0.8", sep = "")
elements$massy <- paste(elements$period, ":0.15", sep = "")
elements$namey <- paste(elements$period, ":0.3", sep = "")

# create figure
p <- figure(title = "Periodic Table", tools = c("resize", "hover"),
  ylim = as.character(c(7:1)), xlim = as.character(1:18),
  xgrid = FALSE, ygrid = FALSE, xlab = "", ylab = "",
  height = 445, width = 800) %>%

# plot rectangles
ly_crect(group, period, data = elements, 0.9, 0.9,
  fill_color = color, line_color = color, fill_alpha = 0.6,
  hover = list(name, atomic.number, type, atomic.mass,
    electronic.configuration)) %>%

# add symbol text
ly_text(symx, period, text = symbol, data = elements,
  font_style = "bold", font_size = "10pt",
  align = "left", baseline = "middle") %>%

# add atomic number text
ly_text(symx, numbery, text = atomic.number, data = elements,
  font_size = "6pt", align = "left", baseline = "middle") %>%

# add name text
ly_text(symx, namey, text = name, data = elements,
  font_size = "4pt", align = "left", baseline = "middle") %>%

# add atomic mass text
ly_text(symx, massy, text = atomic.mass, data = elements,
  font_size = "4pt", align = "left", baseline = "middle")

p
```

## `dygraphs` package

The `dygraphs` package lets you create interactive time series plots from R using the `dygraphs` JavaScript library. \bigskip

The main function syntax is fairly straightforward. Like many of these packages, it allows piping. \bigskip

There is a website with more information on using `dygraphs` available at http://rstudio.github.io/dygraphs/index.html. 

## `dygraphs` package

For example, here is the code to plot monthly deaths from lung diseases in the UK in the 1970s. 

```{r eval = FALSE}
library(dygraphs)
lungDeaths <- cbind(mdeaths, fdeaths)
dygraph(lungDeaths) %>%
  dySeries("mdeaths", label = "Male") %>%
  dySeries("fdeaths", label = "Female")
```

## `dygraphs` package

For example, here is the code to plot monthly deaths from lung diseases in the UK in the 1970s. 

```{r echo = FALSE, fig.height = 3.5}
library(dygraphs)
lungDeaths <- cbind(mdeaths, fdeaths)
dygraph(lungDeaths) %>%
  dySeries("mdeaths", label = "Male") %>%
  dySeries("fdeaths", label = "Female")
```

## `DT` package

The `DT` package provides a way to create interactive tables in R using the JavaScript `DataTables` library. \bigskip

We've already seen some examples of this output in some of the Shiny apps I showed last week. You can also use this package to include interactive tables in R Markdown documents you plan to render to HTML. \bigskip

There is a website with more information on this package at http://rstudio.github.io/DT/. 

## `DT` package

```{r out.height = 200, out.width = 400, fig.width = 3}
library(DT)
datatable(worldcup)
```


## `networkD3` package

The `networkd3` package allows you to create different networks. For example, a simple network: 

```{r eval = FALSE}
library(networkD3)
src <- c("A", "A", "A", "A",
        "B", "B", "C", "C", "D")
target <- c("B", "C", "D", "J",
            "E", "F", "G", "H", "I")
networkData <- data.frame(src, target)
simpleNetwork(networkData)
```

## `networkD3` package

```{r echo = FALSE, fig.height = 4}
library(networkD3)
src <- c("A", "A", "A", "A",
        "B", "B", "C", "C", "D")
target <- c("B", "C", "D", "J",
            "E", "F", "G", "H", "I")
networkData <- data.frame(src, target)
simpleNetwork(networkData)
```

## `networkD3` package

A more complex network: 

```{r eval = FALSE}
data(MisLinks)
data(MisNodes)
forceNetwork(Links = MisLinks, Nodes = MisNodes,
            Source = "source", Target = "target",
            Value = "value", NodeID = "name",
            Group = "group", opacity = 0.8)
```

## `networkD3` package

```{r echo = FALSE, fig.height = 4}
# Load data
data(MisLinks)
data(MisNodes)

# Plot
forceNetwork(Links = MisLinks, Nodes = MisNodes,
            Source = "source", Target = "target",
            Value = "value", NodeID = "name",
            Group = "group", opacity = 0.8)
```

## `networkD3` package

A Sankey diagram: 

```{r eval = FALSE}
URL <- paste0(
        "https://cdn.rawgit.com/christophergandrud/networkD3/",
        "master/JSONdata/energy.json")
Energy <- jsonlite::fromJSON(URL)
sankeyNetwork(Links = Energy$links, Nodes = Energy$nodes, 
              Source = "source", Target = "target", 
              Value = "value", NodeID = "name",
              units = "TWh", fontSize = 12, 
              nodeWidth = 30)
```

## `networkD3` package

```{r echo = FALSE, fig.height = 4.5}
URL <- paste0(
        "https://cdn.rawgit.com/christophergandrud/networkD3/",
        "master/JSONdata/energy.json")
Energy <- jsonlite::fromJSON(URL)
sankeyNetwork(Links = Energy$links, Nodes = Energy$nodes, Source = "source",
             Target = "target", Value = "value", NodeID = "name",
             units = "TWh", fontSize = 12, nodeWidth = 30)
```


## Creating your own widget

If you find a JavaScript visualization library and would like to create bindings to R, you can create your own package for a new htmlWidget. \bigskip

There is advice on creating your own widget for R available at http://www.htmlwidgets.org/develop_intro.html.